### 2022-08-05 supp/01-size.class.Rmd
<init analysis>

# Define size classes

## Load packages + data

```{r load.packages.szclass, out.width = "100%", dpi = 300, echo=TRUE, warning=FALSE}
# load required analysis packages
library(here)
library(StereoMorph)
library(geomorph)
library(tidyverse)
library(wesanderson)
library(ggplot2)

# read data and define number of sLMs ----
shapes <- readShapes("shapes")
shapesGM <- readland.shapes(shapes, nCurvePts = c(15, 10, 15))

# read qualitative data
qdata <- read.csv("qdata.csv", 
                  header = TRUE, 
                  row.names = 1)
```

## Generalised Procrustes Analysis

Landmark data were aligned to a global coordinate system [@RN11622;@RN11623;@RN11563], achieved through generalized Procrustes superimposition [@RN478] performed in R 4.1.3 [@R] using the `geomorph` library v. 4.0.3 [@RN11530;@RN1774;@RN9565]. Procrustes superimposition translates, scales, and rotates the coordinate data to allow for comparisons among objects [@RN11564;@RN478]. The `geomorph` package uses a partial Procrustes superimposition that projects the aligned specimens into tangent space subsequent to alignment in preparation for the use of multivariate methods that assume linear space [@RN1646;@RN11563].

```{r outlines.szclass, out.width = "100%", dpi = 300, echo=TRUE, warning=FALSE}
# gpa ----
Y.gpa <- gpagen(shapesGM, print.progress = FALSE)

# plot
plot(Y.gpa)

# geomorph data frame ----
gdf <- geomorph.data.frame(shape = Y.gpa$coords, 
                           size = Y.gpa$Csize,
                           site = qdata$site)

# add centroid size to qdata
qdata$csz <- Y.gpa$Csize
```

## Assign size class

```{r setwd.szclass ld define variables 3, echo=TRUE, eval = FALSE}
#select vars
pre.data <- qdata %>% 
  select(site:csz)

# millsap subset----
millsap <- filter(pre.data,
                  site == "Millsap Cache")

# calculate mean centroid size
mean.csz <- mean(millsap$csz)

# assign L/S varieties based on mean centroid size
lcheck2 <- millsap$csz >= mean.csz
lcheck3 <- millsap$csz >= 0 & millsap$csz < mean.csz

millsap$size <- ""
millsap$size <- ifelse(lcheck2, "L", millsap$size)
millsap$size <- ifelse(lcheck3, "S", millsap$size)

# an13 subset----
an13 <- subset(pre.data,
               site == "Jowell Farm")

# calculate the mean length
mean.csz <- mean(an13$csz)

# assign L/S varieties based on mean length
lcheck2 <- an13$csz >= mean.csz
lcheck3 <- an13$csz >= 0 & an13$csz < mean.csz

an13$size <- ""
an13$size <- ifelse(lcheck2, "L", an13$size)
an13$size <- ifelse(lcheck3, "S", an13$size)

# aggregate data
mx <- bind_rows(millsap, an13)

# join site and size class
qdata2 <- mx %>% 
  unite("merged", site, size, remove = FALSE)

# output csv for use in gm analysis
write.csv(qdata2, "qdata.merged.csv")
```

### 2022-08-05 supp/02-2dgm-rawmat.Rmd
<init analysis>

# GM - Site/Raw Material

Elliptical bifaces from the Millsap Cache and Jowell Farm were used to examine whether their morphology remains stable, or expresses morphological variability by site/raw meterial. Variability in elliptical biface morphology provides evidence associated with differences in morphology as a function of geography/raw material.

This analysis follows recent analyses of Perdiz arrow points ([Selden and Dockall, in review](https://github.com/seldenlab/perdiz3/blob/main/ms/perdiz3/perdiz3.pdf)) and Gahagan biface morphology across the southern Caddo area and central Texas [@RN11783;@RN21001;@RN11264], where Perdiz arrow points and Gahagan bifaces have been found to differ in shape across the same geography as Caddo bottles [@RN11801;@RN11782;@RN11716;@RN8312]. This analysis also follows another geometric morphometric analysis in which Perdiz arrow points from the ancestral Caddo area were found to differ as a function of _time_, _raw material_, and _burial context_ [@RN9551].

## Load packages + data

```{r load.packagessiteraw, out.width = "100%", dpi = 300, echo=TRUE, warning=FALSE}
# load required analysis packages
library(here)
library(StereoMorph)
library(geomorph)
library(tidyverse)
library(wesanderson)
library(ggplot2)

# read qualitative data
qdata <- read.csv("qdata.merged.csv", 
                  header = TRUE, 
                  row.names = 1)
```

## Generalised Procrustes Analysis

Landmark data were aligned to a global coordinate system [@RN11622;@RN11623;@RN11563], achieved through generalized Procrustes superimposition [@RN478] performed in R 4.1.3 [@R] using the `geomorph` library v. 4.0.3 [@RN11530;@RN1774;@RN9565]. Procrustes superimposition translates, scales, and rotates the coordinate data to allow for comparisons among objects [@RN11564;@RN478]. The `geomorph` package uses a partial Procrustes superimposition that projects the aligned specimens into tangent space subsequent to alignment in preparation for the use of multivariate methods that assume linear space [@RN1646;@RN11563].

```{r outlinessiteraw, out.width = "100%", dpi = 300, echo=TRUE, warning=FALSE}
# gpa ----
Y.gpa <- gpagen(shapesGM, print.progress = FALSE)

# plot
plot(Y.gpa)

# geomorph data frame ----
gdf <- geomorph.data.frame(shape = Y.gpa$coords, 
                           size = Y.gpa$Csize,
                           site = qdata$site)

# add centroid size to qdata
qdata$csz <- Y.gpa$Csize
```

## Principal Components Analysis

Principal components analysis [@RN1746] was used to visualise shape variation among the Perdiz arrow points. The shape changes described by each principal axis are commonly visualized using thin-plate spline warping of a reference 3D mesh [@RN1731;@RN479].

```{r pca.plotsiteraw, out.width = "100%", dpi = 300, echo=TRUE, warning=FALSE}
# principal components analysis ----
pca<-gm.prcomp(Y.gpa$coords)
summary(pca)

# set plot parameters
pch.gps <- c(1:2)[as.factor(qdata$site)]
col.gps <- wes_palette("Moonrise2")[as.factor(qdata$site)]
col.hull <- c("#C27D38", "#798E87")

# plot pca by merged
pc.plot <- plot(pca, 
                asp = 1,
                pch = pch.gps,
                col = col.gps)
shapeHulls(pc.plot, 
           groups = qdata$site,
           group.cols = col.hull)
```

## Procrustes ANOVA

To assess whether shape and size differ by `size class`, Procrustes ANOVAs [@RN1749] were run that enlist effect-sizes (zscores) computed as standard deviates of the generated sampling distributions [@RN1756]. A residual randomization permutation procedure (RRPP; n = 10,000 permutations) was used for all Procrustes ANOVAs [@RN1655;@RN11775], which has higher statistical power and a greater ability to identify patterns in the data should they be present [@RN1719].

```{r anovasiteraw, out.width = "100%", dpi = 300, echo=TRUE, warning=FALSE}
# ANOVA ----
# shape
fit.sh <- procD.lm(shape ~ site,
                   data = gdf,
                   print.progress = FALSE,
                   iter = 9999)

# shape as a function of site/raw material?
anova(fit.sh)

# size
fit.sz <- procD.lm(size ~ site,
                   data = gdf,
                   print.progress = FALSE,
                   iter = 9999)

# size as a function of site/raw material?
anova(fit.sz)
```

## Morphological Disparity

The trajectory analysis was followed by an analysis of morphological disparity [@RN11107;@RN7041;@RN5694] to identify potential differences in morphological diversity between size classes.

```{r mdispsiteraw, out.width = "100%", dpi = 300, echo=TRUE, warning=FALSE}
# morphological disparity ----
# do any of the groups display greater shape variation among 
# individuals relative to the other group?
morphol.disparity(shape ~ site,
                  groups = qdata$site, 
                  data = gdf, 
                  print.progress = FALSE, 
                  iter = 9999)

# do any of the groups display greater size variation among 
# individuals relative to the other group?
morphol.disparity(size ~ site,
                  groups = qdata$site, 
                  data = gdf, 
                  print.progress = FALSE, 
                  iter = 9999)
```

### 2022-08-05 supp/index.Rmd
<init analysis>

--- 
title: "Caddo retouch results in divergent morphology for elliptical bifaces from sites in the American Southeast"
author:
    - Robert Z. Selden, Jr.^[Heritage Research Center, Stephen F. Austin State University; Department of Biology, Stephen F. Austin State University; and Cultural Heritage Department, Jean Monnet University, zselden@sfasu.edu] and John E. Dockall^[Stantec, Inc.]
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
documentclass: book
bibliography: [book.bib]
biblio-style: apalike
link-citations: yes
description: "Pilot study of elliptical biface morphology."
---

# Preface {-}

>Reproducibility---the ability to recompute results---and replicability---the chances other experimenters will achieve a consistent result---are two foundational characteristics of successful scientific research [@RN20918].

Elliptical bifaces from the Millsap Cache and Jowell Farm (41AN13) sites comprise two of the largest samples of this ill-understood lithic tool from the ancestral Caddo area. Similar to Gahagan bifaces, the bifaces from Jowell Farm are assumed to have been manufactured using Edwards chert---due to their size, the colours that they fluoresce beneath low and high frequency UV light, and the paucity of similarly large locally-available raw material sources---and acquired by the Caddo through trade with central Texas groups. Bifaces from the Millsap Cache are thought to have been manufactured from Kay County flint, providing for an empirical assessment of whether elliptical biface morphology differs as a function of raw material and/or biogeographical range. The sample from each site was subsequently divided into two size classes; one reflective of initial production/manufacture (larger), and the other associated with resharpening and/or retouch (smaller). Size classes were used to assess whether modifications by Caddo knappers may have yielded a similar---convergent---knife shape at the smaller size class, and to test the hypothesis that greater morphological variation would be apparent in that size class due to wear associated with regular use, paired with variable idiosyncracies related to resharpening and/or retouch. Results demonstrate that elliptical biface shape does not differ by raw material, but size does. This finding is suggestive of a shared mental template, but a size differential that is potentially driven by raw material and/or biogeographical differences. Additionally, elliptical biface size associated with the large size class does not differ between sites/raw materials; however, the small size class does. This finding aids in advancing the argument that elliptical biface morphology diverges through the resharpening trajectory, where bifaces from each site/raw material enter their uselife at a similar shape and size, but diverge significantly in size---but not shape---throughout that uselife. In following with our expectations, greater shape diversity occurred in the small size class, where the bifaces from Jowell Farm were found to be more diverse than those from Millsap Cache. Discrete biogeographical ranges and raw material differences are posited as the drivers of these extant morphological differences found in elliptical bifaces from the ancestral Caddo area.

```{r figmap, fig.cap="Geological map illustrating the location of the Millsap Cache and Jowell Farm, the level three ecoregions of the Arkansas Valley (blue), Ouachita Mountains (yellow), South Central Plains (orange), and ancestral Caddo area boundary (white) in the American Southeast."}

knitr::include_graphics('./images/map.png')
```

## Analysis overview

Prior to landmarking, elliptical bifaces were oriented with the most heavily retouched edge at top right (Figure \@ref(fig:figlm)). The landmarking protocol uses three landmarks; two horizontal tangents (top/bottom), and one placed at the furthest extent of the area that includes the heaviest amount of retouch; all were applied using the `StereoMorph` package [@RN8973]. Some bifaces include multiple areas of heavy retouch at the top and bottom of the same lateral edge (Figure \@ref(fig:figlm): b, c, d, and e), while others (Figure \@ref(fig:figlm): f) include one retouched edge at top right and another at bottom left.

```{r figlm, fig.cap="Selected elliptical bifaces from the Jowell Farm site, highlighting the most heavily retouched area identified using a modified version of the approach developed and advanced by Miller [-@RN11459], Smallwood [-@RN8310], and Miller and Smallwood [-@RN11460]."}

knitr::include_graphics('./images/ellipticalbifaces.jpg')
```

To identify which edge was most heavily retouched, we employed a modified approach to the flaking index initially developed by Miller [-@RN11459], and later advanced and refined by Smallwood [-@RN8310], then Miller and Smallwood [-@RN11460]. The modified approach uses counts of flake scars from each edge in the two most heavily worked areas, paired with a measure of edge length (not linear/includes curvature). The number of flake scars was subsequently divided by the length of the worked edge, with the heaviest worked area identified by the greater value.

The pattern that emerged suggested that the longest reworked edge on the elliptical bifaces was always more heavily retouched. This finding was used to incorporate the images of elliptical bifaces from the Millsap Cache [@RN11461], where the longest reworked edges were oriented to the top right prior to landmarking.

### Generalised Procrustes Analysis

Landmark data were subsequently aligned to a global coordinate system [@RN11622;@RN11623;@RN11563], achieved through generalized Procrustes superimposition [@RN478] performed in R 4.2.1 [@R] using the `geomorph` library v. 4.0.4 [@RN11530;@RN1774;@RN9565]. Procrustes superimposition translates, scales, and rotates the coordinate data to allow for comparisons among objects [@RN11564;@RN478]. The `geomorph` package uses a partial Procrustes superimposition that projects the aligned specimens into tangent space subsequent to alignment in preparation for the use of multivariate methods that assume linear space [@RN1646;@RN11563].

### Principal Components Analysis

Principal components analysis [@RN1746] was used to visualise shape variation among the elliptical bifaces. The shape changes described by each principal axis are commonly visualized using thin-plate spline warping of a reference image or 3D mesh [@RN1731;@RN479].

```{r figjowellpca, fig.cap="PCA of shape variables for elogated ovate bifaces."}
knitr::include_graphics('./images/pca.png')
```

### Procrustes ANOVA

To assess whether shape and size differed by `raw material` and `size class`, Procrustes ANOVAs [@RN1749] were run that enlist effect-sizes (zscores) computed as standard deviates of the generated sampling distributions [@RN1756]. A residual randomization permutation procedure (RRPP; n = 10,000 permutations) was used for all Procrustes ANOVAs [@RN1655;@RN11775], which has higher statistical power and a greater ability to identify patterns in the data should they be present [@RN1719].

### Morphological disparity

An analysis of morphological disparity was used [@RN11107;@RN7041;@RN5694] to identify potential differences in morphological diversity between size classes.

## Primary findings

  -   Bifaces _do not differ_ in shape, but **differ significantly** in size by raw material/biogeography.
  -   Bifaces in the _large size class_ recovered from the Millsap Cache and Jowell Farm do not differ in shape or size. This suggests consistent morphology in bifaces from the _large size class_, in support of production-based standardization.
  -   Bifaces in the _small size class_ recovered from the Millsap Cache and Jowell Farm **differ significantly** in size. Thus, while shape remains consistent, size differs, suggesting---at minimum---locally divergent approaches to retouch and/or resharpening.
  -   Morphological disparity in the shape of the _small size class_ is **significantly greater** at the Jowell Farm site, providing evidence for a more standardized approach to resharpening and/or retouch at the Millsap Cache site. That finding also supports the hypothesis that greater diversity in shape is associated with retouch and/or resharpening, rather than production/manufacture.

### Summary

The shape of elliptical bifaces does not differ significantly in either size class; however, size does. Results of the Procrustes ANOVA support the hypothesis for divergent morphology. The analysis of morphological disparity yielded support for the interpretation that greater morphological diversity is associated with resharpening and/or retouch, while providing additional support for morphological consistency in production/manufacture.

## Acknowledgments

Our thanks to the Caddo Nation of Oklahoma, the Caddo Nation Tribal Council, Tribal Chairman, and Tribal Historic Preservation Office for permission and access to NAGPRA and previously repatriated collections. Our gratitude is also extended to Marybeth Tomka and Lauren Bussiere at the Texas Archeological Research Laboratory for their assistance with access to the bifaces and the associated records, and to Scott Hammerstedt and Debra K. Green at the Oklahoma Archeological Survey for their assistance with records requests. Thanks also to John Harman for access to the DStretch plugin for ImageJ that was useful in the analysis of flake scars, and to Harry J. Shafer, Hiram F. (Pete) Gregory, Christian S. Hoggard, and David K. Thulman for their comments and constructive criticisms on the ongoing analyses of Caddo biface morphology, and to Emma Sherratt, Kersten Bergstrom, Dean C. Adams, and Michael L. Collyer for their constructive criticisms, general comments, and suggestions throughout the development of this research program.

## Funding

Components of this analytical work flow were developed and funded by a Preservation Technology and Training grant (P14AP00138) to RZS from the National Center for Preservation Technology and Training (NCPTT), and additional grants to RZS from the Caddo Nation of Oklahoma, National Forests and Grasslands in Texas (15-PA-11081300-033) and the United States Forest Service (20-PA-11081300-074). Funding to analyse the bifaces from the Millsap Cache and Jowell Farm was provided by the Heritage Research Center at Stephen F. Austin State University.

## Data management

The analysis code associated with this project can be accessed through this document, is available in the GitHub repository, and digitally curated on the Open Science Framework. The reproducible nature of this enterprise provides a means for others to critically assess and evaluate the various analytical components [@RN20915;@RN20916;@RN20917], which is a necessary requirement for the production of reliable knowledge.

Reproducibility projects in [psychology](https://osf.io/ezcuj/) and [cancer biology](https://www.cos.io/rpcb) are impacting current research practices across all domains. Examples of reproducible research are becoming more abundant in archaeology [@RN20804;@RN21009;@RN11783;@RN21001;@RN9364], and the next generation of archaeologists are learning those tools and methods needed to reproduce and/or replicate research results [@RN21007]. Reproducible and replicable research work flows are often employed at the highest levels of humanities-based inquiries to mitigate concern or doubt regarding proper execution, and is of particular import should the results have---explicitly or implicitly---a major impact on scientific progress [@RN21008].

## Colophon

This version of the analysis was generated on `r Sys.time()` using the following computational environment and dependencies:

```{r colophon, cache = FALSE}
# what R packages and versions were used?
if ("devtools" %in% installed.packages()) devtools::session_info()
```

Current Git commit details are:

```{r}
# where can I find this commit? 
if ("git2r" %in% installed.packages() & git2r::in_repository(path = ".")) git2r::repository(here::here())  
```

### 2022-08-05 supp/index.Rmd
<edit readme>

@@
 --- 
-title: "Caddo retouch results in divergent morphology for elliptical bifaces from sites in the American Southeast"
+title: "Divergent morphology of elliptical bifaces from sites in the American Southeast articulates with raw material differences"
 author:
@@
 Elliptical bifaces from the Millsap Cache and Jowell Farm (41AN13) sites comprise two of the largest samples of this ill-understood lithic tool from the ancestral Caddo area.
-Similar to Gahagan bifaces, the bifaces from Jowell Farm are assumed to have been manufactured using Edwards chert---due to their size, the colours that they fluoresce beneath low and high frequency UV light, and the paucity of similarly large locally-available raw material sources---and acquired by the Caddo through trade with central Texas groups.
-Bifaces from the Millsap Cache are thought to have been manufactured from Kay County flint, providing for an empirical assessment of whether elliptical biface morphology differs as a function of raw material and/or biogeographical range.
+Similar to Gahagan bifaces, those bifaces from Jowell Farm are assumed to have been manufactured using Edwards chert---due to their size, the colours that they fluoresce beneath low and high frequency UV light, and the paucity of similarly large locally-available raw material sources---and acquired by the Caddo through trade with central Texas groups.
+Bifaces from the Millsap Cache are thought to have been manufactured from Kay County flint, providing for an empirical assessment of whether elliptical biface morphology differs as a function of raw material.
 The sample from each site was subsequently divided into two size classes; one reflective of initial production/manufacture (larger), and the other associated with resharpening and/or retouch (smaller).
-Size classes were used to assess whether modifications by Caddo knappers may have yielded a similar---convergent---knife shape at the smaller size class, and to test the hypothesis that greater morphological variation would be apparent in that size class due to wear associated with regular use, paired with variable idiosyncracies related to resharpening and/or retouch.
+Size classes were used to assess whether modifications by Caddo knappers may have yielded a similar---convergent---knife shape at the smaller size class, and to test the hypothesis that greater morphological variation would be apparent in the small size class due to the variable idiosyncracies related to resharpening and/or retouch.
 Results demonstrate that elliptical biface shape does not differ by raw material, but size does.
-This finding is suggestive of a shared mental template, but a size differential that is potentially driven by raw material and/or biogeographical differences.
-Additionally, elliptical biface size associated with the large size class does not differ between sites/raw materials; however, the small size class does.
-This finding aids in advancing the argument that elliptical biface morphology diverges through the resharpening trajectory, where bifaces from each site/raw material enter their uselife at a similar shape and size, but diverge significantly in size---but not shape---throughout that uselife.
-In following with our expectations, greater shape diversity occurred in the small size class, where the bifaces from Jowell Farm were found to be more diverse than those from Millsap Cache.
-Discrete biogeographical ranges and raw material differences are posited as the drivers of these extant morphological differences found in elliptical bifaces from the ancestral Caddo area.
+This finding suggests both a shared mental template, and a size differential is potentially driven by raw material.
+Additionally, elliptical biface size in the large size class does not differ between raw materials; however, the small size class does.
+This finding aids in advancing the argument that elliptical biface morphology diverges through the resharpening trajectory, where bifaces from each site/raw material enter their uselife at a similar shape and size, diverging significantly in size---but not shape---throughout that uselife.
+In following with our expectations, greater shape diversity was found to occur in the small size class, where those bifaces from Jowell Farm were more diverse than those from the Millsap Cache.
+Discrete raw material differences are posited as the drivers of the extant morphological differences found in elliptical bifaces from the ancestral Caddo area.
 
@@
 
-The analysis code associated with this project can be accessed through this document, is available in the GitHub repository, and digitally curated on the Open Science Framework.
+The analysis code associated with this project can be accessed through this document, is available in the GitHub repository [https://github.com/seldenlab/elliptical.bifaces](https://github.com/seldenlab/elliptical.bifaces), and digitally curated on the Open Science Framework [DOI 10.17605/OSF.IO/PH25W](https://osf.io/ph25w/).
 The reproducible nature of this enterprise provides a means for others to critically assess and evaluate the various analytical components [@RN20915;@RN20916;@RN20917], which is a necessary requirement for the production of reliable knowledge.

### 2022-08-05 supp/index.Rmd
<edit index>

@@
 
-```{r figlm, fig.cap="Selected elliptical bifaces from the Jowell Farm site, highlighting the most heavily retouched area identified using a modified version of the approach developed and advanced by Miller [-@RN11459], Smallwood [-@RN8310], and Miller and Smallwood [-@RN11460]."}
+```{r figlm, fig.cap="Selected elliptical bifaces from the Jowell Farm site highlighting the most heavily retouched area identified using a modified version of the approach developed and advanced by Miller [-@RN11459], Smallwood [-@RN8310], and Miller and Smallwood [-@RN11460]."}
 
@@
 
-```{r figjowellpca, fig.cap="PCA of shape variables for elogated ovate bifaces."}
+```{r figjowellpca, fig.cap="PCA of shape variables for elliptical bifaces."}
 knitr::include_graphics('./images/pca.png')
@@
 
-The analysis code associated with this project can be accessed through this document, is available in the GitHub repository [https://github.com/seldenlab/elliptical.bifaces](https://github.com/seldenlab/elliptical.bifaces), and digitally curated on the Open Science Framework [DOI 10.17605/OSF.IO/PH25W](https://osf.io/ph25w/).
+The analysis code associated with this project can be accessed through this document, is available in the GitHub repository ([https://github.com/seldenlab/elliptical.bifaces](https://github.com/seldenlab/elliptical.bifaces)), and digitally curated on the Open Science Framework ([DOI 10.17605/OSF.IO/PH25W](https://osf.io/ph25w/)).
 The reproducible nature of this enterprise provides a means for others to critically assess and evaluate the various analytical components [@RN20915;@RN20916;@RN20917], which is a necessary requirement for the production of reliable knowledge.
@@
 Reproducibility projects in [psychology](https://osf.io/ezcuj/) and [cancer biology](https://www.cos.io/rpcb) are impacting current research practices across all domains.
-Examples of reproducible research are becoming more abundant in archaeology [@RN20804;@RN21009;@RN11783;@RN21001;@RN9364], and the next generation of archaeologists are learning those tools and methods needed to reproduce and/or replicate research results [@RN21007].
+Examples of reproducible research are becoming more abundant in archaeology [@RN20804;@RN21009;@RN11783;@RN21001;@RN9364;@RN11264], and the next generation of archaeologists are learning those tools and methods needed to reproduce and/or replicate research results [@RN21007].
 Reproducible and replicable research work flows are often employed at the highest levels of humanities-based inquiries to mitigate concern or doubt regarding proper execution, and is of particular import should the results have---explicitly or implicitly---a major impact on scientific progress [@RN21008].

### 2022-08-05 supp/index.Rmd
<edit index>

@@
 Elliptical bifaces from the Millsap Cache and Jowell Farm (41AN13) sites comprise two of the largest samples of this ill-understood lithic tool from the ancestral Caddo area.
-Similar to Gahagan bifaces, those bifaces from Jowell Farm are assumed to have been manufactured using Edwards chert---due to their size, the colours that they fluoresce beneath low and high frequency UV light, and the paucity of similarly large locally-available raw material sources---and acquired by the Caddo through trade with central Texas groups.
+Similar to Gahagan bifaces, elliptical bifaces from Jowell Farm are assumed to have been manufactured from Edwards chert---due to their size, the colours that they fluoresce beneath low and high frequency UV light, and the paucity of similarly large locally-available raw material sources---and acquired by the Caddo through trade with central Texas groups.
 Bifaces from the Millsap Cache are thought to have been manufactured from Kay County flint, providing for an empirical assessment of whether elliptical biface morphology differs as a function of raw material.
 The sample from each site was subsequently divided into two size classes; one reflective of initial production/manufacture (larger), and the other associated with resharpening and/or retouch (smaller).
-Size classes were used to assess whether modifications by Caddo knappers may have yielded a similar---convergent---knife shape at the smaller size class, and to test the hypothesis that greater morphological variation would be apparent in the small size class due to the variable idiosyncracies related to resharpening and/or retouch.
+Size classes were used to assess whether modifications by Caddo knappers may have yielded a similar---convergent---knife shape in the smaller size class, and to test the hypothesis that greater morphological variation would be apparent in the small size class due to the variable idiosyncracies related to resharpening and/or retouch.
 Results demonstrate that elliptical biface shape does not differ by raw material, but size does.
-This finding suggests both a shared mental template, and a size differential is potentially driven by raw material.
-Additionally, elliptical biface size in the large size class does not differ between raw materials; however, the small size class does.
-This finding aids in advancing the argument that elliptical biface morphology diverges through the resharpening trajectory, where bifaces from each site/raw material enter their uselife at a similar shape and size, diverging significantly in size---but not shape---throughout that uselife.
-In following with our expectations, greater shape diversity was found to occur in the small size class, where those bifaces from Jowell Farm were more diverse than those from the Millsap Cache.
-Discrete raw material differences are posited as the drivers of the extant morphological differences found in elliptical bifaces from the ancestral Caddo area.
+This finding suggests a shared mental template, coupled with a size differential driven by raw material differences.
+Additionally, size does not differ between raw materials in the large size class; however, in the small size class, it does.
+This finding aids in advancing the argument that elliptical biface morphology diverges through the resharpening lifecycle, where bifaces manufactured using both raw materials enter their uselife at a similar shape and size, but diverge significantly in size---but not shape---throughout that uselife.
+In following with our expectations, greater shape diversity was found to occur in the small size class, where bifaces from Jowell Farm are more diverse than those from the Millsap Cache.
+Discrete raw material differences are advanced as the drivers of the extant morphological differences found in elliptical bifaces from the ancestral Caddo area in the American Southeast.
 

### 2022-08-05 supp/02-2dgm-rawmat.Rmd
<edit index>

@@
 
-# plot
-plot(Y.gpa)
-
 # geomorph data frame ----
@@
            group.cols = col.hull)
+```
+
+### Minima/maxima of PC1/2 with warp grids
+
+```{r minmaxraw, echo=TRUE, out.width = "100%", dpi = 300, warning=FALSE}
+# plot x/y maxima/minima
+## x - minima
+mean.shape <- mshape(Y.gpa$coords)
+plotRefToTarget(pca$shapes$shapes.comp1$min, 
+                mean.shape)
+
+## x - maxima
+plotRefToTarget(pca$shapes$shapes.comp1$max, 
+                mean.shape)
+
+## y - minima
+plotRefToTarget(pca$shapes$shapes.comp2$min, 
+                mean.shape)
+
+## y - maxima
+plotRefToTarget(pca$shapes$shapes.comp2$max, 
+                mean.shape)
 ```

### 2022-08-05 supp/03-2dgm.Rmd
<edit index>

--- 2022-08-05 supp/02-2dgm-rawmat.Rmd
@@
-# GM - Site/Raw Material
+# GM - Size Class
 
-Elliptical bifaces from the Millsap Cache and Jowell Farm were used to examine whether their morphology remains stable, or expresses morphological variability by site/raw meterial.
-Variability in elliptical biface morphology provides evidence associated with differences in morphology as a function of geography/raw material.
+Two `size classes` (large/small) were used to examine whether the morphology of the elliptical bifaces from the Millsap Cache and Jowell Farm remain morphologically stable, or express morphological variation.
+Variability in elliptical biface morphology potentially provides evidence associated with differential resharpening trajectories.
 
@@
 
-```{r load.packagessiteraw, out.width = "100%", dpi = 300, echo=TRUE, warning=FALSE}
+```{r load.packages, out.width = "100%", dpi = 300, echo=TRUE, warning=FALSE}
 # load required analysis packages
@@
 
-```{r outlinessiteraw, out.width = "100%", dpi = 300, echo=TRUE, warning=FALSE}
+```{r outlines, out.width = "100%", dpi = 300, echo=TRUE, warning=FALSE}
 # gpa ----
@@
                            size = Y.gpa$Csize,
-                           site = qdata$site)
+                           site = qdata$site,
+                           size.class = qdata$size,
+                           merged = qdata$merged)
 
@@
 
-```{r pca.plotsiteraw, out.width = "100%", dpi = 300, echo=TRUE, warning=FALSE}
+```{r pca.plot, out.width = "100%", dpi = 300, echo=TRUE, warning=FALSE}
 # principal components analysis ----
@@
 # set plot parameters
-pch.gps <- c(1:2)[as.factor(qdata$site)]
-col.gps <- wes_palette("Moonrise2")[as.factor(qdata$site)]
-col.hull <- c("#C27D38", "#798E87")
+pch.gps <- c(1:4)[as.factor(qdata$merged)]
+col.gps <- wes_palette("Moonrise2")[as.factor(qdata$merged)]
+col.hull <- c("#CCC591", "#29211F", "#798E87", "#C27D38")
 
@@
 shapeHulls(pc.plot, 
-           groups = qdata$site,
+           groups = qdata$merged,
            group.cols = col.hull)
@@
 
-```{r minmaxraw, echo=TRUE, out.width = "100%", dpi = 300, warning=FALSE}
+```{r min.max, echo=TRUE, out.width = "100%", dpi = 300, warning=FALSE}
 # plot x/y maxima/minima
@@
 
-```{r anovasiteraw, out.width = "100%", dpi = 300, echo=TRUE, warning=FALSE}
+```{r anova, out.width = "100%", dpi = 300, echo=TRUE, warning=FALSE}
 # ANOVA ----
 # shape
-fit.sh <- procD.lm(shape ~ site,
+fit.sh <- procD.lm(shape ~ merged,
                    data = gdf,
@@
 
-# shape as a function of site/raw material?
+# shape as a function of size class?
 anova(fit.sh)
@@
 # size
-fit.sz <- procD.lm(size ~ site,
+fit.sz <- procD.lm(size ~ merged,
                    data = gdf,
@@
 
-# size as a function of site/raw material?
+# size as a function of size class?
 anova(fit.sz)
+
+# pairwise comparison of LS means = which differ?
+pw.sz <- pairwise(fit.sz,
+                  groups = qdata$merged)
+summary(pw.sz, 
+        confidence = 0.95, 
+        test.type = "dist")
 ```
@@
 
-The trajectory analysis was followed by an analysis of morphological disparity [@RN11107;@RN7041;@RN5694] to identify potential differences in morphological diversity between size classes.
+An analysis of morphological disparity was used [@RN11107;@RN7041;@RN5694] to identify potential differences in morphological diversity between the size classes.
 
-```{r mdispsiteraw, out.width = "100%", dpi = 300, echo=TRUE, warning=FALSE}
+```{r mdisp, out.width = "100%", dpi = 300, echo=TRUE, warning=FALSE}
 # morphological disparity ----
@@
 # individuals relative to the other group?
-morphol.disparity(shape ~ site,
-                  groups = qdata$site, 
+morphol.disparity(shape ~ merged,
+                  groups = qdata$merged, 
                   data = gdf, 
@@
 # individuals relative to the other group?
-morphol.disparity(size ~ site,
-                  groups = qdata$site, 
+morphol.disparity(size ~ merged,
+                  groups = qdata$merged, 
                   data = gdf, 

### 2022-08-05 supp/index.Rmd
<edit index>

@@
 Prior to landmarking, elliptical bifaces were oriented with the most heavily retouched edge at top right (Figure \@ref(fig:figlm)).
-The landmarking protocol uses three landmarks; two horizontal tangents (top/bottom), and one placed at the furthest extent of the area that includes the heaviest amount of retouch; all were applied using the `StereoMorph` package [@RN8973].
+The landmarking protocol uses three landmarks; two horizontal tangents (top/bottom), and one placed at the furthest extent of the edge including the heaviest amount of retouch.
+Landmarks and semilandmarks were applied using the `StereoMorph` package [@RN8973].
 Some bifaces include multiple areas of heavy retouch at the top and bottom of the same lateral edge (Figure \@ref(fig:figlm): b, c, d, and e), while others (Figure \@ref(fig:figlm): f) include one retouched edge at top right and another at bottom left.
@@
 To identify which edge was most heavily retouched, we employed a modified approach to the flaking index initially developed by Miller [-@RN11459], and later advanced and refined by Smallwood [-@RN8310], then Miller and Smallwood [-@RN11460].
-The modified approach uses counts of flake scars from each edge in the two most heavily worked areas, paired with a measure of edge length (not linear/includes curvature).
+The modified approach uses counts of flake scars from each edge in the two most heavily worked areas, paired with a measure of edge length (inclusive of curvature - not linear).
 The number of flake scars was subsequently divided by the length of the worked edge, with the heaviest worked area identified by the greater value.
@@
 The pattern that emerged suggested that the longest reworked edge on the elliptical bifaces was always more heavily retouched.
-This finding was used to incorporate the images of elliptical bifaces from the Millsap Cache [@RN11461], where the longest reworked edges were oriented to the top right prior to landmarking.
+This finding was used to incorporate the images of elliptical bifaces from the Millsap Cache [@RN11461], where the longest reworked edges were oriented at top right prior to landmarking.
 
@@
 
+```{r figgpa, out.width = "100%", dpi = 300, fig.cap="Results of generalised Procrustes analysis for elliptical bifaces from the Millsap Cache and Jowell Farm.
+Mean consensus configuration shown in black; individual bifaces in gray."}
+knitr::include_graphics('./images/gpa.png')
+```
+
 ### Principal Components Analysis
@@
 
-```{r figjowellpca, fig.cap="PCA of shape variables for elliptical bifaces."}
+```{r figpca, out.width = "100%", dpi = 300, fig.cap="PCA summarizing shape variation in elliptical bifaces."}
 knitr::include_graphics('./images/pca.png')

### 2022-08-05 supp/index.Rmd
<edit index>

@@
 Bifaces from the Millsap Cache are thought to have been manufactured from Kay County flint, providing for an empirical assessment of whether elliptical biface morphology differs as a function of raw material.
-The sample from each site was subsequently divided into two size classes; one reflective of initial production/manufacture (larger), and the other associated with resharpening and/or retouch (smaller).
-Size classes were used to assess whether modifications by Caddo knappers may have yielded a similar---convergent---knife shape in the smaller size class, and to test the hypothesis that greater morphological variation would be apparent in the small size class due to the variable idiosyncracies related to resharpening and/or retouch.
+The sample from each site was subsequently divided into two size classes; one reflective of initial production/manufacture (large), and the other associated with resharpening and/or retouch (small).
+Size classes were used to assess whether modifications by Caddo knappers may have yielded a similar---convergent---biface shape in the small class, and to test the hypothesis that greater morphological variation would be apparent in this class due to the variable idiosyncracies related to resharpening and/or retouch.
 Results demonstrate that elliptical biface shape does not differ by raw material, but size does.
 This finding suggests a shared mental template, coupled with a size differential driven by raw material differences.
-Additionally, size does not differ between raw materials in the large size class; however, in the small size class, it does.
-This finding aids in advancing the argument that elliptical biface morphology diverges through the resharpening lifecycle, where bifaces manufactured using both raw materials enter their uselife at a similar shape and size, but diverge significantly in size---but not shape---throughout that uselife.
-In following with our expectations, greater shape diversity was found to occur in the small size class, where bifaces from Jowell Farm are more diverse than those from the Millsap Cache.
-Discrete raw material differences are advanced as the drivers of the extant morphological differences found in elliptical bifaces from the ancestral Caddo area in the American Southeast.
+Additionally, size does not differ between raw materials in the large class; however, in the small class, it does.
+This finding aids in advancing the argument that elliptical biface morphology diverges through the resharpening lifecycle, where bifaces manufactured using both raw materials enter use at a similar shape and size, but diverge significantly in size---but not shape---through their uselife.
+In following with expectations, greater shape diversity was found to occur in the small class, where bifaces from Jowell Farm are more diverse than those from the Millsap Cache.
+Discrete raw material differences are advanced as the drivers of the extant morphological differences found in elliptical bifaces from the ancestral Caddo area of the American Southeast.
 

### 2022-08-05 supp/index.Rmd
<edit index>

@@
 The sample from each site was subsequently divided into two size classes; one reflective of initial production/manufacture (large), and the other associated with resharpening and/or retouch (small).
-Size classes were used to assess whether modifications by Caddo knappers may have yielded a similar---convergent---biface shape in the small class, and to test the hypothesis that greater morphological variation would be apparent in this class due to the variable idiosyncracies related to resharpening and/or retouch.
+Size classes were used to assess whether modifications by Caddo knappers may have yielded a similar---convergent---biface shape in the small class, and to test the hypothesis that greater morphological variation would be apparent in this class due to idiosyncratic responses related to resharpening and/or retouch.
 Results demonstrate that elliptical biface shape does not differ by raw material, but size does.
@@
 Additionally, size does not differ between raw materials in the large class; however, in the small class, it does.
-This finding aids in advancing the argument that elliptical biface morphology diverges through the resharpening lifecycle, where bifaces manufactured using both raw materials enter use at a similar shape and size, but diverge significantly in size---but not shape---through their uselife.
+This finding aids in advancing the argument that elliptical biface morphology diverges through the resharpening lifecycle, where bifaces manufactured using both raw materials enter use at a similar shape and size, but diverge significantly in size---but not shape---through that uselife.
 In following with expectations, greater shape diversity was found to occur in the small class, where bifaces from Jowell Farm are more diverse than those from the Millsap Cache.

### 2022-08-05 supp/index.Rmd
<edit index>

@@
 Additionally, size does not differ between raw materials in the large class; however, in the small class, it does.
-This finding aids in advancing the argument that elliptical biface morphology diverges through the resharpening lifecycle, where bifaces manufactured using both raw materials enter use at a similar shape and size, but diverge significantly in size---but not shape---through that uselife.
+This finding aids in advancing the argument that elliptical biface morphology diverges through the resharpening lifecycle, where bifaces manufactured using both raw materials enter use at a similar shape and size, but diverge significantly in size---but not shape---through their uselife.
 In following with expectations, greater shape diversity was found to occur in the small class, where bifaces from Jowell Farm are more diverse than those from the Millsap Cache.

### 2022-08-05 supp/index.Rmd
<edit index>

@@
 
-Elliptical bifaces from the Millsap Cache and Jowell Farm (41AN13) sites comprise two of the largest samples of this ill-understood lithic tool from the ancestral Caddo area.
-Similar to Gahagan bifaces, elliptical bifaces from Jowell Farm are assumed to have been manufactured from Edwards chert---due to their size, the colours that they fluoresce beneath low and high frequency UV light, and the paucity of similarly large locally-available raw material sources---and acquired by the Caddo through trade with central Texas groups.
+Elliptical bifaces from Millsap Cache and Jowell Farm (41AN13) comprise two of the largest samples of this ill-understood lithic tool from the ancestral Caddo area.
+Similar to Gahagan bifaces, elliptical bifaces from Jowell Farm are assumed to have been manufactured from Edwards chert---due to their size, the colours that they fluoresce beneath low and high frequency UV light, and the paucity of similarly large locally-available raw material sources---and acquired by the Caddo through trade and/or exchange with central Texas groups.
 Bifaces from the Millsap Cache are thought to have been manufactured from Kay County flint, providing for an empirical assessment of whether elliptical biface morphology differs as a function of raw material.
@@
 This finding aids in advancing the argument that elliptical biface morphology diverges through the resharpening lifecycle, where bifaces manufactured using both raw materials enter use at a similar shape and size, but diverge significantly in size---but not shape---through their uselife.
-In following with expectations, greater shape diversity was found to occur in the small class, where bifaces from Jowell Farm are more diverse than those from the Millsap Cache.
+In following with expectations, greater shape diversity was found to occur in the small class, where bifaces from Jowell Farm are more morphologically diverse than those from the Millsap Cache.
 Discrete raw material differences are advanced as the drivers of the extant morphological differences found in elliptical bifaces from the ancestral Caddo area of the American Southeast.

### 2022-08-06 supp/03-2dgm-sizeclass.Rmd
<edit index>

--- 2022-08-05 supp/03-2dgm.Rmd
@@
 
-Two `size classes` (large/small) were used to examine whether the morphology of the elliptical bifaces from the Millsap Cache and Jowell Farm remain morphologically stable, or express morphological variation.
-Variability in elliptical biface morphology potentially provides evidence associated with differential resharpening trajectories.
+Two `size classes` (large/small) were used to assess whether the morphology of elliptical bifaces from Millsap Cache and Jowell Farm remain morphologically stable through reduction.
+Variability in elliptical biface morphology through reduction provides evidence associated with differential resharpening trajectories.
 

### 2022-08-06 supp/index.Rmd
<edit index>

@@
 --- 
-title: "Divergent morphology of elliptical bifaces from sites in the American Southeast articulates with raw material differences"
+title: "Size matters: Divergent morphology in elliptical bifaces from sites in the American Southeast articulates with distinct reduction practices and raw material differences"
 author:
@@
 The sample from each site was subsequently divided into two size classes; one reflective of initial production/manufacture (large), and the other associated with resharpening and/or retouch (small).
-Size classes were used to assess whether modifications by Caddo knappers may have yielded a similar---convergent---biface shape in the small class, and to test the hypothesis that greater morphological variation would be apparent in this class due to idiosyncratic responses related to resharpening and/or retouch.
+Size classes were used to assess whether modifications by Caddo knappers may have yielded a similar---convergent---biface shape in the small class, and to test the hypothesis that greater morphological variation would be apparent due to idiosyncratic responses related to resharpening and/or retouch.
 Results demonstrate that elliptical biface shape does not differ by raw material, but size does.
-This finding suggests a shared mental template, coupled with a size differential driven by raw material differences.
-Additionally, size does not differ between raw materials in the large class; however, in the small class, it does.
+This finding suggests a shared mental template, coupled with a size differential driven by raw material differences and reduction practices.
+The analysis of biface morphlogy by size class demonstrates that size does not differ between raw materials in the large class, but does in the small class.
 This finding aids in advancing the argument that elliptical biface morphology diverges through the resharpening lifecycle, where bifaces manufactured using both raw materials enter use at a similar shape and size, but diverge significantly in size---but not shape---through their uselife.
 In following with expectations, greater shape diversity was found to occur in the small class, where bifaces from Jowell Farm are more morphologically diverse than those from the Millsap Cache.
-Discrete raw material differences are advanced as the drivers of the extant morphological differences found in elliptical bifaces from the ancestral Caddo area of the American Southeast.
+Distinct reduction practices and raw material differences are advanced as the drivers of the extant morphological differences found in elliptical bifaces from the ancestral Caddo area of the American Southeast.
 
@@
 To identify which edge was most heavily retouched, we employed a modified approach to the flaking index initially developed by Miller [-@RN11459], and later advanced and refined by Smallwood [-@RN8310], then Miller and Smallwood [-@RN11460].
-The modified approach uses counts of flake scars from each edge in the two most heavily worked areas, paired with a measure of edge length (inclusive of curvature - not linear).
+The modified approach uses counts of flake scars from each edge in the two most heavily worked areas, paired with a measure of edge length (inclusive of curvature - not a linear metric).
 The number of flake scars was subsequently divided by the length of the worked edge, with the heaviest worked area identified by the greater value.
@@
 Thus, while shape remains consistent, size differs, suggesting---at minimum---locally divergent approaches to retouch and/or resharpening.
-  -   Morphological disparity in the shape of the _small size class_ is **significantly greater** at the Jowell Farm site, providing evidence for a more standardized approach to resharpening and/or retouch at the Millsap Cache site.
-That finding also supports the hypothesis that greater diversity in shape is associated with retouch and/or resharpening, rather than production/manufacture.
+  -   Morphological disparity in the shape of the _small size class_ is **significantly greater** at Jowell Farm, providing evidence for a more standardized approach to resharpening and/or retouch at Millsap Cache.
+This finding also provides support for the hypothesis that greater shape diversity is associated with retouch and/or resharpening for elliptical bifaces, rather than production/manufacture.
 

### 2022-08-06 supp/01-size.class.Rmd
<edit index>

@@
 shapes <- readShapes("shapes")
-shapesGM <- readland.shapes(shapes, nCurvePts = c(15, 10, 15))
+shapesGM <- readland.shapes(shapes, nCurvePts = c(15, 20, 35))
 
@@
 ## Assign size class
+
+Centroid size---the square root of the sum of squared distances of all landmarks from the centroid---is the measure of size used in landmark geometric morphometrics [@RN11462].
+The mean centroid size was identified for elliptical bifaces from each site.
+All bifaces with centroid sizes larger than the mean were assigned to the `large class`; those with centroid sizes below the mean were assigned to the `small class`.
 

### 2022-08-06 supp/index.Rmd
<edit index>

@@
 
-An analysis of morphological disparity was used [@RN11107;@RN7041;@RN5694] to identify potential differences in morphological diversity between size classes.
+An analysis of morphological disparity was used to identify potential differences in morphological diversity between size classes [@RN11107;@RN7041;@RN5694].
 

### 2022-08-07 supp/index.Rmd
<edit index>

@@
 --- 
-title: "Size matters: Divergent morphology in elliptical bifaces from sites in the American Southeast articulates with distinct reduction practices and raw material differences"
+title: "Size matters: Divergent morphology in elliptical bifaces from sites in the American Southeast articulates with distinct reduction practices"
 author:
@@
 
-Elliptical bifaces from Millsap Cache and Jowell Farm (41AN13) comprise two of the largest samples of this ill-understood lithic tool from the ancestral Caddo area.
-Similar to Gahagan bifaces, elliptical bifaces from Jowell Farm are assumed to have been manufactured from Edwards chert---due to their size, the colours that they fluoresce beneath low and high frequency UV light, and the paucity of similarly large locally-available raw material sources---and acquired by the Caddo through trade and/or exchange with central Texas groups.
-Bifaces from the Millsap Cache are thought to have been manufactured from Kay County flint, providing for an empirical assessment of whether elliptical biface morphology differs as a function of raw material.
-The sample from each site was subsequently divided into two size classes; one reflective of initial production/manufacture (large), and the other associated with resharpening and/or retouch (small).
-Size classes were used to assess whether modifications by Caddo knappers may have yielded a similar---convergent---biface shape in the small class, and to test the hypothesis that greater morphological variation would be apparent due to idiosyncratic responses related to resharpening and/or retouch.
+Elliptical bifaces from Millsap Cache and Jowell Farm (41AN13) comprise two of the largest samples of this ill-understood object from the ancestral Caddo area.
+Elliptical bifaces from Jowell Farm were manufactured using Edwards chert, and thought to have been acquired by the Caddo through trade and/or exchange with central Texas groups.
+Bifaces from the Millsap Cache were reportedly manufactured from Kay County flint, providing for an empirical assessment of whether biface morphology differs as a function of raw material.
+The sample of elliptical bifaces from each site was divided into two size classes; one reflective of initial production/manufacture (large), and the other associated with resharpening and/or retouch (small).
+Size classes were used to assess whether modifications by Caddo knappers may have yielded similar---convergent---biface shape in the small class.
+They were also used to test the hypothesis that greater morphological variation would be apparent in the small class due to idiosyncratic responses related to resharpening and/or retouch practices.
 Results demonstrate that elliptical biface shape does not differ by raw material, but size does.
-This finding suggests a shared mental template, coupled with a size differential driven by raw material differences and reduction practices.
-The analysis of biface morphlogy by size class demonstrates that size does not differ between raw materials in the large class, but does in the small class.
-This finding aids in advancing the argument that elliptical biface morphology diverges through the resharpening lifecycle, where bifaces manufactured using both raw materials enter use at a similar shape and size, but diverge significantly in size---but not shape---through their uselife.
-In following with expectations, greater shape diversity was found to occur in the small class, where bifaces from Jowell Farm are more morphologically diverse than those from the Millsap Cache.
-Distinct reduction practices and raw material differences are advanced as the drivers of the extant morphological differences found in elliptical bifaces from the ancestral Caddo area of the American Southeast.
+This finding suggests a shared mental template, coupled with a size differential.
+The subsequent analysis of biface morphology by size class demonstrated that size does not differ by raw material in the large class, but does in the small class.
+This finding supports the argument that elliptical biface morphology diverges through local resharpening and/or retouch practices, where bifaces manufactured using both raw materials enter their uselife at similar shapes and sizes prior to diverging in size.
+As expected, greater shape diversity occurs in the small class, where the Jowell Farm bifaces were found to be more morphologically diverse than those from the Millsap Cache.
+Distinct reduction practices are advanced as the driver of extant morphological differences found in elliptical bifaces from the ancestral Caddo area of the American Southeast.
 
@@
 The modified approach uses counts of flake scars from each edge in the two most heavily worked areas, paired with a measure of edge length (inclusive of curvature - not a linear metric).
-The number of flake scars was subsequently divided by the length of the worked edge, with the heaviest worked area identified by the greater value.
+The number of flake scars was then divided by the length of the worked edge, with the heaviest worked area identified by the greater value.
 
@@
 
-Landmark data were subsequently aligned to a global coordinate system [@RN11622;@RN11623;@RN11563], achieved through generalized Procrustes superimposition [@RN478] performed in R 4.2.1 [@R] using the `geomorph` library v.
+Landmark data were aligned to a global coordinate system [@RN11622;@RN11623;@RN11563], achieved through generalized Procrustes superimposition [@RN478] performed in R 4.2.1 [@R] using the `geomorph` library v.
 4.0.4 [@RN11530;@RN1774;@RN9565].
 Procrustes superimposition translates, scales, and rotates the coordinate data to allow for comparisons among objects [@RN11564;@RN478].
-The `geomorph` package uses a partial Procrustes superimposition that projects the aligned specimens into tangent space subsequent to alignment in preparation for the use of multivariate methods that assume linear space [@RN1646;@RN11563].
+The `geomorph` package uses a partial Procrustes superimposition that projects the aligned specimens into tangent space following alignment in preparation for the use of multivariate methods that assume linear space [@RN1646;@RN11563].
 

### 2022-08-07 supp/02-2dgm-rawmat.Rmd
<edit index>

@@
 
+## Procrustes ANOVA: Allometry
+
+To assess whether shape changes with size (allometry), Procrustes ANOVAs [@RN1749] were run that enlist effect-sizes (zscores) computed as standard deviates of the generated sampling distributions [@RN1756].
+A residual randomization permutation procedure (RRPP; n = 10,000 permutations) was used for all Procrustes ANOVAs [@RN1655;@RN11775], which has higher statistical power and a greater ability to identify patterns in the data should they be present [@RN1719].
+
+```{r allomgm1, echo=TRUE, out.width = "100%", dpi = 300, warning=FALSE}
+# allometry
+fit.size <- procD.lm(shape ~ size, 
+                     data = gdf,
+
+                     print.progress = FALSE, 
+                     iter = 9999)
+
+# allometry
+anova(fit.size)
+```
+
 ## Procrustes ANOVA

### 2022-08-07 supp/index.Rmd
<edit index>

@@
 --- 
-title: "Size matters: Divergent morphology in elliptical bifaces from sites in the American Southeast articulates with distinct reduction practices"
+title: "Size matters: Divergent morphology in elliptical bifaces from sites in the American Southeast articulates with distinct local reduction practices"
 author:
@@
 
-Elliptical bifaces from Millsap Cache and Jowell Farm (41AN13) comprise two of the largest samples of this ill-understood object from the ancestral Caddo area.
-Elliptical bifaces from Jowell Farm were manufactured using Edwards chert, and thought to have been acquired by the Caddo through trade and/or exchange with central Texas groups.
-Bifaces from the Millsap Cache were reportedly manufactured from Kay County flint, providing for an empirical assessment of whether biface morphology differs as a function of raw material.
-The sample of elliptical bifaces from each site was divided into two size classes; one reflective of initial production/manufacture (large), and the other associated with resharpening and/or retouch (small).
+Elliptical bifaces from Millsap Cache and Jowell Farm (41AN13) comprise two of the largest samples of this ill-understood lithic artefact from the ancestral Caddo area.
+Elliptical bifaces from Jowell Farm were manufactured using Edwards chert, and are thought to have been acquired by the Caddo through trade or exchange with central Texas groups.
+Bifaces from Millsap Cache were reportedly manufactured from Kay County flint, providing for an empirical test of whether biface morphology differs as a function of raw material.
+The sample of elliptical bifaces from each site was also divided into two size classes; one reflective of initial production (large), and the other associated with retouch (small).
 Size classes were used to assess whether modifications by Caddo knappers may have yielded similar---convergent---biface shape in the small class.
-They were also used to test the hypothesis that greater morphological variation would be apparent in the small class due to idiosyncratic responses related to resharpening and/or retouch practices.
+They were also used to test the hypothesis that greater morphological variation would be apparent in the small class due to idiosyncratic responses related to local retouch practices.
 Results demonstrate that elliptical biface shape does not differ by raw material, but size does.
@@
 The subsequent analysis of biface morphology by size class demonstrated that size does not differ by raw material in the large class, but does in the small class.
-This finding supports the argument that elliptical biface morphology diverges through local resharpening and/or retouch practices, where bifaces manufactured using both raw materials enter their uselife at similar shapes and sizes prior to diverging in size.
-As expected, greater shape diversity occurs in the small class, where the Jowell Farm bifaces were found to be more morphologically diverse than those from the Millsap Cache.
-Distinct reduction practices are advanced as the driver of extant morphological differences found in elliptical bifaces from the ancestral Caddo area of the American Southeast.
+This finding supports the argument that elliptical biface morphology diverges through local retouch practices.
+As expected, greater shape diversity occurs in the small class, where Jowell Farm bifaces were found to be more morphologically diverse than those from Millsap Cache.
+Distinct local reduction practices are advanced as the driver of extant morphological differences found in elliptical bifaces from the ancestral Caddo area of the American Southeast.
 
-```{r figmap, fig.cap="Geological map illustrating the location of the Millsap Cache and Jowell Farm, the level three ecoregions of the Arkansas Valley (blue), Ouachita Mountains (yellow), South Central Plains (orange), and ancestral Caddo area boundary (white) in the American Southeast."}
+### Keywords {-}
+
+American Southeast, Caddo, Jowell biface, archaeology, museum studies, digital humanities, geometric morphometrics
+
+```{r figmap, fig.cap="Geological map illustrating the location of the Millsap Cache and Jowell Farm, and the level three ecoregions of the Arkansas Valley (blue), Ouachita Mountains (yellow), and South Central Plains (orange) in the ancestral Caddo area (white) of the American Southeast."}
 
@@
 Prior to landmarking, elliptical bifaces were oriented with the most heavily retouched edge at top right (Figure \@ref(fig:figlm)).
-The landmarking protocol uses three landmarks; two horizontal tangents (top/bottom), and one placed at the furthest extent of the edge including the heaviest amount of retouch.
+The landmarking protocol uses three landmarks; two horizontal tangents (top/bottom), and the third placed at the furthest extent of the edge that includes the heaviest amount of retouch.
 Landmarks and semilandmarks were applied using the `StereoMorph` package [@RN8973].
@@
 
-```{r figlm, fig.cap="Selected elliptical bifaces from the Jowell Farm site highlighting the most heavily retouched area identified using a modified version of the approach developed and advanced by Miller [-@RN11459], Smallwood [-@RN8310], and Miller and Smallwood [-@RN11460]."}
+```{r figlm, fig.cap="Selected elliptical bifaces from the Jowell Farm site highlighting the most heavily retouched area identified using a modified version of the flaking index developed and advanced by Miller [-@RN11459], Smallwood [-@RN8310], and Miller and Smallwood [-@RN11460]."}
 
@@
 To identify which edge was most heavily retouched, we employed a modified approach to the flaking index initially developed by Miller [-@RN11459], and later advanced and refined by Smallwood [-@RN8310], then Miller and Smallwood [-@RN11460].
-The modified approach uses counts of flake scars from each edge in the two most heavily worked areas, paired with a measure of edge length (inclusive of curvature - not a linear metric).
-The number of flake scars was then divided by the length of the worked edge, with the heaviest worked area identified by the greater value.
+The modified approach uses counts of flake scars from each edge in the two most heavily worked areas, paired with a measure of edge length (inclusive of curvature - this is not a linear metric).
+The number of flake scars was then divided by the length of the worked edge, with the heaviest worked edge identified by the greater value.
 
-The pattern that emerged suggested that the longest reworked edge on the elliptical bifaces was always more heavily retouched.
-This finding was used to incorporate the images of elliptical bifaces from the Millsap Cache [@RN11461], where the longest reworked edges were oriented at top right prior to landmarking.
+The longest reworked edge on the elliptical bifaces was almost always more heavily retouched.
+This finding was used to incorporate the images of elliptical bifaces from Millsap Cache [@RN11461], where the longest reworked edge was oriented at top right prior to landmarking.
 
@@
 
-To assess whether shape and size differed by `raw material` and `size class`, Procrustes ANOVAs [@RN1749] were run that enlist effect-sizes (zscores) computed as standard deviates of the generated sampling distributions [@RN1756].
+To assess whether `shape` changes with `size` (allometry), and whether shape and size differed by `raw material` and `size class`, Procrustes ANOVAs [@RN1749] were run that enlist effect-sizes (zscores) computed as standard deviates of the generated sampling distributions [@RN1756].
 A residual randomization permutation procedure (RRPP; n = 10,000 permutations) was used for all Procrustes ANOVAs [@RN1655;@RN11775], which has higher statistical power and a greater ability to identify patterns in the data should they be present [@RN1719].
@@
   -   Bifaces _do not differ_ in shape, but **differ significantly** in size by raw material/biogeography.
+  -   Allometry (shape change with size) was not significant.
   -   Bifaces in the _large size class_ recovered from the Millsap Cache and Jowell Farm do not differ in shape or size.
@@
 The shape of elliptical bifaces does not differ significantly in either size class; however, size does.
-Results of the Procrustes ANOVA support the hypothesis for divergent morphology.
-The analysis of morphological disparity yielded support for the interpretation that greater morphological diversity is associated with resharpening and/or retouch, while providing additional support for morphological consistency in production/manufacture.
+Results of the Procrustes ANOVAs support the hypothesis for divergent morphology.
+The analysis of morphological disparity yielded support for the interpretation that greater morphological diversity is associated with local retouch, while providing additional support for morphological consistency related to production.
 
@@
 Our thanks to the Caddo Nation of Oklahoma, the Caddo Nation Tribal Council, Tribal Chairman, and Tribal Historic Preservation Office for permission and access to NAGPRA and previously repatriated collections.
-Our gratitude is also extended to Marybeth Tomka and Lauren Bussiere at the Texas Archeological Research Laboratory for their assistance with access to the bifaces and the associated records, and to Scott Hammerstedt and Debra K.
+Our gratitude is also extended to Marybeth Tomka and Lauren Bussiere at the Texas Archeological Research Laboratory for their assistance with access to the bifaces and the associated records, to Sergio Ayala and the Gault School of Archaeological Research for access to the UV light, and to Scott Hammerstedt and Debra K.
 Green at the Oklahoma Archeological Survey for their assistance with records requests.
@@
 Hoggard, and David K.
-Thulman for their comments and constructive criticisms on the ongoing analyses of Caddo biface morphology, and to Emma Sherratt, Kersten Bergstrom, Dean C.
+Thulman for their comments and constructive criticisms on the ongoing analyses of Caddo biface morphology, as well as Emma Sherratt, Kersten Bergstrom, Dean C.
 Adams, and Michael L.

### 2022-08-07 supp/index.Rmd
<edit index>

@@
 Prior to landmarking, elliptical bifaces were oriented with the most heavily retouched edge at top right (Figure \@ref(fig:figlm)).
-The landmarking protocol uses three landmarks; two horizontal tangents (top/bottom), and the third placed at the furthest extent of the edge that includes the heaviest amount of retouch.
+The landmarking protocol uses three landmarks; two horizontal tangents (top/bottom), and the third placed at the furthest extent of the edge with the heaviest amount of retouch.
 Landmarks and semilandmarks were applied using the `StereoMorph` package [@RN8973].
-Some bifaces include multiple areas of heavy retouch at the top and bottom of the same lateral edge (Figure \@ref(fig:figlm): b, c, d, and e), while others (Figure \@ref(fig:figlm): f) include one retouched edge at top right and another at bottom left.
+Some bifaces include multiple areas of heavy retouch at the top and bottom of the same lateral edge (Figure \@ref(fig:figlm): b, c, d, and e), while others (Figure \@ref(fig:figlm): f) include one retouched edge at top right and another at bottom left, similar to alternatively beveled bifaces.
 
@@
   -   Allometry (shape change with size) was not significant.
-  -   Bifaces in the _large size class_ recovered from the Millsap Cache and Jowell Farm do not differ in shape or size.
-This suggests consistent morphology in bifaces from the _large size class_, in support of production-based standardization.
-  -   Bifaces in the _small size class_ recovered from the Millsap Cache and Jowell Farm **differ significantly** in size.
-Thus, while shape remains consistent, size differs, suggesting---at minimum---locally divergent approaches to retouch and/or resharpening.
-  -   Morphological disparity in the shape of the _small size class_ is **significantly greater** at Jowell Farm, providing evidence for a more standardized approach to resharpening and/or retouch at Millsap Cache.
-This finding also provides support for the hypothesis that greater shape diversity is associated with retouch and/or resharpening for elliptical bifaces, rather than production/manufacture.
+  -   Bifaces in the _large size class_ recovered from Millsap Cache and Jowell Farm do not differ in shape or size, suggesting consistent morphology in bifaces from the _large size class_, supporting arguments for production-based standardization.
+  -   Bifaces in the _small size class_ recovered from Millsap Cache and Jowell Farm **differ significantly** in size.
+Thus, while shape remains consistent, size differs, suggesting---at minimum---locally divergent approaches to retouch.
+  -   Morphological disparity in the shape of the _small size class_ is **significantly greater** at Jowell Farm, providing evidence for a more standardized approach to local retouch at Millsap Cache.
+This finding also provides support for the hypothesis that greater shape diversity is associated with retouch efforts, rather than production.
 

### 2022-08-07 supp/index.Rmd
<edit index>

@@
 They were also used to test the hypothesis that greater morphological variation would be apparent in the small class due to idiosyncratic responses related to local retouch practices.
-Results demonstrate that elliptical biface shape does not differ by raw material, but size does.
-This finding suggests a shared mental template, coupled with a size differential.
+Results demonstrate that elliptical biface shape does not differ by raw material, but size does, suggesting a shared mental template is maintained regardless of biface size.
 The subsequent analysis of biface morphology by size class demonstrated that size does not differ by raw material in the large class, but does in the small class.
@@
   -   Allometry (shape change with size) was not significant.
-  -   Bifaces in the _large size class_ recovered from Millsap Cache and Jowell Farm do not differ in shape or size, suggesting consistent morphology in bifaces from the _large size class_, supporting arguments for production-based standardization.
-  -   Bifaces in the _small size class_ recovered from Millsap Cache and Jowell Farm **differ significantly** in size.
+  -   Bifaces in the _large class_ recovered from Millsap Cache and Jowell Farm do not differ in shape or size, suggesting consistent morphology in bifaces from the _large class_, supporting arguments for production-based standardization.
+  -   Bifaces in the _small class_ recovered from Millsap Cache and Jowell Farm **differ significantly** in size.
 Thus, while shape remains consistent, size differs, suggesting---at minimum---locally divergent approaches to retouch.
-  -   Morphological disparity in the shape of the _small size class_ is **significantly greater** at Jowell Farm, providing evidence for a more standardized approach to local retouch at Millsap Cache.
+  -   Morphological disparity in the shape of the _small class_ is **significantly greater** at Jowell Farm, providing evidence for a more standardized approach to local retouch at Millsap Cache.
 This finding also provides support for the hypothesis that greater shape diversity is associated with retouch efforts, rather than production.
@@
 Components of this analytical work flow were developed and funded by a Preservation Technology and Training grant (P14AP00138) to RZS from the National Center for Preservation Technology and Training (NCPTT), and additional grants to RZS from the Caddo Nation of Oklahoma, National Forests and Grasslands in Texas (15-PA-11081300-033) and the United States Forest Service (20-PA-11081300-074).
-Funding to analyse the bifaces from the Millsap Cache and Jowell Farm was provided by the Heritage Research Center at Stephen F.
+Funding to analyse the bifaces from Millsap Cache and Jowell Farm was provided by the Heritage Research Center at Stephen F.
 Austin State University.

### 2022-08-07 supp/index.Rmd
<edit index>

@@
 Bifaces from Millsap Cache were reportedly manufactured from Kay County flint, providing for an empirical test of whether biface morphology differs as a function of raw material.
-The sample of elliptical bifaces from each site was also divided into two size classes; one reflective of initial production (large), and the other associated with retouch (small).
+The sample of elliptical bifaces from each site was also divided into two size classes; one reflective of production (large), and the other associated with retouch (small).
 Size classes were used to assess whether modifications by Caddo knappers may have yielded similar---convergent---biface shape in the small class.
 They were also used to test the hypothesis that greater morphological variation would be apparent in the small class due to idiosyncratic responses related to local retouch practices.
-Results demonstrate that elliptical biface shape does not differ by raw material, but size does, suggesting a shared mental template is maintained regardless of biface size.
-The subsequent analysis of biface morphology by size class demonstrated that size does not differ by raw material in the large class, but does in the small class.
+Results demonstrate that elliptical biface shape does not differ by raw material, but size does.
+This suggests a shared, and consistent, mental template that was maintained independently, regardless of biface size.
+The subsequent analysis of biface morphology by size class demonstrated that size does not differ by raw material in the large class, but in the small class, it does.
 This finding supports the argument that elliptical biface morphology diverges through local retouch practices.
@@
 
-American Southeast, Caddo, Jowell biface, archaeology, museum studies, digital humanities, geometric morphometrics
+American Southeast, Caddo, Jowell biface, archaeology, museum studies, digital humanities, geometric morphometrics, STEM, STEAM
 

### 2022-08-07 supp/index.Rmd
<edit index>

@@
 Results demonstrate that elliptical biface shape does not differ by raw material, but size does.
-This suggests a shared, and consistent, mental template that was maintained independently, regardless of biface size.
+This suggests a shared, and consistent, mental template that was consistently maintained independent of biface size.
 The subsequent analysis of biface morphology by size class demonstrated that size does not differ by raw material in the large class, but in the small class, it does.
@@
 
-  -   Bifaces _do not differ_ in shape, but **differ significantly** in size by raw material/biogeography.
+  -   Bifaces _do not differ_ in shape, but **differ significantly** in size by raw material.
   -   Allometry (shape change with size) was not significant.

### 2022-08-08 supp/index.Rmd
<edit index>

@@
 Results demonstrate that elliptical biface shape does not differ by raw material, but size does.
-This suggests a shared, and consistent, mental template that was consistently maintained independent of biface size.
-The subsequent analysis of biface morphology by size class demonstrated that size does not differ by raw material in the large class, but in the small class, it does.
+This suggests a shared, and consistent, mental template that was maintained independent of biface size.
+The subsequent analysis of elliptical biface morphology by size class demonstrated that size does not differ by raw material in the large class, but in the small class, it does.
 This finding supports the argument that elliptical biface morphology diverges through local retouch practices.
@@
 
-American Southeast, Caddo, Jowell biface, archaeology, museum studies, digital humanities, geometric morphometrics, STEM, STEAM
+American Southeast, Caddo, Jowell biface, archaeology, archaeoinformatics, museum studies, digital humanities, geometric morphometrics, STEM, STEAM
 
@@
 Landmarks and semilandmarks were applied using the `StereoMorph` package [@RN8973].
-Some bifaces include multiple areas of heavy retouch at the top and bottom of the same lateral edge (Figure \@ref(fig:figlm): b, c, d, and e), while others (Figure \@ref(fig:figlm): f) include one retouched edge at top right and another at bottom left, similar to alternatively beveled bifaces.
+Some bifaces include multiple areas of heavy retouch at the top and bottom of the same lateral edge (Figure \@ref(fig:figlm): b, c, d, and e), while others (Figure \@ref(fig:figlm): f) include one retouched edge at top right and another at bottom left, similar to alternately beveled bifaces.
 

### 2022-08-08 supp/index.Rmd
<edit index>

@@
 Elliptical bifaces from Millsap Cache and Jowell Farm (41AN13) comprise two of the largest samples of this ill-understood lithic artefact from the ancestral Caddo area.
-Elliptical bifaces from Jowell Farm were manufactured using Edwards chert, and are thought to have been acquired by the Caddo through trade or exchange with central Texas groups.
-Bifaces from Millsap Cache were reportedly manufactured from Kay County flint, providing for an empirical test of whether biface morphology differs as a function of raw material.
+Elliptical bifaces from Jowell Farm (n = 11) were manufactured using Edwards chert, and are thought to have been acquired by the Caddo through trade or exchange with central Texas groups.
+Bifaces from Millsap Cache (n = 19) were reportedly manufactured from Kay County flint, providing for an empirical test of whether biface morphology differs as a function of raw material.
 The sample of elliptical bifaces from each site was also divided into two size classes; one reflective of production (large), and the other associated with retouch (small).
@@
 The subsequent analysis of elliptical biface morphology by size class demonstrated that size does not differ by raw material in the large class, but in the small class, it does.
-This finding supports the argument that elliptical biface morphology diverges through local retouch practices.
+This finding supports the argument that elliptical biface morphology diverges through local reduction practices.
 As expected, greater shape diversity occurs in the small class, where Jowell Farm bifaces were found to be more morphologically diverse than those from Millsap Cache.

### 2022-08-08 supp/index.Rmd
<edit index>

@@
 Bifaces from Millsap Cache (n = 19) were reportedly manufactured from Kay County flint, providing for an empirical test of whether biface morphology differs as a function of raw material.
-The sample of elliptical bifaces from each site was also divided into two size classes; one reflective of production (large), and the other associated with retouch (small).
+The sample of elliptical bifaces from each site was also divided into two size classes; one conceptually reflective of production (large), and the other with local reduction practices (small).
 Size classes were used to assess whether modifications by Caddo knappers may have yielded similar---convergent---biface shape in the small class.
-They were also used to test the hypothesis that greater morphological variation would be apparent in the small class due to idiosyncratic responses related to local retouch practices.
+Size classes were also used to test the hypothesis that greater morphological variation would be apparent in the small class due to idiosyncratic responses related to local retouch practices.
 Results demonstrate that elliptical biface shape does not differ by raw material, but size does.
@@
 Prior to landmarking, elliptical bifaces were oriented with the most heavily retouched edge at top right (Figure \@ref(fig:figlm)).
-The landmarking protocol uses three landmarks; two horizontal tangents (top/bottom), and the third placed at the furthest extent of the edge with the heaviest amount of retouch.
-Landmarks and semilandmarks were applied using the `StereoMorph` package [@RN8973].
+The landmarking protocol uses three landmarks; two horizontal tangents (top/bottom), and the third placed at the furthest extent of the retouched edge bearing the heaviest amount of retouch.
+Landmarks and semilandmarks were applied in R 4.2.1 [@R] using the `StereoMorph` library v.
+1.6.7 [@RN8973].
 Some bifaces include multiple areas of heavy retouch at the top and bottom of the same lateral edge (Figure \@ref(fig:figlm): b, c, d, and e), while others (Figure \@ref(fig:figlm): f) include one retouched edge at top right and another at bottom left, similar to alternately beveled bifaces.

### 2022-08-08 supp/index.Rmd
<edit index>

@@
 
-Elliptical bifaces from Millsap Cache and Jowell Farm (41AN13) comprise two of the largest samples of this ill-understood lithic artefact from the ancestral Caddo area.
+Elliptical bifaces from Millsap Cache and Jowell Farm (41AN13) comprise two of the largest samples of this ill-understood stone tool from the ancestral Caddo area.
 Elliptical bifaces from Jowell Farm (n = 11) were manufactured using Edwards chert, and are thought to have been acquired by the Caddo through trade or exchange with central Texas groups.

### 2022-08-08 supp/index.Rmd
<edit index>

@@
 Elliptical bifaces from Millsap Cache and Jowell Farm (41AN13) comprise two of the largest samples of this ill-understood stone tool from the ancestral Caddo area.
-Elliptical bifaces from Jowell Farm (n = 11) were manufactured using Edwards chert, and are thought to have been acquired by the Caddo through trade or exchange with central Texas groups.
-Bifaces from Millsap Cache (n = 19) were reportedly manufactured from Kay County flint, providing for an empirical test of whether biface morphology differs as a function of raw material.
+Elliptical bifaces from Millsap Cache (n = 19) were reportedly produced using Kay County flint, while those from Jowell Farm (n = 11) were manufactured using Edwards chert, providing for an empirical test of whether biface morphology differs as a function of raw material.
 The sample of elliptical bifaces from each site was also divided into two size classes; one conceptually reflective of production (large), and the other with local reduction practices (small).
@@
 Results demonstrate that elliptical biface shape does not differ by raw material, but size does.
-This suggests a shared, and consistent, mental template that was maintained independent of biface size.
+This suggests that a shared, and consistent, mental template was maintained independent of biface size.
 The subsequent analysis of elliptical biface morphology by size class demonstrated that size does not differ by raw material in the large class, but in the small class, it does.
@@
 
-American Southeast, Caddo, Jowell biface, archaeology, archaeoinformatics, museum studies, digital humanities, geometric morphometrics, STEM, STEAM
+American Southeast, Caddo, Jowell biface, lithics, archaeology, archaeoinformatics, museum studies, digital humanities, geometric morphometrics, STEM, STEAM
 

### 2022-08-08 supp/index.Rmd
<edit index>

@@
 
-American Southeast, Caddo, Jowell biface, lithics, archaeology, archaeoinformatics, museum studies, digital humanities, geometric morphometrics, STEM, STEAM
+American Southeast, Caddo, archaeology, archaeoinformatics, museum studies, digital humanities, geometric morphometrics, STEM, STEAM, Jowell biface, lithics, parallel-oblique flake scars
 

### 2022-08-23 supp/index.Rmd
<edit index>

--- 2022-08-08 supp/index.Rmd
@@
 
-American Southeast, Caddo, Jowell biface, lithics, archaeology, archaeoinformatics, museum studies, digital humanities, geometric morphometrics, STEM, STEAM
+American Southeast, Caddo, archaeology, anthroinformatics, archaeoinformatics, museum studies, digital humanities, geometric morphometrics, STEM, STEAM, Jowell biface, lithics, parallel-oblique flaking
 
@@
 
-knitr::include_graphics('./images/ellipticalbifaces.jpg')
+knitr::include_graphics('./images/ellipticalbifaces.png')
 ```

### 2022-08-23 supp/index.Rmd
<edit index>

@@
 
-American Southeast, Caddo, archaeology, anthroinformatics, archaeoinformatics, museum studies, digital humanities, geometric morphometrics, STEM, STEAM, Jowell biface, lithics, parallel-oblique flaking
+American Southeast, Caddo, archaeology, anthroinformatics, archaeoinformatics, museum studies, digital humanities, geometric morphometrics, STEM, STEAM, Jowell knife, Jowell knives, lithics, parallel-oblique flaking
 

### 2022-08-29 supp/index.Rmd
<edit index>

@@
 
-Elliptical bifaces from Millsap Cache and Jowell Farm (41AN13) comprise two of the largest samples of this ill-understood stone tool from the ancestral Caddo area.
-Elliptical bifaces from Millsap Cache (n = 19) were reportedly produced using Kay County flint, while those from Jowell Farm (n = 11) were manufactured using Edwards chert, providing for an empirical test of whether biface morphology differs as a function of raw material.
-The sample of elliptical bifaces from each site was also divided into two size classes; one conceptually reflective of production (large), and the other with local reduction practices (small).
+Elliptical bifaces from Millsap Cache and Jowell Farm (41AN13) comprise two of the largest samples of this ill-understood stone tool from the ancestral Caddo area of the American Southeast.
+Elliptical bifaces from Millsap Cache were reportedly produced using Kay County flint, while those from Jowell Farm were manufactured using Edwards chert, providing for an empirical test of whether biface morphology differs as a function of raw material.
+The sample of elliptical bifaces was divided into two size classes; one conceptually reflective of production (large), and the other with local reduction practices (small).
 Size classes were used to assess whether modifications by Caddo knappers may have yielded similar---convergent---biface shape in the small class.
@@
 As expected, greater shape diversity occurs in the small class, where Jowell Farm bifaces were found to be more morphologically diverse than those from Millsap Cache.
-Distinct local reduction practices are advanced as the driver of extant morphological differences found in elliptical bifaces from the ancestral Caddo area of the American Southeast.
+Distinct local reduction practices are advanced as the driver of extant morphological differences found in elliptical bifaces.
 
@@
 
+```{r figillust, fig.cap="Illustrated elliptical bifaces from the ancestral Caddo area highlighting parallel-oblique flaking used in production."}
+
+knitr::include_graphics('./images/elliptical.illustration.png')
+```
+
   -   Bifaces _do not differ_ in shape, but **differ significantly** in size by raw material.

### 2022-09-01 supp/index.Rmd
<edit index>

@@
 
-American Southeast, Caddo, archaeology, anthroinformatics, archaeoinformatics, museum studies, digital humanities, geometric morphometrics, STEM, STEAM, Jowell knife, Jowell knives, lithics, parallel-oblique flaking
+American Southeast, Caddo, archaeology, anthroinformatics, archaeoinformatics, museum studies, digital humanities, geometric morphometrics, STEM, STEAM, Jowell knife, Jowell knives, lithics, oblique parallel flaking
 
@@
 
-```{r figillust, fig.cap="Illustrated elliptical bifaces from the ancestral Caddo area highlighting parallel-oblique flaking used in production."}
-
-knitr::include_graphics('./images/elliptical.illustration.png')
-```
-
   -   Bifaces _do not differ_ in shape, but **differ significantly** in size by raw material.
@@
 This finding also provides support for the hypothesis that greater shape diversity is associated with retouch efforts, rather than production.
+
+```{r figillust, fig.cap="Illustrated elliptical bifaces from the ancestral Caddo area highlighting oblique parallel flaking used in production."}
+
+knitr::include_graphics('./images/elliptical.illustration.png')
+```
 

### 2022-09-10 supp/index.Rmd
<edit index>

@@
 
-Elliptical bifaces from Millsap Cache and Jowell Farm (41AN13) comprise two of the largest samples of this ill-understood stone tool from the ancestral Caddo area of the American Southeast.
+Elliptical bifaces are prevalent in Mississipian sites throughout the American Southeast.
+Those from Millsap Cache and Jowell Farm (41AN13) comprise two of the largest samples of this ill-understood stone tool from the ancestral Caddo area of the American Southeast.
 Elliptical bifaces from Millsap Cache were reportedly produced using Kay County flint, while those from Jowell Farm were manufactured using Edwards chert, providing for an empirical test of whether biface morphology differs as a function of raw material.

### 2022-09-10 supp/index.Rmd
<edit index>

--- 2022-09-01 supp/index.Rmd
@@
 
-Elliptical bifaces from Millsap Cache and Jowell Farm (41AN13) comprise two of the largest samples of this ill-understood stone tool from the ancestral Caddo area of the American Southeast.
+Elliptical bifaces are prevalent at Mississipian sites throughout the American Southeast.
+Those from Millsap Cache and Jowell Farm (41AN13) comprise two of the largest samples of this ill-understood stone tool from the ancestral Caddo area.
 Elliptical bifaces from Millsap Cache were reportedly produced using Kay County flint, while those from Jowell Farm were manufactured using Edwards chert, providing for an empirical test of whether biface morphology differs as a function of raw material.

### 2022-09-15 supp/index.Rmd
<edit index>

@@
 Austin State University; Department of Biology, Stephen F.
-Austin State University; and Cultural Heritage Department, Jean Monnet University, zselden@sfasu.edu] and John E.
+Austin State University; Texas Archeological Research Laboratory, The University of Texas at Austin; and Cultural Heritage Department, Jean Monnet University, zselden@sfasu.edu] and John E.
 Dockall^[Stantec, Inc.]
@@
 
->Reproducibility---the ability to recompute results---and replicability---the chances other experimenters will achieve a consistent result---are two foundational characteristics of successful scientific research [@RN20918].
+>This process of comparison, of recognising in one form a definite permutation or _deformation_ of another, apart altogether from a precise and adequate understanding of the original "type" or standard of comparison, lies within the immediate province of mathematics, and finds its solution in the elementary use of a certain method of the mathematician [@RN7522].
 
-Elliptical bifaces are prevalent at Mississipian sites throughout the American Southeast.
+Elliptical bifaces are found at some regularity in Mississipian sites throughout the American Southeast.
 Those from Millsap Cache and Jowell Farm (41AN13) comprise two of the largest samples of this ill-understood stone tool from the ancestral Caddo area.
@@
 The sample of elliptical bifaces was divided into two size classes; one conceptually reflective of production (large), and the other with local reduction practices (small).
-Size classes were used to assess whether modifications by Caddo knappers may have yielded similar---convergent---biface shape in the small class.
+Size classes were used to assess whether modifications by Caddo knappers may have yielded similar---convergent---biface shapes in the small class.
 Size classes were also used to test the hypothesis that greater morphological variation would be apparent in the small class due to idiosyncratic responses related to local retouch practices.
@@
 This finding supports the argument that elliptical biface morphology diverges through local reduction practices.
-As expected, greater shape diversity occurs in the small class, where Jowell Farm bifaces were found to be more morphologically diverse than those from Millsap Cache.
-Distinct local reduction practices are advanced as the driver of extant morphological differences found in elliptical bifaces.
+As expected, greater shape diversity occurs in the small class, where small bifaces from Jowell Farm were found to be more morphologically diverse than small bifaces from Millsap Cache.
+Distinct local reduction practices are advanced as the driver of extant morphological differences that occur in elliptical bifaces.
 
@@
 
-```{r figillust, fig.cap="Illustrated elliptical bifaces from the ancestral Caddo area highlighting oblique parallel flaking used in production."}
+```{r figillust, fig.cap="Illustrated elliptical bifaces from a, an unknown location; b, Jowell Farm; and c, Belcher Mound highlighting oblique parallel flaking used in production.
+The specimen from Belcher Mound is a potential preform exhibiting oblique parallel removals consistent with elliptical bifaces from other sites in the region."}
 
@@
 Our thanks to the Caddo Nation of Oklahoma, the Caddo Nation Tribal Council, Tribal Chairman, and Tribal Historic Preservation Office for permission and access to NAGPRA and previously repatriated collections.
-Our gratitude is also extended to Marybeth Tomka and Lauren Bussiere at the Texas Archeological Research Laboratory for their assistance with access to the bifaces and the associated records, to Sergio Ayala and the Gault School of Archaeological Research for access to the UV light, and to Scott Hammerstedt and Debra K.
+Our gratitude is also extended to Marybeth Tomka and Lauren Bussiere at the Texas Archeological Research Laboratory for their assistance with access to the bifaces and associated records, to Sergio Ayala and the Gault School of Archaeological Research for access to the UV light, and to Scott Hammerstedt and Debra K.
 Green at the Oklahoma Archeological Survey for their assistance with records requests.
@@
 Hoggard, and David K.
-Thulman for their comments and constructive criticisms on the ongoing analyses of Caddo biface morphology, as well as Emma Sherratt, Kersten Bergstrom, Dean C.
+Thulman for their comments and constructive criticisms on the ongoing analyses of Caddo biface morphology, as well as Emma Sherratt, Kersten Bergstrom, Lauren Butaric, Dean C.
 Adams, and Michael L.
@@
 Components of this analytical work flow were developed and funded by a Preservation Technology and Training grant (P14AP00138) to RZS from the National Center for Preservation Technology and Training (NCPTT), and additional grants to RZS from the Caddo Nation of Oklahoma, National Forests and Grasslands in Texas (15-PA-11081300-033) and the United States Forest Service (20-PA-11081300-074).
-Funding to analyse the bifaces from Millsap Cache and Jowell Farm was provided by the Heritage Research Center at Stephen F.
+Funding to analyse the bifaces from Millsap Cache and Jowell Farm was made possible by a Research Support Fund Grant from the Texas Archeological Society.
+Additional support for this project was provided by the Heritage Research Center at Stephen F.
 Austin State University.
@@
 
+Reproducibility---the ability to recompute results---and replicability---the chances other experimenters will achieve a consistent result---are two foundational characteristics of successful scientific research [@RN20918].
 The analysis code associated with this project can be accessed through this document, is available in the GitHub repository ([https://github.com/seldenlab/elliptical.bifaces](https://github.com/seldenlab/elliptical.bifaces)), and digitally curated on the Open Science Framework ([DOI 10.17605/OSF.IO/PH25W](https://osf.io/ph25w/)).

### 2022-09-15 supp/index.Rmd
<edit index>

@@
 
-Elliptical bifaces are found at some regularity in Mississipian sites throughout the American Southeast.
+Elliptical bifaces are regularly found at Mississipian sites throughout the American Southeast.
 Those from Millsap Cache and Jowell Farm (41AN13) comprise two of the largest samples of this ill-understood stone tool from the ancestral Caddo area.
@@
 
-```{r figmap, fig.cap="Geological map illustrating the location of the Millsap Cache and Jowell Farm, and the level three ecoregions of the Arkansas Valley (blue), Ouachita Mountains (yellow), and South Central Plains (orange) in the ancestral Caddo area (white) of the American Southeast."}
+```{r figmap, fig.cap="Location of Millsap Cache, Jowell Farm, and other sites mentioned in the manuscript.
+Elliptical bifaces have been found in every major river basin in the ancestral Caddo area (white)."}
 

### 2022-09-16 supp/index.Rmd
<edit index>

--- 2022-09-15 supp/index.Rmd
@@
 
-American Southeast, Caddo, archaeology, anthroinformatics, archaeoinformatics, museum studies, digital humanities, geometric morphometrics, STEM, STEAM, Jowell knife, Jowell knives, lithics, oblique parallel flaking
+American Southeast, Caddo, archaeology, anthroinformatics, archaeoinformatics, museum studies, digital humanities, geometric morphometrics, STEM, STEAM, ovoid biface, Jowell knife, Jowell knives, lithics, oblique parallel flaking
 

### 2022-09-16 supp/index.Rmd

--- 2022-09-15 supp/index.Rmd
@@
 Components of this analytical work flow were developed and funded by a Preservation Technology and Training grant (P14AP00138) to RZS from the National Center for Preservation Technology and Training (NCPTT), and additional grants to RZS from the Caddo Nation of Oklahoma, National Forests and Grasslands in Texas (15-PA-11081300-033) and the United States Forest Service (20-PA-11081300-074).
-Funding to analyse the bifaces from Millsap Cache and Jowell Farm was made possible by a Research Support Fund Grant from the Texas Archeological Society.
-Additional support for this project was provided by the Heritage Research Center at Stephen F.
+Funding to analyse the bifaces from Millsap Cache and Jowell Farm was made possible by support from the Heritage Research Center at Stephen F.
 Austin State University.
+A Research Support Fund Grant from the Texas Archeological Society provided the tool needed to illustrate the bifaces.
 

### 2022-10-05 supp/index.Rmd
<edit index>

--- 2022-09-16 supp/index.Rmd
@@
 Austin State University; Department of Biology, Stephen F.
-Austin State University; Texas Archeological Research Laboratory, The University of Texas at Austin; and Cultural Heritage Department, Jean Monnet University, zselden@sfasu.edu] and John E.
+Austin State University; and Cultural Heritage Department, Jean Monnet University, zselden@sfasu.edu] and John E.
 Dockall^[Stantec, Inc.]
@@
 
->This process of comparison, of recognising in one form a definite permutation or _deformation_ of another, apart altogether from a precise and adequate understanding of the original "type" or standard of comparison, lies within the immediate province of mathematics, and finds its solution in the elementary use of a certain method of the mathematician [@RN7522].
+>Reproducibility---the ability to recompute results---and replicability---the chances other experimenters will achieve a consistent result---are two foundational characteristics of successful scientific research [@RN20918].
 
-Elliptical bifaces are regularly found at Mississipian sites throughout the American Southeast.
+Elliptical bifaces are prevalent at Mississipian sites throughout the American Southeast.
 Those from Millsap Cache and Jowell Farm (41AN13) comprise two of the largest samples of this ill-understood stone tool from the ancestral Caddo area.
-Elliptical bifaces from Millsap Cache were reportedly produced using Kay County flint, while those from Jowell Farm were manufactured using Edwards chert, providing for an empirical test of whether biface morphology differs as a function of raw material.
+Elliptical bifaces from Millsap Cache were reportedly produced using Kay County flint, while those from Jowell Farm were manufactured using Edwards chert, providing for an empirical test of morphological differences as a function of raw material.
 The sample of elliptical bifaces was divided into two size classes; one conceptually reflective of production (large), and the other with local reduction practices (small).
-Size classes were used to assess whether modifications by Caddo knappers may have yielded similar---convergent---biface shapes in the small class.
+Size classes were used to assess whether modifications by Caddo knappers may have yielded similar---convergent---biface shape in the small class.
 Size classes were also used to test the hypothesis that greater morphological variation would be apparent in the small class due to idiosyncratic responses related to local retouch practices.
@@
 This finding supports the argument that elliptical biface morphology diverges through local reduction practices.
-As expected, greater shape diversity occurs in the small class, where small bifaces from Jowell Farm were found to be more morphologically diverse than small bifaces from Millsap Cache.
-Distinct local reduction practices are advanced as the driver of extant morphological differences that occur in elliptical bifaces.
+As expected, greater shape diversity occurs in the small class, where Jowell Farm bifaces were found to be more morphologically diverse than those from Millsap Cache.
+Distinct local reduction practices are advanced as the driver of extant morphological differences found in elliptical bifaces.
 
@@
 
-American Southeast, Caddo, archaeology, anthroinformatics, archaeoinformatics, museum studies, digital humanities, geometric morphometrics, STEM, STEAM, ovoid biface, Jowell knife, Jowell knives, lithics, oblique parallel flaking
+American Southeast, Caddo, archaeology, anthroinformatics, archaeoinformatics, museum studies, digital humanities, geometric morphometrics, STEM, STEAM, Jowell knife, Jowell knives, lithics, oblique parallel flaking
 
@@
 
-```{r figillust, fig.cap="Illustrated elliptical bifaces from a, an unknown location; b, Jowell Farm; and c, Belcher Mound highlighting oblique parallel flaking used in production.
-The specimen from Belcher Mound is a potential preform exhibiting oblique parallel removals consistent with elliptical bifaces from other sites in the region."}
+```{r figillust, fig.cap="Illustrated elliptical bifaces from the ancestral Caddo area highlighting oblique parallel flaking used in production."}
 
@@
 Our thanks to the Caddo Nation of Oklahoma, the Caddo Nation Tribal Council, Tribal Chairman, and Tribal Historic Preservation Office for permission and access to NAGPRA and previously repatriated collections.
-Our gratitude is also extended to Marybeth Tomka and Lauren Bussiere at the Texas Archeological Research Laboratory for their assistance with access to the bifaces and associated records, to Sergio Ayala and the Gault School of Archaeological Research for access to the UV light, and to Scott Hammerstedt and Debra K.
+Our gratitude is also extended to Marybeth Tomka and Lauren Bussiere at the Texas Archeological Research Laboratory for their assistance with access to the bifaces and the associated records, to Sergio Ayala and the Gault School of Archaeological Research for access to the UV light, and to Scott Hammerstedt and Debra K.
 Green at the Oklahoma Archeological Survey for their assistance with records requests.
@@
 Hoggard, and David K.
-Thulman for their comments and constructive criticisms on the ongoing analyses of Caddo biface morphology, as well as Emma Sherratt, Kersten Bergstrom, Lauren Butaric, Dean C.
+Thulman for their comments and constructive criticisms on the ongoing analyses of Caddo biface morphology, as well as Emma Sherratt, Kersten Bergstrom, Dean C.
 Adams, and Michael L.
@@
 Components of this analytical work flow were developed and funded by a Preservation Technology and Training grant (P14AP00138) to RZS from the National Center for Preservation Technology and Training (NCPTT), and additional grants to RZS from the Caddo Nation of Oklahoma, National Forests and Grasslands in Texas (15-PA-11081300-033) and the United States Forest Service (20-PA-11081300-074).
-Funding to analyse the bifaces from Millsap Cache and Jowell Farm was made possible by a Research Support Fund Grant from the Texas Archeological Society.
-Additional support for this project was provided by the Heritage Research Center at Stephen F.
+Funding to analyse the bifaces from Millsap Cache and Jowell Farm was provided by the Heritage Research Center at Stephen F.
 Austin State University.
@@
 
-Reproducibility---the ability to recompute results---and replicability---the chances other experimenters will achieve a consistent result---are two foundational characteristics of successful scientific research [@RN20918].
 The analysis code associated with this project can be accessed through this document, is available in the GitHub repository ([https://github.com/seldenlab/elliptical.bifaces](https://github.com/seldenlab/elliptical.bifaces)), and digitally curated on the Open Science Framework ([DOI 10.17605/OSF.IO/PH25W](https://osf.io/ph25w/)).

### 2022-10-05 supp/index.Rmd

--- 2022-09-16 supp/index.Rmd
@@
 
-Elliptical bifaces are regularly found at Mississipian sites throughout the American Southeast.
+Elliptical bifaces are prevalent at Mississipian sites throughout the American Southeast.
 Those from Millsap Cache and Jowell Farm (41AN13) comprise two of the largest samples of this ill-understood stone tool from the ancestral Caddo area.
-Elliptical bifaces from Millsap Cache were reportedly produced using Kay County flint, while those from Jowell Farm were manufactured using Edwards chert, providing for an empirical test of whether biface morphology differs as a function of raw material.
+Elliptical bifaces from Millsap Cache were reportedly produced using Kay County flint, while those from Jowell Farm were manufactured using Edwards chert, providing for an empirical test of morphological differences as a function of raw material.
 The sample of elliptical bifaces was divided into two size classes; one conceptually reflective of production (large), and the other with local reduction practices (small).
-Size classes were used to assess whether modifications by Caddo knappers may have yielded similar---convergent---biface shapes in the small class.
+Size classes were used to assess whether modifications by Caddo knappers may have yielded similar---convergent---biface shape in the small class.
 Size classes were also used to test the hypothesis that greater morphological variation would be apparent in the small class due to idiosyncratic responses related to local retouch practices.

### 2022-10-18 supp/index.Rmd
<edit index>

@@
 
-American Southeast, Caddo, archaeology, anthroinformatics, archaeoinformatics, museum studies, digital humanities, geometric morphometrics, STEM, STEAM, ovoid biface, Jowell knife, Jowell knives, lithics, oblique parallel flaking
+American Southeast, Caddo, NAGPRA, archaeoinformatics, museum studies, digital humanities, geometric morphometrics, STEM, STEAM, ovoid biface, Jowell knife, Jowell knives, lithics, oblique parallel flaking
 
@@
   -   Bifaces in the _small class_ recovered from Millsap Cache and Jowell Farm **differ significantly** in size.
-Thus, while shape remains consistent, size differs, suggesting---at minimum---locally divergent approaches to retouch.
+Thus, while shape remains consistent, size differs, suggesting---at minimum---locally divergent approaches to retouch or retouch intensity.
   -   Morphological disparity in the shape of the _small class_ is **significantly greater** at Jowell Farm, providing evidence for a more standardized approach to local retouch at Millsap Cache.
-This finding also provides support for the hypothesis that greater shape diversity is associated with retouch efforts, rather than production.
+This finding provides additional support for the hypothesis that greater shape diversity is associated with retouch efforts, rather than production.
 

### 2022-10-18 supp/index.Rmd
<edit index>

--- 2022-10-05 supp/index.Rmd
@@
 
-Elliptical bifaces are prevalent at Mississipian sites throughout the American Southeast.
-Those from Millsap Cache and Jowell Farm (41AN13) comprise two of the largest samples of this ill-understood stone tool from the ancestral Caddo area.
-Elliptical bifaces from Millsap Cache were reportedly produced using Kay County flint, while those from Jowell Farm were manufactured using Edwards chert, providing for an empirical test of morphological differences as a function of raw material.
+Elliptical bifaces are regularly found at Mississipian sites throughout the American Southeast.
+Those from Millsap Cache and Jowell Farm comprise two of the largest samples of this ill-understood stone tool from the ancestral Caddo area.
+Elliptical bifaces from Millsap Cache were reportedly produced using Kay County flint, while those from Jowell Farm were manufactured using Edwards chert, providing for an empirical test of whether biface morphology differs as a function of raw material.
 The sample of elliptical bifaces was divided into two size classes; one conceptually reflective of production (large), and the other with local reduction practices (small).
-Size classes were used to assess whether modifications by Caddo knappers may have yielded similar---convergent---biface shape in the small class.
+Size classes were used to assess whether modifications by Caddo knappers may have yielded similar---convergent---biface shapes in the small class.
 Size classes were also used to test the hypothesis that greater morphological variation would be apparent in the small class due to idiosyncratic responses related to local retouch practices.
@@
 
-American Southeast, Caddo, archaeology, anthroinformatics, archaeoinformatics, museum studies, digital humanities, geometric morphometrics, STEM, STEAM, ovoid biface, Jowell knife, Jowell knives, lithics, oblique parallel flaking
+American Southeast, Caddo, archaeology, anthroinformatics, archaeoinformatics, museum studies, digital humanities, geometric morphometrics, STEM, STEAM, ovoid biface, Jowell knife, Jowell knives, lithics, oblique parallel flaking, 41AN13
 
@@
 
+The same landmark/semilandmark configuration was used for both analyses, and the landmarking protocol employs three landmarks; two horizontal tangents (top/bottom), and the third placed at the furthest extent of the retouched edge bearing the heaviest amount of retouch (see Figure \@ref(fig:figlm)).
+Equidistant semilandmarks were placed between each landmark, and all landmarks and semilandmarks were applied using R [@R] and the `StereoMorph` package [@RN9091].
+
 ### Generalised Procrustes Analysis
 
-Landmark data were aligned to a global coordinate system [@RN11622;@RN11623;@RN11563], achieved through generalized Procrustes superimposition [@RN478] performed in R 4.2.1 [@R] using the `geomorph` library v.
-4.0.4 [@RN11530;@RN1774;@RN9565].
-Procrustes superimposition translates, scales, and rotates the coordinate data to allow for comparisons among objects [@RN11564;@RN478].
-The `geomorph` package uses a partial Procrustes superimposition that projects the aligned specimens into tangent space following alignment in preparation for the use of multivariate methods that assume linear space [@RN1646;@RN11563].
+Landmark data were aligned to a global coordinate system [@RN8477;@RN7502;@RN11622;@RN11623;@RN11563], achieved through generalised Procrustes superimposition [@RN11138;@RN478;@RN1646] in R using the `geomorph` and `RRPP` packages [@RN1655;@RN11775;@RN11530;@RN1774;@RN9565].
+Procrustes superimposition translates and rotates the coordinate data to allow for comparisons among objects, while also scaling each biface using unit-centroid size---the square root of the sum of squared distances from each landmark to the specimen’s centroid [@RN11139;@RN11140;@RN11564;@RN478].
+The `geomorph` package uses a partial Procrustes superimposition that projects the aligned specimens into tangent space subsequent to alignment in preparation for the use of multivariate methods that assume linear space [@RN11141;@RN11142;@RN1646;@RN11563].
 
@@
 
-Principal components analysis [@RN1746] was used to visualise shape variation among the elliptical bifaces.
-The shape changes described by each principal axis are commonly visualized using thin-plate spline warping of a reference image or 3D mesh [@RN1731;@RN479].
+Principal components analysis [@RN1746] was used to visualise shape variation among the elliptical bifaces, and the scatterplot represents the dispersion of shapes in tangent space [@RN8633;@RN5616;@RN11143;@RN7550].
+Those shape changes described by each principal axis are commonly visualized using thin-plate spline warping of a reference image [@RN1731;@RN479].
 

### 2022-10-31 ms/ms/ms.Rmd
<init ms>

---
title: |
  Size matters: Divergent morphology in elliptical bifaces from sites in the American Southeast articulates with distinct local reduction practices
type: ARTICLE TEMPLATE
author:
  - name: Robert Z. Selden, Jr.
    affil: a
    email: zselden@sfasu.edu
  - name: John E. Dockall
    affil: b
    email: john.dockall@stantec.com
  - name: David H. Dye
    affil: c
    email: daviddye@memphis.edu
affiliation:
  - num: a
    address: |
      Heritage Research Center, Stephen F. Austin State University; Department of Biology, Stephen F. Austin State University; Texas Archeological Research Laboratory, The University of Texas at Austin; and Cultural Heritage Department, Jean Monnet University 
  - num: b
    address: |
      Stantec, Inc.
  - num: c
    address: |
      Department of Earth Sciences, The University of Memphis
bibliography: interactcadsample.bib
appendix: appendix.tex
abstract: |
  Elliptical bifaces are prevalent at Mississippian sites throughout the American Southeast. Those from Millsap Cache and Jowell Farm (41AN13) comprise two of the largest samples of this ill-understood stone tool from the ancestral Caddo area. Elliptical bifaces from Millsap Cache were produced using Kay County flint, while those from Jowell Farm were manufactured using Edwards chert, providing for an empirical test of morphological differences as a function of raw material. The sample of elliptical bifaces was divided into two size classes; one conceptually reflective of production (large), and the other with local reduction practices (small). Size classes were used to assess whether modifications by Caddo knappers may have yielded similar—convergent—biface shape in the small class. Size classes were also used to test the hypothesis that greater morphological variation would be apparent in the small class due to idiosyncratic responses related to local retouch practices. Results demonstrate that elliptical biface shape does not differ by raw material, but size does. This suggests that a shared, and consistent, mental template was maintained independent of biface size. The subsequent analysis of elliptical biface morphology by size class demonstrated that size does not differ by raw material in the large class, but in the small class, it does. This finding supports the argument that elliptical biface morphology diverges through local reduction practices. As expected, greater shape diversity occurs in the small class, where Jowell Farm bifaces were found to be more morphologically diverse than those from Millsap Cache. Distinct local reduction practices are advanced as the driver of extant morphological differences found in elliptical bifaces.
keywords: |
  American Southeast; Caddo; NAGPRA; archaeology; ovoid biface; Jowell knife; Jowell knives; lithics; oblique parallel flaking; archaeoinformatics; museum studies; digital humanities; non-western art history; geometric morphometrics; STEM; STEAM
header-includes: |
  \usepackage{hyperref}
  \usepackage[utf8]{inputenc}
  \def\tightlist{}
output: rticles::tf_article
---

\begin{quote}
\textit{This process of comparison, of recognising in one form a definite permutation or deformation of another, apart altogether from a precise and adequate understanding of the original 'type' or standard of comparison, lies within the immediate province of mathematics, and finds its solution in the elementary use of a certain method of the mathematician} \citep{RN7522}.
\end{quote}

# Introduction

\begin{figure}\centering
\includegraphics[width=\linewidth]{figs/map.png}
\caption{Location of Millsap Cache, Jowell Farm, and other sites mentioned in the manuscript. Elliptical bifaces have been found in every major river basin in the ancestral Caddo area (white).}
\label{fig:map}
\end{figure}

\begin{figure}\centering
\includegraphics[width=\linewidth]{figs/elliptical.illustration.png}
\caption{Illustrated elliptical bifaces from a, an unknown location; b, Jowell Farm; and c, Belcher Mound highlighting oblique parallel flaking used in production. The specimen from Belcher Mound is a potential preform exhibiting oblique parallel removals consistent with elliptical bifaces from other sites in the region.}
\label{fig:illustrated}
\end{figure}

# Methods

To assess the variable impacts that Caddo retouch practices may have had on general morphology, the sample of elliptical bifaces was binned into two size classes using centroid size. Due to potential morphological differences associated with the use of two distinct raw materials (Kay County flint and Edwards chert), the sample was subset by site prior to calculating the mean centroid size for each. Bifaces larger than the mean were then assigned to the large category, those smaller than the mean were assigned to the small category, and the two datasets were then joined in advance of analysis ([Supplementary Materials, Ch. 1](https://seldenlab.github.io/elliptical.bifaces/define-size-classes.html)).

Elliptical bifaces from the Millsap Cache and Jowell Farm were used to examine whether biface morphology remains stable or expresses morphological variability by site/raw material ([Supplementary Materials, Ch. 2](https://seldenlab.github.io/elliptical.bifaces/gm---siteraw-material.html)). Prior to landmarking, elliptical bifaces were oriented with the most heavily retouched edge at top right (Figure  \ref{fig:elliptical}). Some bifaces include multiple areas of heavy retouch at the top and bottom of the same lateral edge (Figure \ref{fig:elliptical}:b, c, d, and e), while others (Figure \ref{fig:elliptical}:f) include one retouched edge at top right and another at bottom left.

\begin{figure}\centering
\includegraphics[width=\linewidth]{figs/ellipticalbifaces.png}
\caption{Selected elliptical bifaces from the Jowell Farm site highlighting the most heavily retouched area identified using a modified version of the flaking index developed and advanced by \citet{RN11099}, \citet{RN9242}, and \citet{RN11098}.}
\label{fig:elliptical}
\end{figure}

To identify which edge was most heavily retouched, we employed a modified approach to the flaking index initially developed by \citet{RN11099}, and later advanced and refined by \citet{RN9242}, then \citet{RN11098}. The modified approach uses counts of flake scars from each edge in the two most heavily worked areas, paired with a measure of edge length inclusive of curvature, which was calculated using ImageJ (Abramoff et al. 2004, Rasband 1997, Schneider et al. 2012). The number of flake scars was then divided by the length of the worked edge, with the most heavily worked edge identified by the greater value.

## Geometric morphometrics

The same landmark/semilandmark configuration was used for both analyses, and the landmarking protocol employs three landmarks; two horizontal tangents (top/bottom), and the third placed at the furthest extent of the retouched edge bearing the heaviest amount of retouch (see Figure \ref{fig:elliptical}). Equidistant semilandmarks were placed between each landmark, and all landmarks and semilandmarks were applied using R 4.2.1 (R Core Development Team, 2022) and the StereoMorph package (Olsen and Westneat 2015).

Landmark data were aligned to a global coordinate system (Bookstein et al. 1999, Gunz et al. 2005, Kendall 1981, 1984, Slice 2001), achieved through generalised Procrustes superimposition (Bookstein 1986, Rohlf and Slice 1990, Rohlf 1999) in R using the geomorph (Adams et al. 2017, Adams and Otarola-Castillo 2013, Baken et al. 2021) and RRPP packages (Adams and Collyer 2015, Collyer and Adams 2018). Procrustes superimposition translates and rotates the coordinate data to allow for comparisons among objects, while also scaling each biface using unit-centroid size—the square root of the sum of squared distances from each landmark to the specimen’s centroid (Chapman 1990, Dryden and Mardia 1998, Gower 1975, Rohlf and Slice 1990). The geomorph package uses a partial Procrustes superimposition that projects the aligned specimens into tangent space subsequent to alignment in preparation for the use of multivariate methods that assume linear space (Dryden and Mardia 1993, Kent and Mardia 2001, Rohlf 1999, Slice 2001).

# Acknowledgement(s) {-}

Our thanks to the Caddo Nation of Oklahoma, the Caddo Nation Tribal Council, Tribal Chairman, and Tribal Historic Preservation Office for permission and access to NAGPRA and previously repatriated collections. Our gratitude is also extended to Marybeth Tomka and Lauren Bussiere at the Texas Archeological Research Laboratory for their assistance with access to the bifaces and associated records, to Sergio Ayala and the Gault School of Archaeological Research for access to the UV light, and to Scott Hammerstedt and Debra K. Green at the Oklahoma Archeological Survey for their assistance with records requests. Thanks also to John Harman for access to the DStretch plugin for ImageJ that was useful in the analysis of flake scars, and to Harry J. Shafer, Hiram F. (Pete) Gregory, Christian S. Hoggard, and David K. Thulman for their comments and constructive criticisms on the ongoing analyses of Caddo biface morphology, as well as Emma Sherratt, Kersten Bergstrom, Lauren Butaric, Dean C. Adams, and Michael L. Collyer for their constructive criticisms, general comments, and suggestions throughout the development of this research program.

# Funding {-}

Components of this analytical work flow were developed and funded by a Preservation Technology and Training grant (P14AP00138) to RZS from the National Center for Preservation Technology and Training (NCPTT), and additional grants to RZS from the Caddo Nation of Oklahoma, National Forests and Grasslands in Texas (15-PA-11081300-033) and the United States Forest Service (20-PA-11081300-074). Funding to analyse the bifaces from Millsap Cache and Jowell Farm was made possible by support from the Heritage Research Center at Stephen F. Austin State University. A Research Support Fund Grant from the Texas Archeological Society provided the tool needed to illustrate the bifaces.

# Data management

Reproducibility—the ability to recompute results—and replicability—the chances other experimenters will achieve a consistent result—are two foundational characteristics of successful scientific research (Leek and Peng 2015). The analysis code associated with this project can be accessed through the [Supplementary Materials](https://seldenlab.github.io/elliptical.bifaces/), is available through the GitHub repository [https://github.com/seldenlab/elliptical.bifaces](https://github.com/seldenlab/elliptical.bifaces), and digitally curated on the Open Science Framework [DOI 10.17605/OSF.IO/PH25W](https://osf.io/ph25w/). The reproducible nature of this enterprise provides a means for others to critically assess and evaluate the various analytical components (Gray and Marwick 2019; Peng 2011; Gandrud 2014), which is a necessary requirement for the production of reliable knowledge.

Reproducibility projects in psychology and cancer biology are impacting current research practices across all domains. Examples of reproducible research are becoming more abundant in archaeology (Marwick 2016; Ivanovaite et al. 2020; Selden Jr., Dockall, and Shafer 2018; Selden Jr., Dockall, and Dubied 2020; Selden Jr et al. 2021a; Selden 2022), and the next generation of archaeologists are learning those tools and methods needed to reproduce and/or replicate research results (Marwick et al. 2019). Reproducible and replicable research work flows are often employed at the highest levels of humanities-based inquiries to mitigate concern or doubt regarding proper execution, and is of particular import should the results have—explicitly or implicitly—a major impact on scientific progress (Peels and Bouter 2018).

### 2022-10-31 ms/ms/ms.tex
<init ms>

--- 2022-10-31 ms/ms/ms.Rmd
@@
----
-title: |
-  Size matters: Divergent morphology in elliptical bifaces from sites in the American Southeast articulates with distinct local reduction practices
-type: ARTICLE TEMPLATE
-author:
-  - name: Robert Z. Selden, Jr.
-    affil: a
-    email: zselden@sfasu.edu
-  - name: John E. Dockall
-    affil: b
-    email: john.dockall@stantec.com
-  - name: David H. Dye
-    affil: c
-    email: daviddye@memphis.edu
-affiliation:
-  - num: a
-    address: |
-      Heritage Research Center, Stephen F.
-Austin State University; Department of Biology, Stephen F.
-Austin State University; Texas Archeological Research Laboratory, The University of Texas at Austin; and Cultural Heritage Department, Jean Monnet University
-  - num: b
-    address: |
-      Stantec, Inc.
-  - num: c
-    address: |
-      Department of Earth Sciences, The University of Memphis
-bibliography: interactcadsample.bib
-appendix: appendix.tex
-abstract: |
-  Elliptical bifaces are prevalent at Mississippian sites throughout the American Southeast.
-Those from Millsap Cache and Jowell Farm (41AN13) comprise two of the largest samples of this ill-understood stone tool from the ancestral Caddo area.
-Elliptical bifaces from Millsap Cache were produced using Kay County flint, while those from Jowell Farm were manufactured using Edwards chert, providing for an empirical test of morphological differences as a function of raw material.
-The sample of elliptical bifaces was divided into two size classes; one conceptually reflective of production (large), and the other with local reduction practices (small).
-Size classes were used to assess whether modifications by Caddo knappers may have yielded similar—convergent—biface shape in the small class.
-Size classes were also used to test the hypothesis that greater morphological variation would be apparent in the small class due to idiosyncratic responses related to local retouch practices.
-Results demonstrate that elliptical biface shape does not differ by raw material, but size does.
-This suggests that a shared, and consistent, mental template was maintained independent of biface size.
-The subsequent analysis of elliptical biface morphology by size class demonstrated that size does not differ by raw material in the large class, but in the small class, it does.
-This finding supports the argument that elliptical biface morphology diverges through local reduction practices.
-As expected, greater shape diversity occurs in the small class, where Jowell Farm bifaces were found to be more morphologically diverse than those from Millsap Cache.
-Distinct local reduction practices are advanced as the driver of extant morphological differences found in elliptical bifaces.
-keywords: |
-  American Southeast; Caddo; NAGPRA; archaeology; ovoid biface; Jowell knife; Jowell knives; lithics; oblique parallel flaking; archaeoinformatics; museum studies; digital humanities; non-western art history; geometric morphometrics; STEM; STEAM
-header-includes: |
+% interactcadsample.tex
+% v1.03 - April 2017
+
+\documentclass[]{interact}
+
+\usepackage{epstopdf}% To incorporate .eps illustrations using PDFLaTeX, etc.
+\usepackage{subfigure}% Support for small, `sub' figures and tables
+%\usepackage[nolists,tablesfirst]{endfloat}% To `separate' figures and tables from text if required
+
+\usepackage{natbib}% Citation support using natbib.sty
+\bibpunct[, ]{(}{)}{;}{a}{}{,}% Citation support using natbib.sty
+\renewcommand\bibfont{\fontsize{10}{12}\selectfont}% Bibliography support using natbib.sty
+
+\theoremstyle{plain}% Theorem-like structures provided by amsthm.sty
+\newtheorem{theorem}{Theorem}[section]
+\newtheorem{lemma}[theorem]{Lemma}
+\newtheorem{corollary}[theorem]{Corollary}
+\newtheorem{proposition}[theorem]{Proposition}
+
+\theoremstyle{definition}
+\newtheorem{definition}[theorem]{Definition}
+\newtheorem{example}[theorem]{Example}
+
+\theoremstyle{remark}
+\newtheorem{remark}{Remark}
+\newtheorem{notation}{Notation}
+
+% tightlist command for lists without linebreak
+\providecommand{\tightlist}{%
+  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
+
 \usepackage{hyperref}
@@
 \def\tightlist{}
-output: rticles::tf_article
----
+
+\begin{document}
+
+\articletype{ARTICLE TEMPLATE}
+
+\title{Size matters: Divergent morphology in elliptical bifaces from
+sites in the American Southeast articulates with distinct local
+reduction practices}
+
+\author{\name{Robert Z. Selden, Jr.$^{a}$, John E. Dockall$^{b}$, David
+H. Dye$^{c}$}
+\affil{$^{a}$Heritage Research Center, Stephen F. Austin State
+University; Department of Biology, Stephen F. Austin State University;
+Texas Archeological Research Laboratory, The University of Texas at
+Austin; and Cultural Heritage Department, Jean Monnet
+University; $^{b}$Stantec, Inc.; $^{c}$Department of Earth Sciences, The
+University of Memphis}
+}
+
+\thanks{CONTACT Robert Z. Selden,
+Jr.. Email: \href{mailto:zselden@sfasu.edu}{\nolinkurl{zselden@sfasu.edu}}, John
+E.
+Dockall. Email: \href{mailto:john.dockall@stantec.com}{\nolinkurl{john.dockall@stantec.com}}, David
+H.
+Dye. Email: \href{mailto:daviddye@memphis.edu}{\nolinkurl{daviddye@memphis.edu}}}
+
+\maketitle
+
+\begin{abstract}
+Elliptical bifaces are prevalent at Mississippian sites throughout the
+American Southeast. Those from Millsap Cache and Jowell Farm (41AN13)
+comprise two of the largest samples of this ill-understood stone tool
+from the ancestral Caddo area. Elliptical bifaces from Millsap Cache
+were produced using Kay County flint, while those from Jowell Farm were
+manufactured using Edwards chert, providing for an empirical test of
+morphological differences as a function of raw material. The sample of
+elliptical bifaces was divided into two size classes; one conceptually
+reflective of production (large), and the other with local reduction
+practices (small). Size classes were used to assess whether
+modifications by Caddo knappers may have yielded
+similar---convergent---biface shape in the small class. Size classes
+were also used to test the hypothesis that greater morphological
+variation would be apparent in the small class due to idiosyncratic
+responses related to local retouch practices. Results demonstrate that
+elliptical biface shape does not differ by raw material, but size does.
+This suggests that a shared, and consistent, mental template was
+maintained independent of biface size. The subsequent analysis of
+elliptical biface morphology by size class demonstrated that size does
+not differ by raw material in the large class, but in the small class,
+it does. This finding supports the argument that elliptical biface
+morphology diverges through local reduction practices. As expected,
+greater shape diversity occurs in the small class, where Jowell Farm
+bifaces were found to be more morphologically diverse than those from
+Millsap Cache. Distinct local reduction practices are advanced as the
+driver of extant morphological differences found in elliptical bifaces.
+\end{abstract}
+
+\begin{keywords}
+American Southeast; Caddo; NAGPRA; archaeology; ovoid biface; Jowell
+knife; Jowell knives; lithics; oblique parallel flaking;
+archaeoinformatics; museum studies; digital humanities; non-western art
+history; geometric morphometrics; STEM; STEAM
+\end{keywords}
 
@@
 
-# Introduction
+\hypertarget{introduction}{%
+\section{Introduction}\label{introduction}}
 
@@
 
-# Methods
+\hypertarget{methods}{%
+\section{Methods}\label{methods}}
 
-To assess the variable impacts that Caddo retouch practices may have had on general morphology, the sample of elliptical bifaces was binned into two size classes using centroid size.
-Due to potential morphological differences associated with the use of two distinct raw materials (Kay County flint and Edwards chert), the sample was subset by site prior to calculating the mean centroid size for each.
-Bifaces larger than the mean were then assigned to the large category, those smaller than the mean were assigned to the small category, and the two datasets were then joined in advance of analysis ([Supplementary Materials, Ch.
-1](https://seldenlab.github.io/elliptical.bifaces/define-size-classes.html)).
+To assess the variable impacts that Caddo retouch practices may have had
+on general morphology, the sample of elliptical bifaces was binned into
+two size classes using centroid size. Due to potential morphological
+differences associated with the use of two distinct raw materials (Kay
+County flint and Edwards chert), the sample was subset by site prior to
+calculating the mean centroid size for each. Bifaces larger than the
+mean were then assigned to the large category, those smaller than the
+mean were assigned to the small category, and the two datasets were then
+joined in advance of analysis
+(\href{https://seldenlab.github.io/elliptical.bifaces/define-size-classes.html}{Supplementary
+Materials, Ch. 1}).
 
-Elliptical bifaces from the Millsap Cache and Jowell Farm were used to examine whether biface morphology remains stable or expresses morphological variability by site/raw material ([Supplementary Materials, Ch.
-2](https://seldenlab.github.io/elliptical.bifaces/gm---siteraw-material.html)).
-Prior to landmarking, elliptical bifaces were oriented with the most heavily retouched edge at top right (Figure  \ref{fig:elliptical}).
-Some bifaces include multiple areas of heavy retouch at the top and bottom of the same lateral edge (Figure \ref{fig:elliptical}:b, c, d, and e), while others (Figure \ref{fig:elliptical}:f) include one retouched edge at top right and another at bottom left.
+Elliptical bifaces from the Millsap Cache and Jowell Farm were used to
+examine whether biface morphology remains stable or expresses
+morphological variability by site/raw material
+(\href{https://seldenlab.github.io/elliptical.bifaces/gm---siteraw-material.html}{Supplementary
+Materials, Ch. 2}). Prior to landmarking, elliptical bifaces were
+oriented with the most heavily retouched edge at top right (Figure
+\ref{fig:elliptical}). Some bifaces include multiple areas of heavy
+retouch at the top and bottom of the same lateral edge (Figure
+\ref{fig:elliptical}:b, c, d, and e), while others (Figure
+\ref{fig:elliptical}:f) include one retouched edge at top right and
+another at bottom left.
 
@@
 
-To identify which edge was most heavily retouched, we employed a modified approach to the flaking index initially developed by \citet{RN11099}, and later advanced and refined by \citet{RN9242}, then \citet{RN11098}.
-The modified approach uses counts of flake scars from each edge in the two most heavily worked areas, paired with a measure of edge length inclusive of curvature, which was calculated using ImageJ (Abramoff et al.
-2004, Rasband 1997, Schneider et al.
-2012).
-The number of flake scars was then divided by the length of the worked edge, with the most heavily worked edge identified by the greater value.
+To identify which edge was most heavily retouched, we employed a
+modified approach to the flaking index initially developed by
+\citet{RN11099}, and later advanced and refined by \citet{RN9242}, then
+\citet{RN11098}. The modified approach uses counts of flake scars from
+each edge in the two most heavily worked areas, paired with a measure of
+edge length inclusive of curvature, which was calculated using ImageJ
+(Abramoff et al.~2004, Rasband 1997, Schneider et al.~2012). The number
+of flake scars was then divided by the length of the worked edge, with
+the most heavily worked edge identified by the greater value.
 
-## Geometric morphometrics
+\hypertarget{geometric-morphometrics}{%
+\subsection{Geometric morphometrics}\label{geometric-morphometrics}}
 
-The same landmark/semilandmark configuration was used for both analyses, and the landmarking protocol employs three landmarks; two horizontal tangents (top/bottom), and the third placed at the furthest extent of the retouched edge bearing the heaviest amount of retouch (see Figure \ref{fig:elliptical}).
-Equidistant semilandmarks were placed between each landmark, and all landmarks and semilandmarks were applied using R 4.2.1 (R Core Development Team, 2022) and the StereoMorph package (Olsen and Westneat 2015).
+The same landmark/semilandmark configuration was used for both analyses,
+and the landmarking protocol employs three landmarks; two horizontal
+tangents (top/bottom), and the third placed at the furthest extent of
+the retouched edge bearing the heaviest amount of retouch (see Figure
+\ref{fig:elliptical}). Equidistant semilandmarks were placed between
+each landmark, and all landmarks and semilandmarks were applied using R
+4.2.1 (R Core Development Team, 2022) and the StereoMorph package (Olsen
+and Westneat 2015).
 
-Landmark data were aligned to a global coordinate system (Bookstein et al.
-1999, Gunz et al.
-2005, Kendall 1981, 1984, Slice 2001), achieved through generalised Procrustes superimposition (Bookstein 1986, Rohlf and Slice 1990, Rohlf 1999) in R using the geomorph (Adams et al.
-2017, Adams and Otarola-Castillo 2013, Baken et al.
-2021) and RRPP packages (Adams and Collyer 2015, Collyer and Adams 2018).
-Procrustes superimposition translates and rotates the coordinate data to allow for comparisons among objects, while also scaling each biface using unit-centroid size—the square root of the sum of squared distances from each landmark to the specimen’s centroid (Chapman 1990, Dryden and Mardia 1998, Gower 1975, Rohlf and Slice 1990).
-The geomorph package uses a partial Procrustes superimposition that projects the aligned specimens into tangent space subsequent to alignment in preparation for the use of multivariate methods that assume linear space (Dryden and Mardia 1993, Kent and Mardia 2001, Rohlf 1999, Slice 2001).
+Landmark data were aligned to a global coordinate system (Bookstein et
+al.~1999, Gunz et al.~2005, Kendall 1981, 1984, Slice 2001), achieved
+through generalised Procrustes superimposition (Bookstein 1986, Rohlf
+and Slice 1990, Rohlf 1999) in R using the geomorph (Adams et al.~2017,
+Adams and Otarola-Castillo 2013, Baken et al.~2021) and RRPP packages
+(Adams and Collyer 2015, Collyer and Adams 2018). Procrustes
+superimposition translates and rotates the coordinate data to allow for
+comparisons among objects, while also scaling each biface using
+unit-centroid size---the square root of the sum of squared distances
+from each landmark to the specimen's centroid (Chapman 1990, Dryden and
+Mardia 1998, Gower 1975, Rohlf and Slice 1990). The geomorph package
+uses a partial Procrustes superimposition that projects the aligned
+specimens into tangent space subsequent to alignment in preparation for
+the use of multivariate methods that assume linear space (Dryden and
+Mardia 1993, Kent and Mardia 2001, Rohlf 1999, Slice 2001).
 
-# Acknowledgement(s) {-}
+\hypertarget{acknowledgements}{%
+\section*{Acknowledgement(s)}\label{acknowledgements}}
+\addcontentsline{toc}{section}{Acknowledgement(s)}
 
-Our thanks to the Caddo Nation of Oklahoma, the Caddo Nation Tribal Council, Tribal Chairman, and Tribal Historic Preservation Office for permission and access to NAGPRA and previously repatriated collections.
-Our gratitude is also extended to Marybeth Tomka and Lauren Bussiere at the Texas Archeological Research Laboratory for their assistance with access to the bifaces and associated records, to Sergio Ayala and the Gault School of Archaeological Research for access to the UV light, and to Scott Hammerstedt and Debra K.
-Green at the Oklahoma Archeological Survey for their assistance with records requests.
-Thanks also to John Harman for access to the DStretch plugin for ImageJ that was useful in the analysis of flake scars, and to Harry J.
-Shafer, Hiram F.
-(Pete) Gregory, Christian S.
-Hoggard, and David K.
-Thulman for their comments and constructive criticisms on the ongoing analyses of Caddo biface morphology, as well as Emma Sherratt, Kersten Bergstrom, Lauren Butaric, Dean C.
-Adams, and Michael L.
-Collyer for their constructive criticisms, general comments, and suggestions throughout the development of this research program.
+Our thanks to the Caddo Nation of Oklahoma, the Caddo Nation Tribal
+Council, Tribal Chairman, and Tribal Historic Preservation Office for
+permission and access to NAGPRA and previously repatriated collections.
+Our gratitude is also extended to Marybeth Tomka and Lauren Bussiere at
+the Texas Archeological Research Laboratory for their assistance with
+access to the bifaces and associated records, to Sergio Ayala and the
+Gault School of Archaeological Research for access to the UV light, and
+to Scott Hammerstedt and Debra K. Green at the Oklahoma Archeological
+Survey for their assistance with records requests. Thanks also to John
+Harman for access to the DStretch plugin for ImageJ that was useful in
+the analysis of flake scars, and to Harry J. Shafer, Hiram F. (Pete)
+Gregory, Christian S. Hoggard, and David K. Thulman for their comments
+and constructive criticisms on the ongoing analyses of Caddo biface
+morphology, as well as Emma Sherratt, Kersten Bergstrom, Lauren Butaric,
+Dean C. Adams, and Michael L. Collyer for their constructive criticisms,
+general comments, and suggestions throughout the development of this
+research program.
 
-# Funding {-}
+\hypertarget{funding}{%
+\section*{Funding}\label{funding}}
+\addcontentsline{toc}{section}{Funding}
 
-Components of this analytical work flow were developed and funded by a Preservation Technology and Training grant (P14AP00138) to RZS from the National Center for Preservation Technology and Training (NCPTT), and additional grants to RZS from the Caddo Nation of Oklahoma, National Forests and Grasslands in Texas (15-PA-11081300-033) and the United States Forest Service (20-PA-11081300-074).
-Funding to analyse the bifaces from Millsap Cache and Jowell Farm was made possible by support from the Heritage Research Center at Stephen F.
-Austin State University.
-A Research Support Fund Grant from the Texas Archeological Society provided the tool needed to illustrate the bifaces.
+Components of this analytical work flow were developed and funded by a
+Preservation Technology and Training grant (P14AP00138) to RZS from the
+National Center for Preservation Technology and Training (NCPTT), and
+additional grants to RZS from the Caddo Nation of Oklahoma, National
+Forests and Grasslands in Texas (15-PA-11081300-033) and the United
+States Forest Service (20-PA-11081300-074). Funding to analyse the
+bifaces from Millsap Cache and Jowell Farm was made possible by support
+from the Heritage Research Center at Stephen F. Austin State University.
+A Research Support Fund Grant from the Texas Archeological Society
+provided the tool needed to illustrate the bifaces.
 
-# Data management
+\hypertarget{data-management}{%
+\section{Data management}\label{data-management}}
 
-Reproducibility—the ability to recompute results—and replicability—the chances other experimenters will achieve a consistent result—are two foundational characteristics of successful scientific research (Leek and Peng 2015).
-The analysis code associated with this project can be accessed through the [Supplementary Materials](https://seldenlab.github.io/elliptical.bifaces/), is available through the GitHub repository [https://github.com/seldenlab/elliptical.bifaces](https://github.com/seldenlab/elliptical.bifaces), and digitally curated on the Open Science Framework [DOI 10.17605/OSF.IO/PH25W](https://osf.io/ph25w/).
-The reproducible nature of this enterprise provides a means for others to critically assess and evaluate the various analytical components (Gray and Marwick 2019; Peng 2011; Gandrud 2014), which is a necessary requirement for the production of reliable knowledge.
+Reproducibility---the ability to recompute results---and
+replicability---the chances other experimenters will achieve a
+consistent result---are two foundational characteristics of successful
+scientific research (Leek and Peng 2015). The analysis code associated
+with this project can be accessed through the
+\href{https://seldenlab.github.io/elliptical.bifaces/}{Supplementary
+Materials}, is available through the GitHub repository
+\url{https://github.com/seldenlab/elliptical.bifaces}, and digitally
+curated on the Open Science Framework \href{https://osf.io/ph25w/}{DOI
+10.17605/OSF.IO/PH25W}. The reproducible nature of this enterprise
+provides a means for others to critically assess and evaluate the
+various analytical components (Gray and Marwick 2019; Peng 2011; Gandrud
+2014), which is a necessary requirement for the production of reliable
+knowledge.
 
-Reproducibility projects in psychology and cancer biology are impacting current research practices across all domains.
-Examples of reproducible research are becoming more abundant in archaeology (Marwick 2016; Ivanovaite et al.
-2020; Selden Jr., Dockall, and Shafer 2018; Selden Jr., Dockall, and Dubied 2020; Selden Jr et al.
-2021a; Selden 2022), and the next generation of archaeologists are learning those tools and methods needed to reproduce and/or replicate research results (Marwick et al.
-2019).
-Reproducible and replicable research work flows are often employed at the highest levels of humanities-based inquiries to mitigate concern or doubt regarding proper execution, and is of particular import should the results have—explicitly or implicitly—a major impact on scientific progress (Peels and Bouter 2018).
+Reproducibility projects in psychology and cancer biology are impacting
+current research practices across all domains. Examples of reproducible
+research are becoming more abundant in archaeology (Marwick 2016;
+Ivanovaite et al.~2020; Selden Jr., Dockall, and Shafer 2018; Selden
+Jr., Dockall, and Dubied 2020; Selden Jr et al.~2021a; Selden 2022), and
+the next generation of archaeologists are learning those tools and
+methods needed to reproduce and/or replicate research results (Marwick
+et al.~2019). Reproducible and replicable research work flows are often
+employed at the highest levels of humanities-based inquiries to mitigate
+concern or doubt regarding proper execution, and is of particular import
+should the results have---explicitly or implicitly---a major impact on
+scientific progress (Peels and Bouter 2018).
+
+\bibliographystyle{tfcad}
+\bibliography{interactcadsample.bib}
+
+\input{"appendix.tex"}
+
+\end{document}

### 2022-11-01 ms/ms/ms.Rmd
<edit ms>

@@
   Size matters: Divergent morphology in elliptical bifaces from sites in the American Southeast articulates with distinct local reduction practices
-type: ARTICLE TEMPLATE
+type: ORIGINAL RESEARCH ARTICLE
 author:
@@
     email: john.dockall@stantec.com
-  - name: David H. Dye
+  - name: and David H. Dye
     affil: c
@@
 abstract: |
-  Elliptical bifaces are prevalent at Mississippian sites throughout the American Southeast.
-Those from Millsap Cache and Jowell Farm (41AN13) comprise two of the largest samples of this ill-understood stone tool from the ancestral Caddo area.
-Elliptical bifaces from Millsap Cache were produced using Kay County flint, while those from Jowell Farm were manufactured using Edwards chert, providing for an empirical test of morphological differences as a function of raw material.
+  Elliptical bifaces are prevalent at Mississipian sites throughout the American Southeast.
+Those from Millsap Cache and Jowell Farm comprise two of the largest samples of this ill-understood stone tool from the ancestral Caddo area.
+Elliptical bifaces from Millsap Cache were reportedly produced using Kay County flint, while those from Jowell Farm were manufactured using Edwards chert, providing for an empirical test of morphological differences as a function of raw material.
 The sample of elliptical bifaces was divided into two size classes; one conceptually reflective of production (large), and the other with local reduction practices (small).
-Size classes were used to assess whether modifications by Caddo knappers may have yielded similar—convergent—biface shape in the small class.
+Size classes were used to assess whether modifications by Caddo knappers may have yielded similar---convergent---biface shapes in the small class.
 Size classes were also used to test the hypothesis that greater morphological variation would be apparent in the small class due to idiosyncratic responses related to local retouch practices.
@@
 keywords: |
-  American Southeast; Caddo; NAGPRA; archaeology; ovoid biface; Jowell knife; Jowell knives; lithics; oblique parallel flaking; archaeoinformatics; museum studies; digital humanities; non-western art history; geometric morphometrics; STEM; STEAM
+  American Southeast; Caddo; NAGPRA; archaeology; ovoid biface; Jowell knife; Jowell knives; Jowell biface; lithics; oblique parallel flaking; archaeoinformatics; museum studies; digital humanities; non-western art history; geometric morphometrics; STEM; STEAM
 header-includes: |
@@
 
-\begin{figure}\centering
-\includegraphics[width=\linewidth]{figs/elliptical.illustration.png}
-\caption{Illustrated elliptical bifaces from a, an unknown location; b, Jowell Farm; and c, Belcher Mound highlighting oblique parallel flaking used in production.
-The specimen from Belcher Mound is a potential preform exhibiting oblique parallel removals consistent with elliptical bifaces from other sites in the region.}
-\label{fig:illustrated}
-\end{figure}
+# Methods and results
 
-# Methods
-
-To assess the variable impacts that Caddo retouch practices may have had on general morphology, the sample of elliptical bifaces was binned into two size classes using centroid size.
+To assess the variable impacts that Caddo retouch practices may have had on general morphology, the sample of elliptical bifaces was first binned into two size classes using centroid size ([Supplementary Materials](https://seldenlab.github.io/elliptical.bifaces/)).
 Due to potential morphological differences associated with the use of two distinct raw materials (Kay County flint and Edwards chert), the sample was subset by site prior to calculating the mean centroid size for each.
-Bifaces larger than the mean were then assigned to the large category, those smaller than the mean were assigned to the small category, and the two datasets were then joined in advance of analysis ([Supplementary Materials, Ch.
-1](https://seldenlab.github.io/elliptical.bifaces/define-size-classes.html)).
+Raw material for bifaces from Millsap Cache were identified by \citet{RN11461} as Kay County flint, and those from Jowell Farm were identified as Edwards Chert due to their size, the colours that they fluoresce beneath low and high frequency UV light, and the paucity of similarly large locally-available raw material sources.
+Bifaces larger than the mean were then assigned to the large category, those smaller than the mean were assigned to the small category, and the two datasets were then joined in advance of analysis ([Supplementary Materials](https://seldenlab.github.io/elliptical.bifaces/)).
 
@@
 \includegraphics[width=\linewidth]{figs/ellipticalbifaces.png}
-\caption{Selected elliptical bifaces from the Jowell Farm site highlighting the most heavily retouched area identified using a modified version of the flaking index developed and advanced by \citet{RN11099}, \citet{RN9242}, and \citet{RN11098}.}
+\caption{Selected elliptical bifaces from the Jowell Farm site highlighting the most heavily retouched area at top right identified using a modified version of the flaking index developed and advanced by \citet{RN11099}, \citet{RN9242}, and \citet{RN11098}.}
 \label{fig:elliptical}
@@
 To identify which edge was most heavily retouched, we employed a modified approach to the flaking index initially developed by \citet{RN11099}, and later advanced and refined by \citet{RN9242}, then \citet{RN11098}.
-The modified approach uses counts of flake scars from each edge in the two most heavily worked areas, paired with a measure of edge length inclusive of curvature, which was calculated using ImageJ (Abramoff et al.
-2004, Rasband 1997, Schneider et al.
-2012).
+The modified approach uses counts of flake scars from each edge in the two most heavily worked areas, paired with a measure of edge length inclusive of curvature, which was calculated using ImageJ \citep{RN11146,RN11147,RN11148}.
 The number of flake scars was then divided by the length of the worked edge, with the most heavily worked edge identified by the greater value.
@@
 
-The same landmark/semilandmark configuration was used for both analyses, and the landmarking protocol employs three landmarks; two horizontal tangents (top/bottom), and the third placed at the furthest extent of the retouched edge bearing the heaviest amount of retouch (see Figure \ref{fig:elliptical}).
-Equidistant semilandmarks were placed between each landmark, and all landmarks and semilandmarks were applied using R 4.2.1 (R Core Development Team, 2022) and the StereoMorph package (Olsen and Westneat 2015).
+The same landmark/semilandmark configuration was used for both analyses, and the landmarking protocol employs three landmarks; two horizontal tangents (top/bottom), and the third placed at the furthest extent of the retouched edge bearing the heaviest amount of retouch (see Figure \@ref(fig:elliptical)).
+Equidistant semilandmarks were placed between each landmark, and all landmarks and semilandmarks were applied using R \citep{R} and the `StereoMorph` package \citep{RN9091}.
 
-Landmark data were aligned to a global coordinate system (Bookstein et al.
-1999, Gunz et al.
-2005, Kendall 1981, 1984, Slice 2001), achieved through generalised Procrustes superimposition (Bookstein 1986, Rohlf and Slice 1990, Rohlf 1999) in R using the geomorph (Adams et al.
-2017, Adams and Otarola-Castillo 2013, Baken et al.
-2021) and RRPP packages (Adams and Collyer 2015, Collyer and Adams 2018).
-Procrustes superimposition translates and rotates the coordinate data to allow for comparisons among objects, while also scaling each biface using unit-centroid size—the square root of the sum of squared distances from each landmark to the specimen’s centroid (Chapman 1990, Dryden and Mardia 1998, Gower 1975, Rohlf and Slice 1990).
-The geomorph package uses a partial Procrustes superimposition that projects the aligned specimens into tangent space subsequent to alignment in preparation for the use of multivariate methods that assume linear space (Dryden and Mardia 1993, Kent and Mardia 2001, Rohlf 1999, Slice 2001).
+### Generalised Procrustes Analysis
+
+Landmark data were aligned to a global coordinate system \citep{RN8477,RN7502,RN11622,RN11623,RN11563}, achieved through generalised Procrustes superimposition \citep{RN11138,RN478,RN1646} in R using the `geomorph` and `RRPP` packages \citep{RN1655,RN11775,RN11530,RN1774,RN9565}.
+Procrustes superimposition translates and rotates the coordinate data to allow for comparisons among objects, while also scaling each biface using unit-centroid size---the square root of the sum of squared distances from each landmark to the specimen’s centroid \citep{RN11139,RN11140,RN11564,RN478}.
+The `geomorph` package uses a partial Procrustes superimposition that projects the aligned specimens into tangent space subsequent to alignment in preparation for the use of multivariate methods that assume linear space (Figure \ref{fig:gpa}) \citep{RN11141,RN11142,RN1646,RN11563}.
+
+\begin{figure}\centering
+\includegraphics[width=\linewidth]{figs/gpa.png}
+\caption{Results of generalised Procrustes analysis for elliptical bifaces from the Millsap Cache and Jowell Farm. Mean consensus configuration shown in black; individual bifaces in gray.}
+\label{fig:gpa}
+\end{figure}
+
+### Principal Components Analysis
+
+Principal components analysis \citep{RN1746} was used to visualise shape variation among the elliptical bifaces, and the scatterplot and convex hulls represent the dispersion of shapes in tangent space (Figure \ref{fig:pca}) \citep{RN8633,RN5616,RN11143,RN7550}.
+The shape space described by each principal axis is commonly visualized using thin-plate spline warping of a reference image \citep{RN1731,RN479}.
+
+\begin{figure}\centering
+\includegraphics[width=\linewidth]{figs/pca.png}
+\caption{PCA summarizing shape variation in elliptical bifaces by site and size class. }
+\label{fig:pca}
+\end{figure}
+
+### Procrustes ANOVA
+
+To assess whether shape changes with size (allometry), and whether shape and size differed by raw material and size class, Procrustes ANOVAs \citep{RN1749} were run that enlist effect-sizes (zscores) computed as standard deviates of the generated sampling distributions \citep{RN1756}.
+A residual randomization permutation procedure (RRPP; n = 10,000 permutations) was used for all Procrustes ANOVAs \citep{RN1655,RN11775}, which has higher statistical power and a greater ability to identify patterns in the data should they be present \citep{RN1719}.
+
+The test for allometry in the sample did not return a significant result, suggesting that elliptical biface shape and size may follow a pattern of isometry, where shape remains constant through the reduction process ([Supplementary Materials](https://seldenlab.github.io/elliptical.bifaces/)).
+Elliptical bifaces were also found to express a consistency of shape across all analyses ([Supplementary Materials](https://seldenlab.github.io/elliptical.bifaces/)).
+Results demonstrated a significant difference in size by site/raw material (RRPP = 10,000, Pr(>F) = 0.0007); however, the subsequent analysis of elliptical bifaces by size class demonstrates that the pattern is more readily attributable to reduction practices/intensity rather than nodule and raw material size ([Supplementary Materials](https://seldenlab.github.io/elliptical.bifaces/)).
+
+The analysis of morphology by size class demonstrates a significant difference in size between size classes; however, that pattern deviated from our expectations.
+Elliptical bifaces from Jowell Farm differ in neither shape nor size between the size classes ([Supplementary Materials](https://seldenlab.github.io/elliptical.bifaces/)).
+Elliptical bifaces in the large and small size classes from Millsap Cache differ significantly in size (RRPP = 10,000, Pr>d = 0.0035), but not shape ([Supplementary Materials](https://seldenlab.github.io/elliptical.bifaces/)).
+Pairwise comparisons demonstrate that elliptical bifaces from Millsap Cache and Jowell Farm in the large size class are morphologically consistent, but those in the small size class differ in size (RRPP = 10,000, Pr>d = 0.0001) ([Supplementary Materials](https://seldenlab.github.io/elliptical.bifaces/)).
+Thus, the pattern that emerges is one of differential retouch _intensity_, where isometric reduction strategies of one group (Millsap Cache) may correlate with differing retouch practices that articulate with locally abundant raw material (Kay County flint).
+This stands in contrast to the comparatively minimal retouch intensity of another group (Jowell Farm) where local raw material size is insufficient, making it necessary to import the requisite raw material from an extralocal source; in this case, Edwards chert from central Texas.
+
+### Morphological disparity
+
+An analysis of morphological disparity was used to identify potential differences in morphological diversity between size classes \citep{RN11107,RN7041,RN5694}.
+No difference was found for elliptical biface shape or size by site/raw material, but the comparison of morphological diversity by size class demonstrated that biface shapes in the small size class from Jowell Farm occupy a significantly greater range of morphospace than those from Millsap Cache.
+This pattern suggests that the comparatively minimal retouch intensity associated with Caddo knappers at Jowell Farm may have been more targeted---potentially due to a lack of abundant local raw material---resulting in a broader range of biface shapes in the small size class.
+Conversely, these results also suggest that Caddo knappers at Millsap Cache employed a more standardised (specialised?) approach to reduction practices.
+
+# Discussion and conclusion
 
@@
 Austin State University.
-A Research Support Fund Grant from the Texas Archeological Society provided the tool needed to illustrate the bifaces.
 

### 2022-11-01 ms/ms/ms.tex
<edit ms>

@@
 
-\articletype{ARTICLE TEMPLATE}
+\articletype{ORIGINAL RESEARCH ARTICLE}
 
@@
 
-\author{\name{Robert Z. Selden, Jr.$^{a}$, John E. Dockall$^{b}$, David
-H. Dye$^{c}$}
+\author{\name{Robert Z. Selden, Jr.$^{a}$, John E. Dockall$^{b}$, and
+David H. Dye$^{c}$}
 \affil{$^{a}$Heritage Research Center, Stephen F. Austin State
@@
 E.
-Dockall. Email: \href{mailto:john.dockall@stantec.com}{\nolinkurl{john.dockall@stantec.com}}, David
-H.
+Dockall. Email: \href{mailto:john.dockall@stantec.com}{\nolinkurl{john.dockall@stantec.com}}, and
+David H.
 Dye. Email: \href{mailto:daviddye@memphis.edu}{\nolinkurl{daviddye@memphis.edu}}}
@@
 \begin{abstract}
-Elliptical bifaces are prevalent at Mississippian sites throughout the
-American Southeast. Those from Millsap Cache and Jowell Farm (41AN13)
-comprise two of the largest samples of this ill-understood stone tool
-from the ancestral Caddo area. Elliptical bifaces from Millsap Cache
-were produced using Kay County flint, while those from Jowell Farm were
-manufactured using Edwards chert, providing for an empirical test of
-morphological differences as a function of raw material. The sample of
-elliptical bifaces was divided into two size classes; one conceptually
-reflective of production (large), and the other with local reduction
-practices (small). Size classes were used to assess whether
+Elliptical bifaces are prevalent at Mississipian sites throughout the
+American Southeast. Those from Millsap Cache and Jowell Farm comprise
+two of the largest samples of this ill-understood stone tool from the
+ancestral Caddo area. Elliptical bifaces from Millsap Cache were
+reportedly produced using Kay County flint, while those from Jowell Farm
+were manufactured using Edwards chert, providing for an empirical test
+of morphological differences as a function of raw material. The sample
+of elliptical bifaces was divided into two size classes; one
+conceptually reflective of production (large), and the other with local
+reduction practices (small). Size classes were used to assess whether
 modifications by Caddo knappers may have yielded
-similar---convergent---biface shape in the small class. Size classes
+similar---convergent---biface shapes in the small class. Size classes
 were also used to test the hypothesis that greater morphological
@@
 American Southeast; Caddo; NAGPRA; archaeology; ovoid biface; Jowell
-knife; Jowell knives; lithics; oblique parallel flaking;
+knife; Jowell knives; Jowell biface; lithics; oblique parallel flaking;
 archaeoinformatics; museum studies; digital humanities; non-western art
@@
 
-\begin{figure}\centering
-\includegraphics[width=\linewidth]{figs/elliptical.illustration.png}
-\caption{Illustrated elliptical bifaces from a, an unknown location; b, Jowell Farm; and c, Belcher Mound highlighting oblique parallel flaking used in production.
-The specimen from Belcher Mound is a potential preform exhibiting oblique parallel removals consistent with elliptical bifaces from other sites in the region.}
-\label{fig:illustrated}
-\end{figure}
-
-\hypertarget{methods}{%
-\section{Methods}\label{methods}}
+\hypertarget{methods-and-results}{%
+\section{Methods and results}\label{methods-and-results}}
 
 To assess the variable impacts that Caddo retouch practices may have had
-on general morphology, the sample of elliptical bifaces was binned into
-two size classes using centroid size. Due to potential morphological
-differences associated with the use of two distinct raw materials (Kay
-County flint and Edwards chert), the sample was subset by site prior to
-calculating the mean centroid size for each. Bifaces larger than the
-mean were then assigned to the large category, those smaller than the
-mean were assigned to the small category, and the two datasets were then
-joined in advance of analysis
-(\href{https://seldenlab.github.io/elliptical.bifaces/define-size-classes.html}{Supplementary
-Materials, Ch. 1}).
+on general morphology, the sample of elliptical bifaces was first binned
+into two size classes using centroid size
+(\href{https://seldenlab.github.io/elliptical.bifaces/}{Supplementary
+Materials}). Due to potential morphological differences associated with
+the use of two distinct raw materials (Kay County flint and Edwards
+chert), the sample was subset by site prior to calculating the mean
+centroid size for each. Raw material for bifaces from Millsap Cache were
+identified by \citet{RN11461} as Kay County flint, and those from Jowell
+Farm were identified as Edwards Chert due to their size, the colours
+that they fluoresce beneath low and high frequency UV light, and the
+paucity of similarly large locally-available raw material sources.
+Bifaces larger than the mean were then assigned to the large category,
+those smaller than the mean were assigned to the small category, and the
+two datasets were then joined in advance of analysis
+(\href{https://seldenlab.github.io/elliptical.bifaces/}{Supplementary
+Materials}).
 
@@
 edge length inclusive of curvature, which was calculated using ImageJ
-(Abramoff et al.~2004, Rasband 1997, Schneider et al.~2012). The number
-of flake scars was then divided by the length of the worked edge, with
-the most heavily worked edge identified by the greater value.
+\citep{RN11146,RN11147,RN11148}. The number of flake scars was then
+divided by the length of the worked edge, with the most heavily worked
+edge identified by the greater value.
 
@@
 the retouched edge bearing the heaviest amount of retouch (see Figure
-\ref{fig:elliptical}). Equidistant semilandmarks were placed between
+@ref(fig:elliptical)). Equidistant semilandmarks were placed between
 each landmark, and all landmarks and semilandmarks were applied using R
-4.2.1 (R Core Development Team, 2022) and the StereoMorph package (Olsen
-and Westneat 2015).
+\citep{R} and the \texttt{StereoMorph} package \citep{RN9091}.
 
-Landmark data were aligned to a global coordinate system (Bookstein et
-al.~1999, Gunz et al.~2005, Kendall 1981, 1984, Slice 2001), achieved
-through generalised Procrustes superimposition (Bookstein 1986, Rohlf
-and Slice 1990, Rohlf 1999) in R using the geomorph (Adams et al.~2017,
-Adams and Otarola-Castillo 2013, Baken et al.~2021) and RRPP packages
-(Adams and Collyer 2015, Collyer and Adams 2018). Procrustes
-superimposition translates and rotates the coordinate data to allow for
-comparisons among objects, while also scaling each biface using
-unit-centroid size---the square root of the sum of squared distances
-from each landmark to the specimen's centroid (Chapman 1990, Dryden and
-Mardia 1998, Gower 1975, Rohlf and Slice 1990). The geomorph package
+\hypertarget{generalised-procrustes-analysis}{%
+\subsubsection{Generalised Procrustes
+Analysis}\label{generalised-procrustes-analysis}}
+
+Landmark data were aligned to a global coordinate system
+\citep{RN8477,RN7502,RN11622,RN11623,RN11563}, achieved through
+generalised Procrustes superimposition \citep{RN11138,RN478,RN1646} in R
+using the \texttt{geomorph} and \texttt{RRPP} packages
+\citep{RN1655,RN11775,RN11530,RN1774,RN9565}. Procrustes superimposition
+translates and rotates the coordinate data to allow for comparisons
+among objects, while also scaling each biface using unit-centroid
+size---the square root of the sum of squared distances from each
+landmark to the specimen's centroid
+\citep{RN11139,RN11140,RN11564,RN478}. The \texttt{geomorph} package
 uses a partial Procrustes superimposition that projects the aligned
 specimens into tangent space subsequent to alignment in preparation for
-the use of multivariate methods that assume linear space (Dryden and
-Mardia 1993, Kent and Mardia 2001, Rohlf 1999, Slice 2001).
+the use of multivariate methods that assume linear space (Figure
+\ref{fig:gpa}) \citep{RN11141,RN11142,RN1646,RN11563}.
+
+\begin{figure}\centering
+\includegraphics[width=\linewidth]{figs/gpa.png}
+\caption{Results of generalised Procrustes analysis for elliptical bifaces from the Millsap Cache and Jowell Farm. Mean consensus configuration shown in black; individual bifaces in gray.}
+\label{fig:gpa}
+\end{figure}
+
+\hypertarget{principal-components-analysis}{%
+\subsubsection{Principal Components
+Analysis}\label{principal-components-analysis}}
+
+Principal components analysis \citep{RN1746} was used to visualise shape
+variation among the elliptical bifaces, and the scatterplot and convex
+hulls represent the dispersion of shapes in tangent space (Figure
+\ref{fig:pca}) \citep{RN8633,RN5616,RN11143,RN7550}. The shape space
+described by each principal axis is commonly visualized using thin-plate
+spline warping of a reference image \citep{RN1731,RN479}.
+
+\begin{figure}\centering
+\includegraphics[width=\linewidth]{figs/pca.png}
+\caption{PCA summarizing shape variation in elliptical bifaces by site and size class. }
+\label{fig:pca}
+\end{figure}
+
+\hypertarget{procrustes-anova}{%
+\subsubsection{Procrustes ANOVA}\label{procrustes-anova}}
+
+To assess whether shape changes with size (allometry), and whether shape
+and size differed by raw material and size class, Procrustes ANOVAs
+\citep{RN1749} were run that enlist effect-sizes (zscores) computed as
+standard deviates of the generated sampling distributions
+\citep{RN1756}. A residual randomization permutation procedure (RRPP; n
+= 10,000 permutations) was used for all Procrustes ANOVAs
+\citep{RN1655,RN11775}, which has higher statistical power and a greater
+ability to identify patterns in the data should they be present
+\citep{RN1719}.
+
+The test for allometry in the sample did not return a significant
+result, suggesting that elliptical biface shape and size may follow a
+pattern of isometry, where shape remains constant through the reduction
+process
+(\href{https://seldenlab.github.io/elliptical.bifaces/}{Supplementary
+Materials}). Elliptical bifaces were also found to express a consistency
+of shape across all analyses
+(\href{https://seldenlab.github.io/elliptical.bifaces/}{Supplementary
+Materials}). Results demonstrated a significant difference in size by
+site/raw material (RRPP = 10,000, Pr(\textgreater F) = 0.0007); however,
+the subsequent analysis of elliptical bifaces by size class demonstrates
+that the pattern is more readily attributable to reduction
+practices/intensity rather than nodule and raw material size
+(\href{https://seldenlab.github.io/elliptical.bifaces/}{Supplementary
+Materials}).
+
+The analysis of morphology by size class demonstrates a significant
+difference in size between size classes; however, that pattern deviated
+from our expectations. Elliptical bifaces from Jowell Farm differ in
+neither shape nor size between the size classes
+(\href{https://seldenlab.github.io/elliptical.bifaces/}{Supplementary
+Materials}). Elliptical bifaces in the large and small size classes from
+Millsap Cache differ significantly in size (RRPP = 10,000,
+Pr\textgreater d = 0.0035), but not shape
+(\href{https://seldenlab.github.io/elliptical.bifaces/}{Supplementary
+Materials}). Pairwise comparisons demonstrate that elliptical bifaces
+from Millsap Cache and Jowell Farm in the large size class are
+morphologically consistent, but those in the small size class differ in
+size (RRPP = 10,000, Pr\textgreater d = 0.0001)
+(\href{https://seldenlab.github.io/elliptical.bifaces/}{Supplementary
+Materials}). Thus, the pattern that emerges is one of differential
+retouch \emph{intensity}, where isometric reduction strategies of one
+group (Millsap Cache) may correlate with differing retouch practices
+that articulate with locally abundant raw material (Kay County flint).
+This stands in contrast to the comparatively minimal retouch intensity
+of another group (Jowell Farm) where local raw material size is
+insufficient, making it necessary to import the requisite raw material
+from an extralocal source; in this case, Edwards chert from central
+Texas.
+
+\hypertarget{morphological-disparity}{%
+\subsubsection{Morphological disparity}\label{morphological-disparity}}
+
+An analysis of morphological disparity was used to identify potential
+differences in morphological diversity between size classes
+\citep{RN11107,RN7041,RN5694}. No difference was found for elliptical
+biface shape or size by site/raw material, but the comparison of
+morphological diversity by size class demonstrated that biface shapes in
+the small size class from Jowell Farm occupy a significantly greater
+range of morphospace than those from Millsap Cache. This pattern
+suggests that the comparatively minimal retouch intensity associated
+with Caddo knappers at Jowell Farm may have been more
+targeted---potentially due to a lack of abundant local raw
+material---resulting in a broader range of biface shapes in the small
+size class. Conversely, these results also suggest that Caddo knappers
+at Millsap Cache employed a more standardised (specialised?) approach to
+reduction practices.
+
+\hypertarget{discussion-and-conclusion}{%
+\section{Discussion and conclusion}\label{discussion-and-conclusion}}
 
@@
 from the Heritage Research Center at Stephen F. Austin State University.
-A Research Support Fund Grant from the Texas Archeological Society
-provided the tool needed to illustrate the bifaces.
 

### 2022-11-01 supp/index.Rmd
<edit ms>
<edit index>

@@
 Austin State University; Department of Biology, Stephen F.
-Austin State University; Texas Archeological Research Laboratory, The University of Texas at Austin; and Cultural Heritage Department, Jean Monnet University, zselden@sfasu.edu] and John E.
-Dockall^[Stantec, Inc.]
+Austin State University; Texas Archeological Research Laboratory, The University of Texas at Austin; and Cultural Heritage Department, Jean Monnet University, zselden@sfasu.edu]
 date: "`r Sys.Date()`"
@@
 The number of flake scars was then divided by the length of the worked edge, with the heaviest worked edge identified by the greater value.
-
-The longest reworked edge on the elliptical bifaces was almost always more heavily retouched.
-This finding was used to incorporate the images of elliptical bifaces from Millsap Cache [@RN11461], where the longest reworked edge was oriented at top right prior to landmarking.
 

### 2022-11-02 supp/index.Rmd
<edit index>

@@
 Elliptical bifaces are prevalent at Mississipian sites throughout the American Southeast.
-Those from Millsap Cache and Jowell Farm (41AN13) comprise two of the largest samples of this ill-understood stone tool from the ancestral Caddo area.
-Elliptical bifaces from Millsap Cache were reportedly produced using Kay County flint, while those from Jowell Farm were manufactured using Edwards chert, providing for an empirical test of morphological differences as a function of raw material.
+Those from Millsap Cache and Jowell Farm comprise two of the largest samples of this ill-understood stone tool from the ancestral Caddo area.
+Bifaces from Millsap Cache were reportedly produced using Kay County flint, while those from Jowell Farm were manufactured using Edwards chert, providing for an empirical test of morphological differences as a function of raw material.
 The sample of elliptical bifaces was divided into two size classes; one conceptually reflective of production (large), and the other with local reduction practices (small).
@@
 Size classes were also used to test the hypothesis that greater morphological variation would be apparent in the small class due to idiosyncratic responses related to local retouch practices.
-Results demonstrate that elliptical biface shape does not differ by raw material, but size does.
-This suggests that a shared, and consistent, mental template was maintained independent of biface size.
+Results demonstrate that elliptical biface shape does not differ by raw material, but size does, suggesting that a shared and morphologically-consistent mental template was maintained independent of biface size.
 The subsequent analysis of elliptical biface morphology by size class demonstrated that size does not differ by raw material in the large class, but in the small class, it does.
@@
 
-American Southeast, Caddo, NAGPRA, archaeoinformatics, museum studies, digital humanities, geometric morphometrics, STEM, STEAM, ovoid biface, Jowell knife, Jowell knives, lithics, oblique parallel flaking
+American Southeast, Caddo, NAGPRA, archaeoinformatics, museum studies, digital humanities, non-Western art history, geometric morphometrics, STEM, STEAM, ovoid biface, Jowell knife, Jowell knives, lithics, oblique parallel flaking
 
-```{r figmap, fig.cap="Location of Millsap Cache, Jowell Farm, and other sites mentioned in the manuscript.
-Elliptical bifaces have been found in every major river basin in the ancestral Caddo area (white)."}
+```{r figmap, fig.cap="Location of Millsap Cache, Jowell Farm, and other Caddo sites mentioned in the text.
+Elliptical bifaces have been found in every major river basin in the ancestral Caddo area (dashed/white)."}
 
@@
 Our thanks to the Caddo Nation of Oklahoma, the Caddo Nation Tribal Council, Tribal Chairman, and Tribal Historic Preservation Office for permission and access to NAGPRA and previously repatriated collections.
-Our gratitude is also extended to Marybeth Tomka and Lauren Bussiere at the Texas Archeological Research Laboratory for their assistance with access to the bifaces and associated records, to Sergio Ayala and the Gault School of Archaeological Research for access to the UV light, and to Scott Hammerstedt and Debra K.
-Green at the Oklahoma Archeological Survey for their assistance with records requests.
-Thanks also to John Harman for access to the DStretch plugin for ImageJ that was useful in the analysis of flake scars, and to Harry J.
+Our gratitude is also extended to Marybeth Tomka and Lauren Bussiere at the Texas Archeological Research Laboratory for their assistance with access to the bifaces and associated records, to Sergio Ayala and the Gault School of Archaeological Research for access to the UV light, to Scott Hammerstedt and Debra K.
+Green at the Oklahoma Archeological Survey for their assistance with records requests, and to Marc Lavine and Susie Fishman-Armstrong at the Sam Noble Oklahoma Museum of Natural History for providing digital images from their collections.
+Thanks also to John Harman for access to the DStretch plugin for ImageJ, which was useful in the analysis of flake scars, and to Harry J.
 Shafer, Hiram F.

### 2022-11-02 ms/ms/ms.Rmd
<edit index>

@@
 Those from Millsap Cache and Jowell Farm comprise two of the largest samples of this ill-understood stone tool from the ancestral Caddo area.
-Elliptical bifaces from Millsap Cache were reportedly produced using Kay County flint, while those from Jowell Farm were manufactured using Edwards chert, providing for an empirical test of morphological differences as a function of raw material.
+Bifaces from Millsap Cache were reportedly produced using Kay County flint, while those from Jowell Farm were manufactured using Edwards chert, providing for an empirical test of morphological differences as a function of raw material.
 The sample of elliptical bifaces was divided into two size classes; one conceptually reflective of production (large), and the other with local reduction practices (small).
-Size classes were used to assess whether modifications by Caddo knappers may have yielded similar---convergent---biface shapes in the small class.
+Size classes were used to assess whether modifications by Caddo knappers may have yielded similar---convergent---biface shape in the small class.
 Size classes were also used to test the hypothesis that greater morphological variation would be apparent in the small class due to idiosyncratic responses related to local retouch practices.
-Results demonstrate that elliptical biface shape does not differ by raw material, but size does.
-This suggests that a shared, and consistent, mental template was maintained independent of biface size.
+Results demonstrate that elliptical biface shape does not differ by raw material, but size does, suggesting that a shared and morphologically-consistent mental template was maintained independent of biface size.
 The subsequent analysis of elliptical biface morphology by size class demonstrated that size does not differ by raw material in the large class, but in the small class, it does.
@@
 keywords: |
-  American Southeast; Caddo; NAGPRA; archaeology; ovoid biface; Jowell knife; Jowell knives; Jowell biface; lithics; oblique parallel flaking; archaeoinformatics; museum studies; digital humanities; non-western art history; geometric morphometrics; STEM; STEAM
+  American Southeast; Caddo; NAGPRA; archaeology; ovoid biface; Jowell knife; Jowell knives; Jowell biface; lithics; oblique parallel flaking; archaeoinformatics; museum studies; digital humanities; non-Western art history; geometric morphometrics; STEM; STEAM
 header-includes: |
@@
 \includegraphics[width=\linewidth]{figs/map.png}
-\caption{Location of Millsap Cache, Jowell Farm, and other sites mentioned in the manuscript. Elliptical bifaces have been found in every major river basin in the ancestral Caddo area (white).}
+\caption{Location of Millsap Cache, Jowell Farm, and other sites mentioned in the text. Elliptical bifaces have been found in every major drainage basin in the ancestral Caddo area (white).}
 \label{fig:map}
@@
 
-Elliptical bifaces from the Millsap Cache and Jowell Farm were used to examine whether biface morphology remains stable or expresses morphological variability by site/raw material ([Supplementary Materials, Ch.
-2](https://seldenlab.github.io/elliptical.bifaces/gm---siteraw-material.html)).
+Elliptical bifaces from the Millsap Cache and Jowell Farm were used to examine whether biface morphology remains stable or expresses morphological variability by site/raw material ([Supplementary Materials](https://seldenlab.github.io/elliptical.bifaces/gm---siteraw-material.html)).
 Prior to landmarking, elliptical bifaces were oriented with the most heavily retouched edge at top right (Figure  \ref{fig:elliptical}).
@@
 Components of this analytical work flow were developed and funded by a Preservation Technology and Training grant (P14AP00138) to RZS from the National Center for Preservation Technology and Training (NCPTT), and additional grants to RZS from the Caddo Nation of Oklahoma, National Forests and Grasslands in Texas (15-PA-11081300-033) and the United States Forest Service (20-PA-11081300-074).
-Funding to analyse the bifaces from Millsap Cache and Jowell Farm was made possible by support from the Heritage Research Center at Stephen F.
+Funding to analyse the bifaces from Millsap Cache and Jowell Farm was provided by the Heritage Research Center at Stephen F.
 Austin State University.

### 2022-11-02 supp/index.Rmd
<edit index>

--- 2022-10-18 supp/index.Rmd
@@
 
-Elliptical bifaces are prevalent at Mississipian sites throughout the American Southeast.
+Elliptical bifaces are regularly found at Mississipian sites throughout the American Southeast.
 Those from Millsap Cache and Jowell Farm (41AN13) comprise two of the largest samples of this ill-understood stone tool from the ancestral Caddo area.
-Elliptical bifaces from Millsap Cache were reportedly produced using Kay County flint, while those from Jowell Farm were manufactured using Edwards chert, providing for an empirical test of morphological differences as a function of raw material.
+Elliptical bifaces from Millsap Cache were reportedly produced using Kay County flint, while those from Jowell Farm were manufactured using Edwards chert, providing for an empirical test of whether biface morphology differs as a function of raw material.
 The sample of elliptical bifaces was divided into two size classes; one conceptually reflective of production (large), and the other with local reduction practices (small).
-Size classes were used to assess whether modifications by Caddo knappers may have yielded similar---convergent---biface shape in the small class.
+Size classes were used to assess whether modifications by Caddo knappers may have yielded similar---convergent---biface shapes in the small class.
 Size classes were also used to test the hypothesis that greater morphological variation would be apparent in the small class due to idiosyncratic responses related to local retouch practices.
@@
 
-American Southeast, Caddo, NAGPRA, archaeoinformatics, museum studies, digital humanities, geometric morphometrics, STEM, STEAM, ovoid biface, Jowell knife, Jowell knives, lithics, oblique parallel flaking
+American Southeast, Caddo, archaeology, anthroinformatics, archaeoinformatics, museum studies, digital humanities, geometric morphometrics, STEM, STEAM, Jowell knife, Jowell knives, lithics, oblique parallel flaking
 
@@
 
-The same landmark/semilandmark configuration was used for both analyses, and the landmarking protocol employs three landmarks; two horizontal tangents (top/bottom), and the third placed at the furthest extent of the retouched edge bearing the heaviest amount of retouch (see Figure \@ref(fig:figlm)).
-Equidistant semilandmarks were placed between each landmark, and all landmarks and semilandmarks were applied using R [@R] and the `StereoMorph` package [@RN9091].
-
 ### Generalised Procrustes Analysis
 
-Landmark data were aligned to a global coordinate system [@RN8477;@RN7502;@RN11622;@RN11623;@RN11563], achieved through generalised Procrustes superimposition [@RN11138;@RN478;@RN1646] in R using the `geomorph` and `RRPP` packages [@RN1655;@RN11775;@RN11530;@RN1774;@RN9565].
-Procrustes superimposition translates and rotates the coordinate data to allow for comparisons among objects, while also scaling each biface using unit-centroid size---the square root of the sum of squared distances from each landmark to the specimen’s centroid [@RN11139;@RN11140;@RN11564;@RN478].
-The `geomorph` package uses a partial Procrustes superimposition that projects the aligned specimens into tangent space subsequent to alignment in preparation for the use of multivariate methods that assume linear space [@RN11141;@RN11142;@RN1646;@RN11563].
+Landmark data were aligned to a global coordinate system [@RN11622;@RN11623;@RN11563], achieved through generalized Procrustes superimposition [@RN478] performed in R 4.2.1 [@R] using the `geomorph` library v.
+4.0.4 [@RN11530;@RN1774;@RN9565].
+Procrustes superimposition translates, scales, and rotates the coordinate data to allow for comparisons among objects [@RN11564;@RN478].
+The `geomorph` package uses a partial Procrustes superimposition that projects the aligned specimens into tangent space following alignment in preparation for the use of multivariate methods that assume linear space [@RN1646;@RN11563].
 
@@
 
-Principal components analysis [@RN1746] was used to visualise shape variation among the elliptical bifaces, and the scatterplot represents the dispersion of shapes in tangent space [@RN8633;@RN5616;@RN11143;@RN7550].
-Those shape changes described by each principal axis are commonly visualized using thin-plate spline warping of a reference image [@RN1731;@RN479].
+Principal components analysis [@RN1746] was used to visualise shape variation among the elliptical bifaces.
+The shape changes described by each principal axis are commonly visualized using thin-plate spline warping of a reference image or 3D mesh [@RN1731;@RN479].
 
@@
   -   Bifaces in the _small class_ recovered from Millsap Cache and Jowell Farm **differ significantly** in size.
-Thus, while shape remains consistent, size differs, suggesting---at minimum---locally divergent approaches to retouch or retouch intensity.
+Thus, while shape remains consistent, size differs, suggesting---at minimum---locally divergent approaches to retouch.
   -   Morphological disparity in the shape of the _small class_ is **significantly greater** at Jowell Farm, providing evidence for a more standardized approach to local retouch at Millsap Cache.
-This finding provides additional support for the hypothesis that greater shape diversity is associated with retouch efforts, rather than production.
+This finding also provides support for the hypothesis that greater shape diversity is associated with retouch efforts, rather than production.
 
@@
 Components of this analytical work flow were developed and funded by a Preservation Technology and Training grant (P14AP00138) to RZS from the National Center for Preservation Technology and Training (NCPTT), and additional grants to RZS from the Caddo Nation of Oklahoma, National Forests and Grasslands in Texas (15-PA-11081300-033) and the United States Forest Service (20-PA-11081300-074).
-Funding to analyse the bifaces from Millsap Cache and Jowell Farm was made possible by support from the Heritage Research Center at Stephen F.
+Funding to analyse the bifaces from Millsap Cache and Jowell Farm was made possible by a Research Support Fund Grant from the Texas Archeological Society.
+Additional support for this project was provided by the Heritage Research Center at Stephen F.
 Austin State University.
-A Research Support Fund Grant from the Texas Archeological Society provided the tool needed to illustrate the bifaces.
 

### 2022-11-02 ms/ms/ms.tex

@@
 two of the largest samples of this ill-understood stone tool from the
-ancestral Caddo area. Elliptical bifaces from Millsap Cache were
-reportedly produced using Kay County flint, while those from Jowell Farm
-were manufactured using Edwards chert, providing for an empirical test
-of morphological differences as a function of raw material. The sample
-of elliptical bifaces was divided into two size classes; one
-conceptually reflective of production (large), and the other with local
-reduction practices (small). Size classes were used to assess whether
+ancestral Caddo area. Bifaces from Millsap Cache were reportedly
+produced using Kay County flint, while those from Jowell Farm were
+manufactured using Edwards chert, providing for an empirical test of
+morphological differences as a function of raw material. The sample of
+elliptical bifaces was divided into two size classes; one conceptually
+reflective of production (large), and the other with local reduction
+practices (small). Size classes were used to assess whether
 modifications by Caddo knappers may have yielded
-similar---convergent---biface shapes in the small class. Size classes
+similar---convergent---biface shape in the small class. Size classes
 were also used to test the hypothesis that greater morphological
@@
 responses related to local retouch practices. Results demonstrate that
-elliptical biface shape does not differ by raw material, but size does.
-This suggests that a shared, and consistent, mental template was
-maintained independent of biface size. The subsequent analysis of
+elliptical biface shape does not differ by raw material, but size does,
+suggesting that a shared and morphologically-consistent mental template
+was maintained independent of biface size. The subsequent analysis of
 elliptical biface morphology by size class demonstrated that size does
@@
 knife; Jowell knives; Jowell biface; lithics; oblique parallel flaking;
-archaeoinformatics; museum studies; digital humanities; non-western art
+archaeoinformatics; museum studies; digital humanities; non-Western art
 history; geometric morphometrics; STEM; STEAM
@@
 (\href{https://seldenlab.github.io/elliptical.bifaces/gm---siteraw-material.html}{Supplementary
-Materials, Ch. 2}). Prior to landmarking, elliptical bifaces were
-oriented with the most heavily retouched edge at top right (Figure
+Materials}). Prior to landmarking, elliptical bifaces were oriented with
+the most heavily retouched edge at top right (Figure
 \ref{fig:elliptical}). Some bifaces include multiple areas of heavy
@@
 access to the bifaces and associated records, to Sergio Ayala and the
-Gault School of Archaeological Research for access to the UV light, and
-to Scott Hammerstedt and Debra K. Green at the Oklahoma Archeological
-Survey for their assistance with records requests. Thanks also to John
-Harman for access to the DStretch plugin for ImageJ that was useful in
-the analysis of flake scars, and to Harry J. Shafer, Hiram F. (Pete)
-Gregory, Christian S. Hoggard, and David K. Thulman for their comments
-and constructive criticisms on the ongoing analyses of Caddo biface
-morphology, as well as Emma Sherratt, Kersten Bergstrom, Lauren Butaric,
-Dean C. Adams, and Michael L. Collyer for their constructive criticisms,
-general comments, and suggestions throughout the development of this
-research program.
+Gault School of Archaeological Research for access to the UV light, to
+Scott Hammerstedt and Debra K. Green at the Oklahoma Archeological
+Survey for their assistance with records requests, and to Marc Lavine
+and Susie Fishman-Armstrong at the Sam Noble Oklahoma Museum of Natural
+History for providing digital images from their collections. Thanks also
+to John Harman for access to the DStretch plugin for ImageJ, which was
+useful in the analysis of flake scars, and to Harry J. Shafer, Hiram F.
+(Pete) Gregory, Christian S. Hoggard, and David K. Thulman for their
+comments and constructive criticisms on the ongoing analyses of Caddo
+biface morphology, as well as Emma Sherratt, Kersten Bergstrom, Lauren
+Butaric, Dean C. Adams, and Michael L. Collyer for their constructive
+criticisms, general comments, and suggestions throughout the development
+of this research program.
 
@@
 States Forest Service (20-PA-11081300-074). Funding to analyse the
-bifaces from Millsap Cache and Jowell Farm was made possible by support
-from the Heritage Research Center at Stephen F. Austin State University.
+bifaces from Millsap Cache and Jowell Farm was provided by the Heritage
+Research Center at Stephen F. Austin State University.
 

### 2022-11-02 supp/03-2dgm-sizeclass.Rmd
<edit index>

@@
 
-Two `size classes` (large/small) were used to assess whether the morphology of elliptical bifaces from Millsap Cache and Jowell Farm remain morphologically stable through reduction.
-Variability in elliptical biface morphology through reduction provides evidence associated with differential resharpening trajectories.
+Two `size classes` (large/small) were used to assess whether the morphology of elliptical bifaces from Millsap Cache and Jowell Farm remains morphologically stable through reduction
 
@@
 Landmark data were aligned to a global coordinate system [@RN11622;@RN11623;@RN11563], achieved through generalized Procrustes superimposition [@RN478] performed in R 4.1.3 [@R] using the `geomorph` library v.
-4.0.3 [@RN11530;@RN1774;@RN9565].
+4.0.4 [@RN11530;@RN1774;@RN9565].
 Procrustes superimposition translates, scales, and rotates the coordinate data to allow for comparisons among objects [@RN11564;@RN478].

### 2022-11-02 ms/ms/ms.Rmd
<edit ms>

@@
 
-To assess the variable impacts that Caddo retouch practices may have had on general morphology, the sample of elliptical bifaces was first binned into two size classes using centroid size ([Supplementary Materials](https://seldenlab.github.io/elliptical.bifaces/)).
+To assess the variable impacts that Caddo retouch practices may have had on general morphology, the sample of elliptical bifaces was binned into two size classes using centroid size ([Supplementary Materials](https://seldenlab.github.io/elliptical.bifaces/)).
 Due to potential morphological differences associated with the use of two distinct raw materials (Kay County flint and Edwards chert), the sample was subset by site prior to calculating the mean centroid size for each.

### 2022-11-03 supp/index.Rmd
<edit index>

--- 2022-11-02 supp/index.Rmd
@@
 Principal components analysis [@RN1746] was used to visualise shape variation among the elliptical bifaces, and the scatterplot represents the dispersion of shapes in tangent space [@RN8633;@RN5616;@RN11143;@RN7550].
-Those shape changes described by each principal axis are commonly visualized using thin-plate spline warping of a reference image [@RN1731;@RN479].
+Shape ranges described by each principal axis are commonly visualized using thin-plate spline warping of a reference image [@RN1731;@RN479].
 
@@
 
-To assess whether `shape` changes with `size` (allometry), and whether shape and size differed by `raw material` and `size class`, Procrustes ANOVAs [@RN1749] were run that enlist effect-sizes (zscores) computed as standard deviates of the generated sampling distributions [@RN1756].
+To assess whether `shape` differs by `size` (allometry), and whether shape and size differed by `raw material` and `size class`, Procrustes ANOVAs [@RN1749] were run that enlist effect-sizes (zscores) computed as standard deviates of the generated sampling distributions [@RN1756].
 A residual randomization permutation procedure (RRPP; n = 10,000 permutations) was used for all Procrustes ANOVAs [@RN1655;@RN11775], which has higher statistical power and a greater ability to identify patterns in the data should they be present [@RN1719].
@@
 
-The shape of elliptical bifaces does not differ significantly in either size class; however, size does.
-Results of the Procrustes ANOVAs support the hypothesis for divergent morphology.
-The analysis of morphological disparity yielded support for the interpretation that greater morphological diversity is associated with local retouch, while providing additional support for morphological consistency related to production.
+The pattern that emerges is one of differential retouch intensity, where the isometric reduction strategies of one group (Millsap Cache) may correlate with differing retouch practices that articulate with locally abundant raw material (Kay County flint).
+This stands in contrast to the comparatively minimal retouch intensity of another group (Jowell Farm) where local raw material was of an insufficient size, yielding a situation where it was necessary to import raw material from an extralocal source (Edwards chert) in central Texas.
+
+Further, the comparison of morphological diversity by size class demonstrated that elliptical biface shapes in the small size class at Jowell Farm occupy a significantly greater range of morphospace than those from Millsap Cache.
+This pattern suggests that the comparatively minimal retouch intensity associated with Caddo knappers at Jowell Farm may have been more targeted---potentially due to a lack of abundant local raw material of sufficient size---resulting in a broader range of biface shapes in the small size class.
+Conversely, these results also suggest that Caddo knappers at Millsap Cache employed a more standardized (specialized?) approach to reduction.
 

### 2022-11-03 ms/ms/ms.tex

@@
 To assess the variable impacts that Caddo retouch practices may have had
-on general morphology, the sample of elliptical bifaces was first binned
-into two size classes using centroid size
+on general morphology, the sample of elliptical bifaces was binned into
+two size classes using centroid size
 (\href{https://seldenlab.github.io/elliptical.bifaces/}{Supplementary

### 2022-11-03 supp/index.Rmd
<edit index>

@@
 
-The pattern that emerges is one of differential retouch intensity, where the isometric reduction strategies of one group (Millsap Cache) may correlate with differing retouch practices that articulate with locally abundant raw material (Kay County flint).
-This stands in contrast to the comparatively minimal retouch intensity of another group (Jowell Farm) where local raw material was of an insufficient size, yielding a situation where it was necessary to import raw material from an extralocal source (Edwards chert) in central Texas.
+The pattern that emerges is one of differential retouch _intensity_, where the isometric reduction strategies of one group (Millsap Cache) may correlate with differing retouch practices that articulate with locally abundant raw material (Kay County flint).
+This stands in contrast to the comparatively minimal retouch intensity of another group (Jowell Farm) where local raw material was of an insufficient size, yielding a situation where it was necessary to import raw material from an extralocal source in central Texas (Edwards chert).
 
 Further, the comparison of morphological diversity by size class demonstrated that elliptical biface shapes in the small size class at Jowell Farm occupy a significantly greater range of morphospace than those from Millsap Cache.
-This pattern suggests that the comparatively minimal retouch intensity associated with Caddo knappers at Jowell Farm may have been more targeted---potentially due to a lack of abundant local raw material of sufficient size---resulting in a broader range of biface shapes in the small size class.
+This pattern suggests that the comparatively minimal retouch intensity associated with Caddo knappers at Jowell Farm may have been more _targeted_---potentially due to a lack of abundant local raw material of sufficient size---resulting in a broader range of biface shapes in the small size class.
 Conversely, these results also suggest that Caddo knappers at Millsap Cache employed a more standardized (specialized?) approach to reduction.

### 2022-11-03 supp/index.Rmd
<edit index>

@@
 ```{r figmap, fig.cap="Location of Millsap Cache, Jowell Farm, and other Caddo sites mentioned in the text.
-Elliptical bifaces have been found in every major river basin in the ancestral Caddo area (dashed/white)."}
+Elliptical bifaces have been found in all but one major drainage basin in the ancestral Caddo area (dashed/white)."}
 
@@
 
-The pattern that emerges is one of differential retouch _intensity_, where the isometric reduction strategies of one group (Millsap Cache) may correlate with differing retouch practices that articulate with locally abundant raw material (Kay County flint).
-This stands in contrast to the comparatively minimal retouch intensity of another group (Jowell Farm) where local raw material was of an insufficient size, yielding a situation where it was necessary to import raw material from an extralocal source in central Texas (Edwards chert).
+The pattern that emerges is one of differential retouch _intensity_, where the isometric reduction strategies of one group (Millsap Cache) articulates with a different reduction practice thought to be associated with locally abundant raw material (Kay County flint).
+This stands in contrast to the comparatively minimal retouch intensity of another group (Jowell Farm) where local raw material was of poor quality and insufficient size, yielding a situation where it was necessary to import raw material from an extralocal source (Edwards chert).
 

### 2022-11-03 supp/index.Rmd
<edit index>

@@
 
-American Southeast, Caddo, NAGPRA, archaeoinformatics, museum studies, digital humanities, non-Western art history, geometric morphometrics, STEM, STEAM, ovoid biface, Jowell knife, Jowell knives, lithics, oblique parallel flaking
+American Southeast, Caddo, NAGPRA, archaeoinformatics, museum studies, digital humanities, non-Western art history, geometric morphometrics, STEM, STEAM, ovoid biface, Jowell knife, Jowell knives, Jowell biface, lithics, oblique parallel flaking
 
@@
 
-The pattern that emerges is one of differential retouch _intensity_, where the isometric reduction strategies of one group (Millsap Cache) articulates with a different reduction practice thought to be associated with locally abundant raw material (Kay County flint).
-This stands in contrast to the comparatively minimal retouch intensity of another group (Jowell Farm) where local raw material was of poor quality and insufficient size, yielding a situation where it was necessary to import raw material from an extralocal source (Edwards chert).
+The pattern that emerges is one of differential isometric retouch _intensity_, where the retouch strategies of Caddo knappers from Millsap Cache articulate with a different reduction practice thought to be associated with locally abundant raw material (Kay County flint).
+This stands in contrast to the comparatively minimal isometric retouch intensity of another group (Jowell Farm) where local raw material was of poor quality and insufficient size, yielding a situation where it was necessary to import raw material from an extralocal source (Edwards chert).
 

### 2022-11-03 supp/index.Rmd
<edit index>

@@
 
-The pattern that emerges is one of differential isometric retouch _intensity_, where the retouch strategies of Caddo knappers from Millsap Cache articulate with a different reduction practice thought to be associated with locally abundant raw material (Kay County flint).
-This stands in contrast to the comparatively minimal isometric retouch intensity of another group (Jowell Farm) where local raw material was of poor quality and insufficient size, yielding a situation where it was necessary to import raw material from an extralocal source (Edwards chert).
+The pattern that emerges is one of differential retouch _intensity_, where the isometric retouch strategies of Caddo knappers from Millsap Cache articulate with a different reduction practice thought to be associated with locally abundant raw material (Kay County flint).
+This stands in contrast to the comparatively minimal retouch intensity of another group (Jowell Farm) where local raw materials were of poor quality and insufficient size, yielding a situation where it was necessary to import raw material from an extralocal source (Edwards chert).
 
 Further, the comparison of morphological diversity by size class demonstrated that elliptical biface shapes in the small size class at Jowell Farm occupy a significantly greater range of morphospace than those from Millsap Cache.
-This pattern suggests that the comparatively minimal retouch intensity associated with Caddo knappers at Jowell Farm may have been more _targeted_---potentially due to a lack of abundant local raw material of sufficient size---resulting in a broader range of biface shapes in the small size class.
+This pattern suggests that the comparatively minimal retouch intensity associated with Caddo knappers at Jowell Farm may have been more _targeted_---potentially due to a lack of abundant local raw material of sufficient quality and size---resulting in a broader range of biface shapes in the small size class.
 Conversely, these results also suggest that Caddo knappers at Millsap Cache employed a more standardized (specialized?) approach to reduction.

### 2022-11-07 supp/02-2dgm-rawmat.Rmd
<edit index>

@@
 
-Elliptical bifaces from the Millsap Cache and Jowell Farm were used to examine whether their morphology remains stable, or expresses morphological variability by site/raw meterial.
+Elliptical bifaces from the Millsap Cache and Jowell Farm were used to examine whether their morphology remains stable, or expresses morphological variability by site/raw material.
 Variability in elliptical biface morphology provides evidence associated with differences in morphology as a function of geography/raw material.
@@
 
-# allometry
+# shape change with size?
 anova(fit.size)

### 2022-11-07 supp/03-2dgm-sizeclass.Rmd
<edit index>

@@
 
-Two `size classes` (large/small) were used to assess whether the morphology of elliptical bifaces from Millsap Cache and Jowell Farm remains morphologically stable through reduction
+Two `size classes` (large/small) were used to assess whether the morphology of elliptical bifaces from Millsap Cache and Jowell Farm remains morphologically stable/consistent through reduction.
 

### 2022-11-10 ms/ms/ms.Rmd
<edit ms>

@@
 
-Landmark data were aligned to a global coordinate system \citep{RN8477,RN7502,RN11622,RN11623,RN11563}, achieved through generalised Procrustes superimposition \citep{RN11138,RN478,RN1646} in R using the `geomorph` and `RRPP` packages \citep{RN1655,RN11775,RN11530,RN1774,RN9565}.
+Landmark data were aligned to a global coordinate system \citep{RN8477,RN7502,RN11622,RN11623,RN11563}, achieved through generalised Procrustes superimposition \citep{RN11138,RN478,RN1646} in R using the `geomorph` and `RRPP` packages \citep{RN1655,RN11775,RN11530,RN1774,RN8605}.
 Procrustes superimposition translates and rotates the coordinate data to allow for comparisons among objects, while also scaling each biface using unit-centroid size---the square root of the sum of squared distances from each landmark to the specimen’s centroid \citep{RN11139,RN11140,RN11564,RN478}.
@@
 Principal components analysis \citep{RN1746} was used to visualise shape variation among the elliptical bifaces, and the scatterplot and convex hulls represent the dispersion of shapes in tangent space (Figure \ref{fig:pca}) \citep{RN8633,RN5616,RN11143,RN7550}.
-The shape space described by each principal axis is commonly visualized using thin-plate spline warping of a reference image \citep{RN1731,RN479}.
+Shape space described by each principal axis is commonly visualized using thin-plate spline warping of a reference image \citep{RN1731,RN479}.
 
@@
 Pairwise comparisons demonstrate that elliptical bifaces from Millsap Cache and Jowell Farm in the large size class are morphologically consistent, but those in the small size class differ in size (RRPP = 10,000, Pr>d = 0.0001) ([Supplementary Materials](https://seldenlab.github.io/elliptical.bifaces/)).
-Thus, the pattern that emerges is one of differential retouch _intensity_, where isometric reduction strategies of one group (Millsap Cache) may correlate with differing retouch practices that articulate with locally abundant raw material (Kay County flint).
-This stands in contrast to the comparatively minimal retouch intensity of another group (Jowell Farm) where local raw material size is insufficient, making it necessary to import the requisite raw material from an extralocal source; in this case, Edwards chert from central Texas.
+
+Thus, the pattern that emerges is one of differential retouch _intensity_, where the isometric retouch strategies of Caddo knappers from Millsap Cache articulate with a different reduction practice thought to be associated with locally abundant raw material (Kay County flint).
+This stands in contrast to the comparatively minimal retouch intensity of another group (Jowell Farm) where local raw materials were of poor quality and insufficient size, yielding a situation where it was necessary to import raw material from an extralocal source (Edwards chert).
 
@@
 No difference was found for elliptical biface shape or size by site/raw material, but the comparison of morphological diversity by size class demonstrated that biface shapes in the small size class from Jowell Farm occupy a significantly greater range of morphospace than those from Millsap Cache.
-This pattern suggests that the comparatively minimal retouch intensity associated with Caddo knappers at Jowell Farm may have been more targeted---potentially due to a lack of abundant local raw material---resulting in a broader range of biface shapes in the small size class.
-Conversely, these results also suggest that Caddo knappers at Millsap Cache employed a more standardised (specialised?) approach to reduction practices.
+
+The comparison of morphological disparity by size class demonstrated that elliptical biface shapes in the small size class at Jowell Farm occupy a significantly greater range of morphospace than those from Millsap Cache.
+This pattern suggests that the comparatively minimal retouch intensity associated with Caddo knappers at Jowell Farm may have been more targeted---potentially due to a lack of abundant local raw material of sufficient quality and size---resulting in a broader range of biface shapes in the small size class.
+Conversely, these results also suggest that Caddo knappers at Millsap Cache employed a more standardized (specialized?) approach to reduction.
 
 # Discussion and conclusion
+
+\begin{figure}\centering
+\includegraphics[width=\linewidth]{figs/elliptical.illustration.png}
+\caption{Illustrated elliptical bifaces from a, an unknown location; b, Jowell Farm; and c, Belcher Mound highlighting oblique parallel flaking used in production.
+The specimen from Belcher Mound is a potential preform exhibiting oblique parallel removals consistent with elliptical bifaces from other sites in the region.}
+\label{fig:illustration}
+\end{figure}
 

### 2022-11-10 ms/ms/ms.tex
<edit ms>

@@
 using the \texttt{geomorph} and \texttt{RRPP} packages
-\citep{RN1655,RN11775,RN11530,RN1774,RN9565}. Procrustes superimposition
+\citep{RN1655,RN11775,RN11530,RN1774,RN8605}. Procrustes superimposition
 translates and rotates the coordinate data to allow for comparisons
@@
 hulls represent the dispersion of shapes in tangent space (Figure
-\ref{fig:pca}) \citep{RN8633,RN5616,RN11143,RN7550}. The shape space
+\ref{fig:pca}) \citep{RN8633,RN5616,RN11143,RN7550}. Shape space
 described by each principal axis is commonly visualized using thin-plate
@@
 (\href{https://seldenlab.github.io/elliptical.bifaces/}{Supplementary
-Materials}). Thus, the pattern that emerges is one of differential
-retouch \emph{intensity}, where isometric reduction strategies of one
-group (Millsap Cache) may correlate with differing retouch practices
-that articulate with locally abundant raw material (Kay County flint).
-This stands in contrast to the comparatively minimal retouch intensity
-of another group (Jowell Farm) where local raw material size is
-insufficient, making it necessary to import the requisite raw material
-from an extralocal source; in this case, Edwards chert from central
-Texas.
+Materials}).
+
+Thus, the pattern that emerges is one of differential retouch
+\emph{intensity}, where the isometric retouch strategies of Caddo
+knappers from Millsap Cache articulate with a different reduction
+practice thought to be associated with locally abundant raw material
+(Kay County flint). This stands in contrast to the comparatively minimal
+retouch intensity of another group (Jowell Farm) where local raw
+materials were of poor quality and insufficient size, yielding a
+situation where it was necessary to import raw material from an
+extralocal source (Edwards chert).
 
@@
 the small size class from Jowell Farm occupy a significantly greater
-range of morphospace than those from Millsap Cache. This pattern
-suggests that the comparatively minimal retouch intensity associated
-with Caddo knappers at Jowell Farm may have been more
-targeted---potentially due to a lack of abundant local raw
-material---resulting in a broader range of biface shapes in the small
-size class. Conversely, these results also suggest that Caddo knappers
-at Millsap Cache employed a more standardised (specialised?) approach to
-reduction practices.
+range of morphospace than those from Millsap Cache.
+
+The comparison of morphological disparity by size class demonstrated
+that elliptical biface shapes in the small size class at Jowell Farm
+occupy a significantly greater range of morphospace than those from
+Millsap Cache. This pattern suggests that the comparatively minimal
+retouch intensity associated with Caddo knappers at Jowell Farm may have
+been more targeted---potentially due to a lack of abundant local raw
+material of sufficient quality and size---resulting in a broader range
+of biface shapes in the small size class. Conversely, these results also
+suggest that Caddo knappers at Millsap Cache employed a more
+standardized (specialized?) approach to reduction.
 
@@
 \section{Discussion and conclusion}\label{discussion-and-conclusion}}
+
+\begin{figure}\centering
+\includegraphics[width=\linewidth]{figs/elliptical.illustration.png}
+\caption{Illustrated elliptical bifaces from a, an unknown location; b, Jowell Farm; and c, Belcher Mound highlighting oblique parallel flaking used in production.
+The specimen from Belcher Mound is a potential preform exhibiting oblique parallel removals consistent with elliptical bifaces from other sites in the region.}
+\label{fig:illustration}
+\end{figure}
 

### 2022-11-10 ms/ms/ms.Rmd
<edit ms>

@@
 The analysis code associated with this project can be accessed through the [Supplementary Materials](https://seldenlab.github.io/elliptical.bifaces/), is available through the GitHub repository [https://github.com/seldenlab/elliptical.bifaces](https://github.com/seldenlab/elliptical.bifaces), and digitally curated on the Open Science Framework [DOI 10.17605/OSF.IO/PH25W](https://osf.io/ph25w/).
-The reproducible nature of this enterprise provides a means for others to critically assess and evaluate the various analytical components (Gray and Marwick 2019; Peng 2011; Gandrud 2014), which is a necessary requirement for the production of reliable knowledge.
+The reproducible nature of this enterprise provides a means for others to critically assess and evaluate the various analytical components \citep{RN7434,RN7435,RN7427}, which is a necessary requirement for the production of reliable knowledge.
 
 Reproducibility projects in psychology and cancer biology are impacting current research practices across all domains.
-Examples of reproducible research are becoming more abundant in archaeology (Marwick 2016; Ivanovaite et al.
-2020; Selden Jr., Dockall, and Shafer 2018; Selden Jr., Dockall, and Dubied 2020; Selden Jr et al.
-2021a; Selden 2022), and the next generation of archaeologists are learning those tools and methods needed to reproduce and/or replicate research results (Marwick et al.
-2019).
-Reproducible and replicable research work flows are often employed at the highest levels of humanities-based inquiries to mitigate concern or doubt regarding proper execution, and is of particular import should the results have—explicitly or implicitly—a major impact on scientific progress (Peels and Bouter 2018).
+Examples of reproducible research are becoming more abundant in archaeology \citep{RN9022,RN7818,RN10578,RN10576,RN8510,RN11097}, and the next generation of archaeologists are learning those tools and methods needed to reproduce and/or replicate research results \citep{RN10579}.
+Reproducible and replicable research work flows are often employed at the highest levels of humanities-based inquiries to mitigate concern or doubt regarding proper execution, and is of particular import should the results have—explicitly or implicitly—a major impact on scientific progress \citep{RN10580}.

### 2022-11-10 ms/ms/ms.tex
<edit ms>

@@
 provides a means for others to critically assess and evaluate the
-various analytical components (Gray and Marwick 2019; Peng 2011; Gandrud
-2014), which is a necessary requirement for the production of reliable
-knowledge.
+various analytical components \citep{RN7434,RN7435,RN7427}, which is a
+necessary requirement for the production of reliable knowledge.
 
@@
 current research practices across all domains. Examples of reproducible
-research are becoming more abundant in archaeology (Marwick 2016;
-Ivanovaite et al.~2020; Selden Jr., Dockall, and Shafer 2018; Selden
-Jr., Dockall, and Dubied 2020; Selden Jr et al.~2021a; Selden 2022), and
-the next generation of archaeologists are learning those tools and
-methods needed to reproduce and/or replicate research results (Marwick
-et al.~2019). Reproducible and replicable research work flows are often
-employed at the highest levels of humanities-based inquiries to mitigate
-concern or doubt regarding proper execution, and is of particular import
-should the results have---explicitly or implicitly---a major impact on
-scientific progress (Peels and Bouter 2018).
+research are becoming more abundant in archaeology
+\citep{RN9022,RN7818,RN10578,RN10576,RN8510,RN11097}, and the next
+generation of archaeologists are learning those tools and methods needed
+to reproduce and/or replicate research results \citep{RN10579}.
+Reproducible and replicable research work flows are often employed at
+the highest levels of humanities-based inquiries to mitigate concern or
+doubt regarding proper execution, and is of particular import should the
+results have---explicitly or implicitly---a major impact on scientific
+progress \citep{RN10580}.
 

### 2022-11-11 ms/ms.Rmd
<edit ms>

--- 2022-11-10 ms/ms/ms.Rmd
@@
 \includegraphics[width=\linewidth]{figs/ellipticalbifaces.png}
-\caption{Selected elliptical bifaces from the Jowell Farm site highlighting the most heavily retouched area at top right identified using a modified version of the flaking index developed and advanced by \citet{RN11099}, \citet{RN9242}, and \citet{RN11098}.}
+\caption{Selected elliptical bifaces from the Jowell Farm site highlighting the most heavily retouched area at top right identified using a modified version of the flaking index developed by \citet{RN11099}.}
 \label{fig:elliptical}
@@
 
-To identify which edge was most heavily retouched, we employed a modified approach to the flaking index initially developed by \citet{RN11099}, and later advanced and refined by \citet{RN9242}, then \citet{RN11098}.
+To identify which edge was most heavily retouched, we employed a modified approach to the flaking index developed by \citet{RN11099}.
 The modified approach uses counts of flake scars from each edge in the two most heavily worked areas, paired with a measure of edge length inclusive of curvature, which was calculated using ImageJ \citep{RN11146,RN11147,RN11148}.
-The number of flake scars was then divided by the length of the worked edge, with the most heavily worked edge identified by the greater value.
+The number of flake scars was subsequently divided by the length of the worked edge, with the most heavily worked edge identified by the greater value.
 
@@
 
-The same landmark/semilandmark configuration was used for both analyses, and the landmarking protocol employs three landmarks; two horizontal tangents (top/bottom), and the third placed at the furthest extent of the retouched edge bearing the heaviest amount of retouch (see Figure \@ref(fig:elliptical)).
+The same landmark/semilandmark configuration was used for both analyses, and the landmarking protocol employs three landmarks; two horizontal tangents (top/bottom), and the third placed at the furthest extent of the retouched edge bearing the heaviest amount of retouch (see Figure \ref{fig:elliptical}).
 Equidistant semilandmarks were placed between each landmark, and all landmarks and semilandmarks were applied using R \citep{R} and the `StereoMorph` package \citep{RN9091}.
@@
 
-# Discussion and conclusion
+# Discussion
 
@@
 \end{figure}
+
+# Conclusion
 

### 2022-11-11 ms/ms.tex
<edit ms>

--- 2022-11-10 ms/ms/ms.tex
@@
 To identify which edge was most heavily retouched, we employed a
-modified approach to the flaking index initially developed by
-\citet{RN11099}, and later advanced and refined by \citet{RN9242}, then
-\citet{RN11098}. The modified approach uses counts of flake scars from
-each edge in the two most heavily worked areas, paired with a measure of
-edge length inclusive of curvature, which was calculated using ImageJ
-\citep{RN11146,RN11147,RN11148}. The number of flake scars was then
-divided by the length of the worked edge, with the most heavily worked
-edge identified by the greater value.
+modified approach to the flaking index developed by \citet{RN11099}. The
+modified approach uses counts of flake scars from each edge in the two
+most heavily worked areas, paired with a measure of edge length
+inclusive of curvature, which was calculated using ImageJ
+\citep{RN11146,RN11147,RN11148}. The number of flake scars was
+subsequently divided by the length of the worked edge, with the most
+heavily worked edge identified by the greater value.
 
@@
 the retouched edge bearing the heaviest amount of retouch (see Figure
-@ref(fig:elliptical)). Equidistant semilandmarks were placed between
+\ref{fig:elliptical}). Equidistant semilandmarks were placed between
 each landmark, and all landmarks and semilandmarks were applied using R
@@
 
-\hypertarget{discussion-and-conclusion}{%
-\section{Discussion and conclusion}\label{discussion-and-conclusion}}
+\hypertarget{discussion}{%
+\section{Discussion}\label{discussion}}
 
@@
 \end{figure}
+
+\hypertarget{conclusion}{%
+\section{Conclusion}\label{conclusion}}
 

### 2022-11-19 supp/index.Rmd
<edit index>

@@
 
-```{r figlm, fig.cap="Selected elliptical bifaces from the Jowell Farm site highlighting the most heavily retouched area identified using a modified version of the flaking index developed and advanced by Miller [-@RN11459], Smallwood [-@RN8310], and Miller and Smallwood [-@RN11460]."}
+```{r figlm, fig.cap="Selected elliptical bifaces from the Jowell Farm site highlighting the most heavily retouched area identified using a modified version of the flaking index developed and advanced by Miller [-@RN11459]."}
 
@@
 
-To identify which edge was most heavily retouched, we employed a modified approach to the flaking index initially developed by Miller [-@RN11459], and later advanced and refined by Smallwood [-@RN8310], then Miller and Smallwood [-@RN11460].
+To identify which edge was most heavily retouched, we employed a modified approach to the flaking index initially developed by Miller [-@RN11459].
 The modified approach uses counts of flake scars from each edge in the two most heavily worked areas, paired with a measure of edge length (inclusive of curvature - this is not a linear metric).
@@
 The same landmark/semilandmark configuration was used for both analyses, and the landmarking protocol employs three landmarks; two horizontal tangents (top/bottom), and the third placed at the furthest extent of the retouched edge bearing the heaviest amount of retouch (see Figure \@ref(fig:figlm)).
-Equidistant semilandmarks were placed between each landmark, and all landmarks and semilandmarks were applied using R [@R] and the `StereoMorph` package [@RN9091].
 
@@
 
-## Acknowledgments
+## Acknowledgments {-}
 
@@
 
-## Funding
+## Funding {-}
 
@@
 
-## Data management
+## Data management {-}
 
@@
 
-## Colophon
+## Colophon {-}
 

### 2022-11-19 ms/ms.Rmd
<edit index>

@@
 # Introduction
+
+Elliptical bifaces have a deep history in eastern North America extending from the Middle Archaic to well past European contact.
+In many instances they have been deposited in caches, which may represent large, corporate bundles decommissioned amid polity demise or radical transformations in social practice.
+Stylistic changes are evident in both length and width for these bifaces, and result in great variability in shape and size.
+While some undoubtedly were knapped for every-day, mundane uses, others were crafted from exotic cherts and reflect non-utilitarian uses.
+The _oversize_ or _exaggerated_ forms, mimicking combat weapons, are often referred to as symbolic weaponry.
+Such objects are typically those “whose size, delicacy, material of manufacture, and craftsmanship are outside the bounds of any weapon that might have been actually used” \citep[624]{RN11177}.
+
+The earliest elliptical bifaces, associated with Middle Archaic period, recovered as Benton caches (4500—4000 BCE) in the Mid-South \citep{RN11179}, are clearly non-utilitarian, as they “meet the expectations of sacred markers for secular exchange” \citep[143]{RN11178}.
+In the western Middle Tennessee Valley, the Pickwick Burial Complex \citep[94]{RN11180}, dating to the Late Archaic/Gulf Formational period (3000—300 BCE) exhibits a continuation of bipointed blades, some notched or truncated, found in ritual caches and ceremonially broken.
+Elliptical bifaces find their most dramatic expression as Caddoan and Mississippian exaggerated, symbolic weaponry.
+Mississippian Ramey knives, dating from c.
+CE 1000 to 1300 are found associated with Cahokia political dynamics.
+The slightly later Duck River sword-form bifaces (CE 1275-1400) continue the emphasis on high-quality knapping and long-distance circulation.
+Such lanceolate blades, although smaller, continue into the early seventeenth century when they serve as regalia among political and social elites in ritual performances.
 

### 2022-11-19 ms/ms.tex
<edit index>

@@
 \section{Introduction}\label{introduction}}
+
+Elliptical bifaces have a deep history in eastern North America
+extending from the Middle Archaic to well past European contact. In many
+instances they have been deposited in caches, which may represent large,
+corporate bundles decommissioned amid polity demise or radical
+transformations in social practice. Stylistic changes are evident in
+both length and width for these bifaces, and result in great variability
+in shape and size. While some undoubtedly were knapped for every-day,
+mundane uses, others were crafted from exotic cherts and reflect
+non-utilitarian uses. The \emph{oversize} or \emph{exaggerated} forms,
+mimicking combat weapons, are often referred to as symbolic weaponry.
+Such objects are typically those ``whose size, delicacy, material of
+manufacture, and craftsmanship are outside the bounds of any weapon that
+might have been actually used'' \citep[624]{RN11177}.
+
+The earliest elliptical bifaces, associated with Middle Archaic period,
+recovered as Benton caches (4500---4000 BCE) in the Mid-South
+\citep{RN11179}, are clearly non-utilitarian, as they ``meet the
+expectations of sacred markers for secular exchange''
+\citep[143]{RN11178}. In the western Middle Tennessee Valley, the
+Pickwick Burial Complex \citep[94]{RN11180}, dating to the Late
+Archaic/Gulf Formational period (3000---300 BCE) exhibits a continuation
+of bipointed blades, some notched or truncated, found in ritual caches
+and ceremonially broken. Elliptical bifaces find their most dramatic
+expression as Caddoan and Mississippian exaggerated, symbolic weaponry.
+Mississippian Ramey knives, dating from c.~CE 1000 to 1300 are found
+associated with Cahokia political dynamics. The slightly later Duck
+River sword-form bifaces (CE 1275-1400) continue the emphasis on
+high-quality knapping and long-distance circulation. Such lanceolate
+blades, although smaller, continue into the early seventeenth century
+when they serve as regalia among political and social elites in ritual
+performances.
 

### 2022-11-20 ms/ms.Rmd
<edit ms>

@@
 Raw material for bifaces from Millsap Cache were identified by \citet{RN11461} as Kay County flint, and those from Jowell Farm were identified as Edwards Chert due to their size, the colours that they fluoresce beneath low and high frequency UV light, and the paucity of similarly large locally-available raw material sources.
-Bifaces larger than the mean were then assigned to the large category, those smaller than the mean were assigned to the small category, and the two datasets were then joined in advance of analysis ([Supplementary Materials](https://seldenlab.github.io/elliptical.bifaces/)).
+Bifaces larger than the mean were assigned to the large category, those smaller than the mean were assigned to the small category, and the two datasets were joined in advance of analysis ([Supplementary Materials](https://seldenlab.github.io/elliptical.bifaces/)).
 
-Elliptical bifaces from the Millsap Cache and Jowell Farm were used to examine whether biface morphology remains stable or expresses morphological variability by site/raw material ([Supplementary Materials](https://seldenlab.github.io/elliptical.bifaces/gm---siteraw-material.html)).
+Elliptical bifaces from Millsap Cache and Jowell Farm were used to examine whether biface morphology remains stable or expresses morphological variability by site/raw material ([Supplementary Materials](https://seldenlab.github.io/elliptical.bifaces/gm---siteraw-material.html)).
 Prior to landmarking, elliptical bifaces were oriented with the most heavily retouched edge at top right (Figure  \ref{fig:elliptical}).
@@
 To identify which edge was most heavily retouched, we employed a modified approach to the flaking index developed by \citet{RN11099}.
-The modified approach uses counts of flake scars from each edge in the two most heavily worked areas, paired with a measure of edge length inclusive of curvature, which was calculated using ImageJ \citep{RN11146,RN11147,RN11148}.
+The modified approach uses counts of flake scars from both sides of each edge in the two most heavily worked areas, paired with a measure of edge length inclusive of curvature, which was calculated using ImageJ \citep{RN11146,RN11147,RN11148}.
 The number of flake scars was subsequently divided by the length of the worked edge, with the most heavily worked edge identified by the greater value.
@@
 The same landmark/semilandmark configuration was used for both analyses, and the landmarking protocol employs three landmarks; two horizontal tangents (top/bottom), and the third placed at the furthest extent of the retouched edge bearing the heaviest amount of retouch (see Figure \ref{fig:elliptical}).
-Equidistant semilandmarks were placed between each landmark, and all landmarks and semilandmarks were applied using R \citep{R} and the `StereoMorph` package \citep{RN9091}.
+Equidistant semilandmarks were placed between each landmark, applied using R \citep{R} and the `StereoMorph` package \citep{RN9091}.
 
@@
 Principal components analysis \citep{RN1746} was used to visualise shape variation among the elliptical bifaces, and the scatterplot and convex hulls represent the dispersion of shapes in tangent space (Figure \ref{fig:pca}) \citep{RN8633,RN5616,RN11143,RN7550}.
-Shape space described by each principal axis is commonly visualized using thin-plate spline warping of a reference image \citep{RN1731,RN479}.
+Shape space described by each principal axis is commonly visualized using thin-plate spline warping of a reference image or 3D mesh \citep{RN1731,RN479}.
 
@@
 
-To assess whether shape changes with size (allometry), and whether shape and size differed by raw material and size class, Procrustes ANOVAs \citep{RN1749} were run that enlist effect-sizes (zscores) computed as standard deviates of the generated sampling distributions \citep{RN1756}.
+To assess whether shape changed with size (allometry), and whether shape and size differed by raw material and size class, Procrustes ANOVAs \citep{RN1749} were run that enlist effect-sizes (zscores) computed as standard deviates of the generated sampling distributions \citep{RN1756}.
 A residual randomization permutation procedure (RRPP; n = 10,000 permutations) was used for all Procrustes ANOVAs \citep{RN1655,RN11775}, which has higher statistical power and a greater ability to identify patterns in the data should they be present \citep{RN1719}.
 
-The test for allometry in the sample did not return a significant result, suggesting that elliptical biface shape and size may follow a pattern of isometry, where shape remains constant through the reduction process ([Supplementary Materials](https://seldenlab.github.io/elliptical.bifaces/)).
-Elliptical bifaces were also found to express a consistency of shape across all analyses ([Supplementary Materials](https://seldenlab.github.io/elliptical.bifaces/)).
-Results demonstrated a significant difference in size by site/raw material (RRPP = 10,000, Pr(>F) = 0.0007); however, the subsequent analysis of elliptical bifaces by size class demonstrates that the pattern is more readily attributable to reduction practices/intensity rather than nodule and raw material size ([Supplementary Materials](https://seldenlab.github.io/elliptical.bifaces/)).
+The test for allometry did not return a significant result, suggesting that elliptical biface shape and size may follow a pattern of isometry, where shape remains constant throughout the reduction/retouch process ([Supplementary Materials](https://seldenlab.github.io/elliptical.bifaces/)).
+Elliptical bifaces were morphologically consistent in shape across all analyses ([Supplementary Materials](https://seldenlab.github.io/elliptical.bifaces/)).
+Results demonstrated a significant difference in size by site/raw material (RRPP = 10,000, Pr(>F) = 0.0007); however, the subsequent analysis by size class demonstrated that the pattern is more readily attributable to reduction practices/intensity rather than nodule and raw material size ([Supplementary Materials](https://seldenlab.github.io/elliptical.bifaces/)).
 
-The analysis of morphology by size class demonstrates a significant difference in size between size classes; however, that pattern deviated from our expectations.
-Elliptical bifaces from Jowell Farm differ in neither shape nor size between the size classes ([Supplementary Materials](https://seldenlab.github.io/elliptical.bifaces/)).
+The analysis of morphology by size class demonstrated a significant difference in size between size classes (expected); however, the pattern deviated from our expectations.
+Elliptical bifaces from Jowell Farm differ in neither shape nor size between size classes ([Supplementary Materials](https://seldenlab.github.io/elliptical.bifaces/)).
 Elliptical bifaces in the large and small size classes from Millsap Cache differ significantly in size (RRPP = 10,000, Pr>d = 0.0035), but not shape ([Supplementary Materials](https://seldenlab.github.io/elliptical.bifaces/)).
@@
 
-Thus, the pattern that emerges is one of differential retouch _intensity_, where the isometric retouch strategies of Caddo knappers from Millsap Cache articulate with a different reduction practice thought to be associated with locally abundant raw material (Kay County flint).
-This stands in contrast to the comparatively minimal retouch intensity of another group (Jowell Farm) where local raw materials were of poor quality and insufficient size, yielding a situation where it was necessary to import raw material from an extralocal source (Edwards chert).
+Thus, the pattern that emerges is one of differential retouch _intensity_, where isometric retouch strategies of Caddo knappers at Millsap Cache articulate with a different reduction practice thought to be associated with locally abundant raw material (Kay County flint).
+This finding is contrasted with the comparatively minimal retouch intensity of another group (Jowell Farm) where local raw materials were of poor quality and insufficient size, yielding a situation where it was necessary to import raw material from an extralocal source (Edwards chert).
 
@@
 An analysis of morphological disparity was used to identify potential differences in morphological diversity between size classes \citep{RN11107,RN7041,RN5694}.
-No difference was found for elliptical biface shape or size by site/raw material, but the comparison of morphological diversity by size class demonstrated that biface shapes in the small size class from Jowell Farm occupy a significantly greater range of morphospace than those from Millsap Cache.
+No difference was found in elliptical biface shape or size by site/raw material, but the comparison of morphological diversity by size class demonstrated that biface shapes in the small size class from Jowell Farm occupy a significantly greater range of morphospace than those from Millsap Cache.
 
@@
 This pattern suggests that the comparatively minimal retouch intensity associated with Caddo knappers at Jowell Farm may have been more targeted---potentially due to a lack of abundant local raw material of sufficient quality and size---resulting in a broader range of biface shapes in the small size class.
-Conversely, these results also suggest that Caddo knappers at Millsap Cache employed a more standardized (specialized?) approach to reduction.
+Conversely, results also suggest that Caddo knappers at Millsap Cache employed a more standardized (specialized?) approach to reduction.
 

### 2022-11-20 ms/ms.tex
<edit ms>

@@
 paucity of similarly large locally-available raw material sources.
-Bifaces larger than the mean were then assigned to the large category,
-those smaller than the mean were assigned to the small category, and the
-two datasets were then joined in advance of analysis
+Bifaces larger than the mean were assigned to the large category, those
+smaller than the mean were assigned to the small category, and the two
+datasets were joined in advance of analysis
 (\href{https://seldenlab.github.io/elliptical.bifaces/}{Supplementary
@@
 
-Elliptical bifaces from the Millsap Cache and Jowell Farm were used to
+Elliptical bifaces from Millsap Cache and Jowell Farm were used to
 examine whether biface morphology remains stable or expresses
@@
 modified approach to the flaking index developed by \citet{RN11099}. The
-modified approach uses counts of flake scars from each edge in the two
-most heavily worked areas, paired with a measure of edge length
-inclusive of curvature, which was calculated using ImageJ
+modified approach uses counts of flake scars from both sides of each
+edge in the two most heavily worked areas, paired with a measure of edge
+length inclusive of curvature, which was calculated using ImageJ
 \citep{RN11146,RN11147,RN11148}. The number of flake scars was
@@
 \ref{fig:elliptical}). Equidistant semilandmarks were placed between
-each landmark, and all landmarks and semilandmarks were applied using R
-\citep{R} and the \texttt{StereoMorph} package \citep{RN9091}.
+each landmark, applied using R \citep{R} and the \texttt{StereoMorph}
+package \citep{RN9091}.
 
@@
 described by each principal axis is commonly visualized using thin-plate
-spline warping of a reference image \citep{RN1731,RN479}.
+spline warping of a reference image or 3D mesh \citep{RN1731,RN479}.
 
@@
 
-To assess whether shape changes with size (allometry), and whether shape
+To assess whether shape changed with size (allometry), and whether shape
 and size differed by raw material and size class, Procrustes ANOVAs
@@
 
-The test for allometry in the sample did not return a significant
-result, suggesting that elliptical biface shape and size may follow a
-pattern of isometry, where shape remains constant through the reduction
-process
+The test for allometry did not return a significant result, suggesting
+that elliptical biface shape and size may follow a pattern of isometry,
+where shape remains constant throughout the reduction/retouch process
 (\href{https://seldenlab.github.io/elliptical.bifaces/}{Supplementary
-Materials}). Elliptical bifaces were also found to express a consistency
-of shape across all analyses
+Materials}). Elliptical bifaces were morphologically consistent in shape
+across all analyses
 (\href{https://seldenlab.github.io/elliptical.bifaces/}{Supplementary
@@
 site/raw material (RRPP = 10,000, Pr(\textgreater F) = 0.0007); however,
-the subsequent analysis of elliptical bifaces by size class demonstrates
-that the pattern is more readily attributable to reduction
-practices/intensity rather than nodule and raw material size
+the subsequent analysis by size class demonstrated that the pattern is
+more readily attributable to reduction practices/intensity rather than
+nodule and raw material size
 (\href{https://seldenlab.github.io/elliptical.bifaces/}{Supplementary
@@
 
-The analysis of morphology by size class demonstrates a significant
-difference in size between size classes; however, that pattern deviated
-from our expectations. Elliptical bifaces from Jowell Farm differ in
-neither shape nor size between the size classes
+The analysis of morphology by size class demonstrated a significant
+difference in size between size classes (expected); however, the pattern
+deviated from our expectations. Elliptical bifaces from Jowell Farm
+differ in neither shape nor size between size classes
 (\href{https://seldenlab.github.io/elliptical.bifaces/}{Supplementary
@@
 Thus, the pattern that emerges is one of differential retouch
-\emph{intensity}, where the isometric retouch strategies of Caddo
-knappers from Millsap Cache articulate with a different reduction
-practice thought to be associated with locally abundant raw material
-(Kay County flint). This stands in contrast to the comparatively minimal
-retouch intensity of another group (Jowell Farm) where local raw
-materials were of poor quality and insufficient size, yielding a
-situation where it was necessary to import raw material from an
-extralocal source (Edwards chert).
+\emph{intensity}, where isometric retouch strategies of Caddo knappers
+at Millsap Cache articulate with a different reduction practice thought
+to be associated with locally abundant raw material (Kay County flint).
+This finding is contrasted with the comparatively minimal retouch
+intensity of another group (Jowell Farm) where local raw materials were
+of poor quality and insufficient size, yielding a situation where it was
+necessary to import raw material from an extralocal source (Edwards
+chert).
 
@@
 differences in morphological diversity between size classes
-\citep{RN11107,RN7041,RN5694}. No difference was found for elliptical
+\citep{RN11107,RN7041,RN5694}. No difference was found in elliptical
 biface shape or size by site/raw material, but the comparison of
@@
 material of sufficient quality and size---resulting in a broader range
-of biface shapes in the small size class. Conversely, these results also
+of biface shapes in the small size class. Conversely, results also
 suggest that Caddo knappers at Millsap Cache employed a more

### 2022-11-20 supp/index.Rmd
<edit ms>

@@
 The number of flake scars was then divided by the length of the worked edge, with the heaviest worked edge identified by the greater value.
-
-The same landmark/semilandmark configuration was used for both analyses, and the landmarking protocol employs three landmarks; two horizontal tangents (top/bottom), and the third placed at the furthest extent of the retouched edge bearing the heaviest amount of retouch (see Figure \@ref(fig:figlm)).
 

### 2022-11-22 ms/ms.Rmd
<edit ms>

--- 2022-11-19 ms/ms.Rmd
@@
 keywords: |
-  American Southeast; Caddo; NAGPRA; archaeology; ovoid biface; Jowell knife; Jowell knives; Jowell biface; lithics; oblique parallel flaking; archaeoinformatics; museum studies; digital humanities; non-Western art history; geometric morphometrics; STEM; STEAM
+  American Southeast; Caddo; NAGPRA; archaeology; ovoid biface; Jowell knife; Jowell knives; Jowell biface; lithics; parallel oblique flaking; archaeoinformatics; museum studies; digital humanities; non-Western art history; geometric morphometrics; STEM; STEAM
 header-includes: |
@@
 \label{fig:map}
+\end{figure}
+
+## Parallel-oblique flaking
+
+When compared with other large finished bifaces in the region, the distinctive parallel oblique flake scar patterns on both faces of the elliptical bifaces for all, or virtually all, of their length is unique (Figure \ref{fig:illustration}).
+The absence of identified preforms for this artifact type makes technological and/or attribute-based discriminations related to manufacture a challenge.
+The projection of ritual and/or symbolic interpretation/s that articulate with this flake scar pattern is not possible at this time.
+Parallel oblique flake scar patterns are known to be characteristic of Guerrero arrow points \citep{RN9610} found at missions in Texas, several dart point types of the late Paleoindian period \citep{RN8584,RN7935,RN11182, RN9843,RN8143,RN10117,RN11181}, and this distinctive flaking pattern has also been inferred from ceramic motifs in Peru \citep{RN7938}.
+
+Parallel oblique flake scar patterns are not characteristic of any of the larger bifaces and bifacial knives in the American Southeast; thus, technological, cultural, or symbolic relationships between the elliptical bifaces and Guerrero arrow points or Angostura dart points are not suggested here.
+The technological choice to use parallel oblique flaking in finishing these elliptical bifaces most likely represents, at minimum, a stylistic choice by knappers limited to this specific implement; although, that may have been a functionally-relevant choice given their shape and size.
+In some respects, the decision to employ parallel oblique flaking is as distinctive as the presence---although not diagnostic \citep{RN6170}---of the flat and expanding flake scar pattern observed on Gahagan and Copena bifaces.
+
+\begin{figure}\centering
+\includegraphics[width=0.9\linewidth]{figs/elliptical.illustration.png}
+\caption{Illustrated elliptical bifaces from a, an unknown location; b, Jowell Farm; and c, Belcher Mound highlighting parallel oblique flaking used in production.
+The specimen from Belcher Mound is a potential preform exhibiting parallel oblique removals consistent with elliptical bifaces from other sites in the region.}
+\label{fig:illustration}
 \end{figure}
@@
 
-\begin{figure}\centering
-\includegraphics[width=\linewidth]{figs/elliptical.illustration.png}
-\caption{Illustrated elliptical bifaces from a, an unknown location; b, Jowell Farm; and c, Belcher Mound highlighting oblique parallel flaking used in production.
-The specimen from Belcher Mound is a potential preform exhibiting oblique parallel removals consistent with elliptical bifaces from other sites in the region.}
-\label{fig:illustration}
-\end{figure}
-
 # Conclusion

### 2022-11-22 ms/ms.tex
<edit ms>

--- 2022-11-19 ms/ms.tex
@@
 American Southeast; Caddo; NAGPRA; archaeology; ovoid biface; Jowell
-knife; Jowell knives; Jowell biface; lithics; oblique parallel flaking;
+knife; Jowell knives; Jowell biface; lithics; parallel oblique flaking;
 archaeoinformatics; museum studies; digital humanities; non-Western art
@@
 \label{fig:map}
+\end{figure}
+
+\hypertarget{parallel-oblique-flaking}{%
+\subsection{Parallel-oblique flaking}\label{parallel-oblique-flaking}}
+
+When compared with other large finished bifaces in the region, the
+distinctive parallel oblique flake scar patterns on both faces of the
+elliptical bifaces for all, or virtually all, of their length is unique
+(Figure \ref{fig:illustration}). The absence of identified preforms for
+this artifact type makes technological and/or attribute-based
+discriminations related to manufacture a challenge. The projection of
+ritual and/or symbolic interpretation/s that articulate with this flake
+scar pattern is not possible at this time. Parallel oblique flake scar
+patterns are known to be characteristic of Guerrero arrow points
+\citep{RN9610} found at missions in Texas, several dart point types of
+the late Paleoindian period
+\citep{RN8584,RN7935,RN11182, RN9843,RN8143,RN10117,RN11181}, and this
+distinctive flaking pattern has also been inferred from ceramic motifs
+in Peru \citep{RN7938}.
+
+Parallel oblique flake scar patterns are not characteristic of any of
+the larger bifaces and bifacial knives in the American Southeast; thus,
+technological, cultural, or symbolic relationships between the
+elliptical bifaces and Guerrero arrow points or Angostura dart points
+are not suggested here. The technological choice to use parallel oblique
+flaking in finishing these elliptical bifaces most likely represents, at
+minimum, a stylistic choice by knappers limited to this specific
+implement; although, that may have been a functionally-relevant choice
+given their shape and size. In some respects, the decision to employ
+parallel oblique flaking is as distinctive as the presence---although
+not diagnostic \citep{RN6170}---of the flat and expanding flake scar
+pattern observed on Gahagan and Copena bifaces.
+
+\begin{figure}\centering
+\includegraphics[width=0.9\linewidth]{figs/elliptical.illustration.png}
+\caption{Illustrated elliptical bifaces from a, an unknown location; b, Jowell Farm; and c, Belcher Mound highlighting parallel oblique flaking used in production.
+The specimen from Belcher Mound is a potential preform exhibiting parallel oblique removals consistent with elliptical bifaces from other sites in the region.}
+\label{fig:illustration}
 \end{figure}
@@
 
-\begin{figure}\centering
-\includegraphics[width=\linewidth]{figs/elliptical.illustration.png}
-\caption{Illustrated elliptical bifaces from a, an unknown location; b, Jowell Farm; and c, Belcher Mound highlighting oblique parallel flaking used in production.
-The specimen from Belcher Mound is a potential preform exhibiting oblique parallel removals consistent with elliptical bifaces from other sites in the region.}
-\label{fig:illustration}
-\end{figure}
-
 \hypertarget{conclusion}{%

### 2022-11-22 supp/01-size.class.Rmd
<edit index>
<edit ms>

@@
 
-Landmark data were aligned to a global coordinate system [@RN11622;@RN11623;@RN11563], achieved through generalized Procrustes superimposition [@RN478] performed in R 4.1.3 [@R] using the `geomorph` library v.
-4.0.3 [@RN11530;@RN1774;@RN9565].
-Procrustes superimposition translates, scales, and rotates the coordinate data to allow for comparisons among objects [@RN11564;@RN478].
-The `geomorph` package uses a partial Procrustes superimposition that projects the aligned specimens into tangent space subsequent to alignment in preparation for the use of multivariate methods that assume linear space [@RN1646;@RN11563].
+Landmark data were aligned to a global coordinate system [@RN8477;@RN7502;@RN11622;@RN11623;@RN11563], achieved through generalised Procrustes superimposition [@RN11138;@RN478;@RN1646] in R using the `geomorph` and `RRPP` packages [@RN1655;@RN11775;@RN11530;@RN1774;@RN9565].
+Procrustes superimposition translates and rotates the coordinate data to allow for comparisons among objects, while also scaling each biface using unit-centroid size---the square root of the sum of squared distances from each landmark to the specimen’s centroid [@RN11139;@RN11140;@RN11564;@RN478].
+The `geomorph` package uses a partial Procrustes superimposition that projects the aligned specimens into tangent space subsequent to alignment in preparation for the use of multivariate methods that assume linear space [@RN11141;@RN11142;@RN1646;@RN11563].
 
@@
 Y.gpa <- gpagen(shapesGM, print.progress = FALSE)
-
-# plot
-plot(Y.gpa)
 

### 2022-11-22 supp/02-2dgm-rawmat.Rmd
<edit ms>

@@
 
-Landmark data were aligned to a global coordinate system [@RN11622;@RN11623;@RN11563], achieved through generalized Procrustes superimposition [@RN478] performed in R 4.1.3 [@R] using the `geomorph` library v.
-4.0.4 [@RN11530;@RN1774;@RN9565].
-Procrustes superimposition translates, scales, and rotates the coordinate data to allow for comparisons among objects [@RN11564;@RN478].
-The `geomorph` package uses a partial Procrustes superimposition that projects the aligned specimens into tangent space subsequent to alignment in preparation for the use of multivariate methods that assume linear space [@RN1646;@RN11563].
+Landmark data were aligned to a global coordinate system [@RN8477;@RN7502;@RN11622;@RN11623;@RN11563], achieved through generalised Procrustes superimposition [@RN11138;@RN478;@RN1646] in R using the `geomorph` and `RRPP` packages [@RN1655;@RN11775;@RN11530;@RN1774;@RN9565].
+Procrustes superimposition translates and rotates the coordinate data to allow for comparisons among objects, while also scaling each biface using unit-centroid size---the square root of the sum of squared distances from each landmark to the specimen’s centroid [@RN11139;@RN11140;@RN11564;@RN478].
+The `geomorph` package uses a partial Procrustes superimposition that projects the aligned specimens into tangent space subsequent to alignment in preparation for the use of multivariate methods that assume linear space [@RN11141;@RN11142;@RN1646;@RN11563].
 
@@
 
-Principal components analysis [@RN1746] was used to visualise shape variation among the Perdiz arrow points.
-The shape changes described by each principal axis are commonly visualized using thin-plate spline warping of a reference 3D mesh [@RN1731;@RN479].
+Principal components analysis [@RN1746] was used to visualise shape variation among the elliptical bifaces, and the scatterplot represents the dispersion of shapes in tangent space [@RN8633;@RN5616;@RN11143;@RN7550].
+Shape ranges described by each principal axis are commonly visualized using thin-plate spline warping of a reference image [@RN1731;@RN479].
 

### 2022-11-22 ms/ms.Rmd

--- 2022-11-20 ms/ms.Rmd
@@
 \begin{figure}\centering
-\includegraphics[width=\linewidth]{figs/map.png}
+\includegraphics[width=0.9\linewidth]{figs/map.png}
 \caption{Location of Millsap Cache, Jowell Farm, and other sites mentioned in the text. Elliptical bifaces have been found in every major drainage basin in the ancestral Caddo area (white).}
 \label{fig:map}
+\end{figure}
+
+## Parallel-oblique flaking
+
+When compared with other large finished bifaces in the region, the distinctive parallel oblique flake scar patterns on both faces of the elliptical bifaces for all, or virtually all, of their length is unique (Figure \ref{fig:illustration}).
+The absence of identified preforms for this artifact type makes technological and/or attribute-based discriminations related to manufacture a challenge; however, one potential preform was recovered from mortuary contexts at Belcher Mound (Figure \ref{fig:illustration}c).
+The projection of ritual and/or symbolic interpretation/s that articulate with this flake scar pattern is not possible at this time.
+Parallel oblique flake scar patterns are known to be characteristic of Guerrero arrow points found at missions in Texas \citep{RN9610}, several dart point types of the late Paleoindian period \citep{RN8584,RN7935,RN11182, RN9843,RN8143,RN10117,RN11181}, and this distinctive flaking pattern has also been inferred from ceramic motifs in Peru \citep{RN7938}.
+
+Parallel oblique flake scar patterns are not characteristic of any of the larger bifaces and bifacial knives in the American Southeast; thus, technological, cultural, or symbolic relationships between the elliptical bifaces and Guerrero arrow points or Angostura dart points are not suggested here.
+The technological choice to use parallel oblique flaking in finishing these elliptical bifaces most likely represents, at minimum, a stylistic choice by knappers limited to this specific implement; although, that may have been a functionally-relevant choice given their shape and size.
+In some respects, the decision to employ parallel oblique flaking is as distinctive as the presence---although not diagnostic \citep{RN6170}---of the flat and expanding flake scar pattern observed on Gahagan and Copena bifaces.
+
+\begin{figure}\centering
+\includegraphics[width=0.9\linewidth]{figs/elliptical.illustration.png}
+\caption{Illustrated elliptical bifaces from a, an unknown location; b, Jowell Farm; and c, Belcher Mound highlighting parallel oblique flaking used in production.
+The specimen from Belcher Mound is a potential preform exhibiting parallel oblique removals consistent with elliptical bifaces from other sites in the region.}
+\label{fig:illustration}
 \end{figure}

### 2022-11-22 ms/ms.tex

--- 2022-11-20 ms/ms.tex
@@
 \begin{figure}\centering
-\includegraphics[width=\linewidth]{figs/map.png}
+\includegraphics[width=0.9\linewidth]{figs/map.png}
 \caption{Location of Millsap Cache, Jowell Farm, and other sites mentioned in the text. Elliptical bifaces have been found in every major drainage basin in the ancestral Caddo area (white).}
 \label{fig:map}
+\end{figure}
+
+\hypertarget{parallel-oblique-flaking}{%
+\subsection{Parallel-oblique flaking}\label{parallel-oblique-flaking}}
+
+When compared with other large finished bifaces in the region, the
+distinctive parallel oblique flake scar patterns on both faces of the
+elliptical bifaces for all, or virtually all, of their length is unique
+(Figure \ref{fig:illustration}). The absence of identified preforms for
+this artifact type makes technological and/or attribute-based
+discriminations related to manufacture a challenge; however, one
+potential preform was recovered from mortuary contexts at Belcher Mound
+(Figure \ref{fig:illustration}c). The projection of ritual and/or
+symbolic interpretation/s that articulate with this flake scar pattern
+is not possible at this time. Parallel oblique flake scar patterns are
+known to be characteristic of Guerrero arrow points found at missions in
+Texas \citep{RN9610}, several dart point types of the late Paleoindian
+period \citep{RN8584,RN7935,RN11182, RN9843,RN8143,RN10117,RN11181}, and
+this distinctive flaking pattern has also been inferred from ceramic
+motifs in Peru \citep{RN7938}.
+
+Parallel oblique flake scar patterns are not characteristic of any of
+the larger bifaces and bifacial knives in the American Southeast; thus,
+technological, cultural, or symbolic relationships between the
+elliptical bifaces and Guerrero arrow points or Angostura dart points
+are not suggested here. The technological choice to use parallel oblique
+flaking in finishing these elliptical bifaces most likely represents, at
+minimum, a stylistic choice by knappers limited to this specific
+implement; although, that may have been a functionally-relevant choice
+given their shape and size. In some respects, the decision to employ
+parallel oblique flaking is as distinctive as the presence---although
+not diagnostic \citep{RN6170}---of the flat and expanding flake scar
+pattern observed on Gahagan and Copena bifaces.
+
+\begin{figure}\centering
+\includegraphics[width=0.9\linewidth]{figs/elliptical.illustration.png}
+\caption{Illustrated elliptical bifaces from a, an unknown location; b, Jowell Farm; and c, Belcher Mound highlighting parallel oblique flaking used in production.
+The specimen from Belcher Mound is a potential preform exhibiting parallel oblique removals consistent with elliptical bifaces from other sites in the region.}
+\label{fig:illustration}
 \end{figure}

### 2022-11-27 supp/index.qmd
<edit struc>

--- 2022-11-20 supp/index.Rmd
@@
 author:
-    - Robert Z.
-Selden, Jr.^[Heritage Research Center, Stephen F.
+  - name: Robert Z. Selden, Jr.
+    affiliations:
+      - Heritage Research Center, Stephen F.
 Austin State University; Department of Biology, Stephen F.
-Austin State University; Texas Archeological Research Laboratory, The University of Texas at Austin; and Cultural Heritage Department, Jean Monnet University, zselden@sfasu.edu]
-date: "`r Sys.Date()`"
-site: bookdown::bookdown_site
-documentclass: book
+Austin State University; Texas Archeological Research Laboratory, The University of Texas at Austin; Cultural Heritage Department, Jean Monnet University
+    email: zselden@sfasu.edu
+    orcid: 0000-0002-1789-8449
+  - name: John E. Dockall
+    affiliations:
+      - Stantec, Inc.
+    email: john.dockall@stantec.com
+    orcid: 0000-0002-0940-7144
+  - name: David H. Dye
+    affiliations: 
+      - Department of Earth Sciences, The University of Memphis
+    email: daviddye@memphis.edu
 bibliography: [book.bib]
-biblio-style: apalike
+date: "`r format(Sys.time(), '%d %B, %Y')`"
 link-citations: yes
-description: "Pilot study of elliptical biface morphology."
+github-repo: seldenlab/elliptical.bifaces
 ---

### 2022-11-27 supp/index.qmd
<edit supp>

@@
 
-Prior to landmarking, elliptical bifaces were oriented with the most heavily retouched edge at top right (Figure \@ref(fig:figlm)).
+Prior to landmarking, elliptical bifaces were oriented with the most heavily retouched edge at top right.
 The landmarking protocol uses three landmarks; two horizontal tangents (top/bottom), and the third placed at the furthest extent of the retouched edge bearing the heaviest amount of retouch.
@@
 1.6.7 [@RN8973].
-Some bifaces include multiple areas of heavy retouch at the top and bottom of the same lateral edge (Figure \@ref(fig:figlm): b, c, d, and e), while others (Figure \@ref(fig:figlm): f) include one retouched edge at top right and another at bottom left, similar to alternately beveled bifaces.
+Some bifaces include multiple areas of heavy retouch at the top and bottom of the same lateral edge, while others include one retouched edge at top right and another at bottom left, similar to alternately beveled bifaces.
 

### 2022-11-27 supp/index.qmd
<edit index>

@@
 
-American Southeast, Caddo, NAGPRA, archaeoinformatics, museum studies, digital humanities, non-Western art history, geometric morphometrics, STEM, STEAM, ovoid biface, Jowell knife, Jowell knives, Jowell biface, lithics, oblique parallel flaking
+American Southeast, Caddo, NAGPRA, computational archaeology, archaeoinformatics, museum studies, digital humanities, non-Western art history, geometric morphometrics, STEM, STEAM, ovoid biface, Jowell knife, Jowell knives, Jowell biface, lithics, oblique parallel flaking
 
@@
 
-The pattern that emerges is one of differential retouch _intensity_, where the isometric retouch strategies of Caddo knappers from Millsap Cache articulate with a different reduction practice thought to be associated with locally abundant raw material (Kay County flint).
+The pattern that emerges is one of differential retouch _intensity_, where retouch strategies of Caddo knappers from Millsap Cache articulate with a reduction practice thought to be associated with locally abundant raw material (Kay County flint).
 This stands in contrast to the comparatively minimal retouch intensity of another group (Jowell Farm) where local raw materials were of poor quality and insufficient size, yielding a situation where it was necessary to import raw material from an extralocal source (Edwards chert).

### 2022-11-27 supp/sizeclass.qmd
<edit struc>
<edit index>

--- 2022-08-06 supp/01-size.class.Rmd
@@
-# Define size classes
+---
+title: "Define Size classes"
+bibliography: [book.bib]
+---
 
@@
 
-Landmark data were aligned to a global coordinate system [@RN11622;@RN11623;@RN11563], achieved through generalized Procrustes superimposition [@RN478] performed in R 4.1.3 [@R] using the `geomorph` library v.
-4.0.3 [@RN11530;@RN1774;@RN9565].
-Procrustes superimposition translates, scales, and rotates the coordinate data to allow for comparisons among objects [@RN11564;@RN478].
-The `geomorph` package uses a partial Procrustes superimposition that projects the aligned specimens into tangent space subsequent to alignment in preparation for the use of multivariate methods that assume linear space [@RN1646;@RN11563].
+Landmark data were aligned to a global coordinate system [@RN8477; @RN7502; @RN11622; @RN11623; @RN11563], achieved through generalised Procrustes superimposition [@RN11138; @RN478; @RN1646] in R using the `geomorph` and `RRPP` packages [@RN1655; @RN11775; @RN11530; @RN1774; @RN9565].
+Procrustes superimposition translates and rotates the coordinate data to allow for comparisons among objects, while also scaling each biface using unit-centroid size---the square root of the sum of squared distances from each landmark to the specimen's centroid [@RN11139; @RN11140; @RN11564; @RN478].
+The `geomorph` package uses a partial Procrustes superimposition that projects the aligned specimens into tangent space subsequent to alignment in preparation for the use of multivariate methods that assume linear space [@RN11141; @RN11142; @RN1646; @RN11563].
 

### 2022-11-27 supp/gmraw.qmd
<edit struc>
<edit index>

--- 2022-11-07 supp/02-2dgm-rawmat.Rmd
@@
-# GM - Site/Raw Material
+---
+title: "GM: Raw Material"
+bibliography: [book.bib]
+---
 
@@
 
-This analysis follows recent analyses of Perdiz arrow points ([Selden and Dockall, in review](https://github.com/seldenlab/perdiz3/blob/main/ms/perdiz3/perdiz3.pdf)) and Gahagan biface morphology across the southern Caddo area and central Texas [@RN11783;@RN21001;@RN11264], where Perdiz arrow points and Gahagan bifaces have been found to differ in shape across the same geography as Caddo bottles [@RN11801;@RN11782;@RN11716;@RN8312].
-This analysis also follows another geometric morphometric analysis in which Perdiz arrow points from the ancestral Caddo area were found to differ as a function of _time_, _raw material_, and _burial context_ [@RN9551].
+This analysis follows recent analyses of Perdiz arrow points ([Selden and Dockall, in review](https://github.com/seldenlab/perdiz3/blob/main/ms/perdiz3/perdiz3.pdf)) and Gahagan biface morphology across the southern Caddo area and central Texas [@RN11783; @RN21001; @RN11264], where Perdiz arrow points and Gahagan bifaces have been found to differ in shape across the same geography as Caddo bottles [@RN11801; @RN11782; @RN11716; @RN8312].
+This analysis also follows another geometric morphometric analysis in which Perdiz arrow points from the ancestral Caddo area were found to differ as a function of *time*, *raw material*, and *burial context* [@RN9551].
 
@@
 
+# read data and define number of sLMs ----
+shapes <- readShapes("shapes")
+shapesGM <- readland.shapes(shapes, nCurvePts = c(15, 20, 35))
+
 # read qualitative data
@@
 
-Landmark data were aligned to a global coordinate system [@RN11622;@RN11623;@RN11563], achieved through generalized Procrustes superimposition [@RN478] performed in R 4.1.3 [@R] using the `geomorph` library v.
-4.0.4 [@RN11530;@RN1774;@RN9565].
-Procrustes superimposition translates, scales, and rotates the coordinate data to allow for comparisons among objects [@RN11564;@RN478].
-The `geomorph` package uses a partial Procrustes superimposition that projects the aligned specimens into tangent space subsequent to alignment in preparation for the use of multivariate methods that assume linear space [@RN1646;@RN11563].
+Landmark data were aligned to a global coordinate system [@RN8477; @RN7502; @RN11622; @RN11623; @RN11563], achieved through generalised Procrustes superimposition [@RN11138; @RN478; @RN1646] in R using the `geomorph` and `RRPP` packages [@RN1655; @RN11775; @RN11530; @RN1774; @RN9565].
+Procrustes superimposition translates and rotates the coordinate data to allow for comparisons among objects, while also scaling each biface using unit-centroid size---the square root of the sum of squared distances from each landmark to the specimen's centroid [@RN11139; @RN11140; @RN11564; @RN478].
+The `geomorph` package uses a partial Procrustes superimposition that projects the aligned specimens into tangent space subsequent to alignment in preparation for the use of multivariate methods that assume linear space [@RN11141; @RN11142; @RN1646; @RN11563].
 
@@
 
-Principal components analysis [@RN1746] was used to visualise shape variation among the Perdiz arrow points.
-The shape changes described by each principal axis are commonly visualized using thin-plate spline warping of a reference 3D mesh [@RN1731;@RN479].
+Principal components analysis [@RN1746] was used to visualise shape variation among the elliptical bifaces, and the scatterplot represents the dispersion of shapes in tangent space [@RN8633; @RN5616; @RN11143; @RN7550].
+Shape ranges described by each principal axis are commonly visualized using thin-plate spline warping of a reference image [@RN1731; @RN479].
 
@@
 To assess whether shape changes with size (allometry), Procrustes ANOVAs [@RN1749] were run that enlist effect-sizes (zscores) computed as standard deviates of the generated sampling distributions [@RN1756].
-A residual randomization permutation procedure (RRPP; n = 10,000 permutations) was used for all Procrustes ANOVAs [@RN1655;@RN11775], which has higher statistical power and a greater ability to identify patterns in the data should they be present [@RN1719].
+A residual randomization permutation procedure (RRPP; n = 10,000 permutations) was used for all Procrustes ANOVAs [@RN1655; @RN11775], which has higher statistical power and a greater ability to identify patterns in the data should they be present [@RN1719].
 
@@
 To assess whether shape and size differ by `size class`, Procrustes ANOVAs [@RN1749] were run that enlist effect-sizes (zscores) computed as standard deviates of the generated sampling distributions [@RN1756].
-A residual randomization permutation procedure (RRPP; n = 10,000 permutations) was used for all Procrustes ANOVAs [@RN1655;@RN11775], which has higher statistical power and a greater ability to identify patterns in the data should they be present [@RN1719].
+A residual randomization permutation procedure (RRPP; n = 10,000 permutations) was used for all Procrustes ANOVAs [@RN1655; @RN11775], which has higher statistical power and a greater ability to identify patterns in the data should they be present [@RN1719].
 
@@
 
-The trajectory analysis was followed by an analysis of morphological disparity [@RN11107;@RN7041;@RN5694] to identify potential differences in morphological diversity between size classes.
+The trajectory analysis was followed by an analysis of morphological disparity [@RN11107; @RN7041; @RN5694] to identify potential differences in morphological diversity between size classes.
 

### 2022-11-27 supp/gmraw.qmd
<edit index>

@@
 
-Elliptical bifaces from the Millsap Cache and Jowell Farm were used to examine whether their morphology remains stable, or expresses morphological variability by site/raw material.
-Variability in elliptical biface morphology provides evidence associated with differences in morphology as a function of geography/raw material.
+Bifaces from Millsap Cache and Jowell Farm were used to assess morphological stability/variability in elliptical bifaces by site/raw material.
+Variability in elliptical biface morphology would provide evidence of local differences in morphology as a function of geography/raw material.
 
-This analysis follows recent analyses of Perdiz arrow points ([Selden and Dockall, in review](https://github.com/seldenlab/perdiz3/blob/main/ms/perdiz3/perdiz3.pdf)) and Gahagan biface morphology across the southern Caddo area and central Texas [@RN11783; @RN21001; @RN11264], where Perdiz arrow points and Gahagan bifaces have been found to differ in shape across the same geography as Caddo bottles [@RN11801; @RN11782; @RN11716; @RN8312].
+This analysis follows recent formal analyses of Perdiz arrow points [@RN9551;@RN11264] as well as another that is in press ([Selden and Dockall, in press](https://github.com/seldenlab/perdiz3/blob/main/ms/perdiz3.pdf)), and Gahagan biface morphology across the southern Caddo area and central Texas [@RN11783; @RN21001; @RN11264], where Perdiz arrow points and Gahagan bifaces have been found to differ in shape across the same geography as Caddo bottles [@RN11801; @RN11782; @RN11716; @RN8312].
 This analysis also follows another geometric morphometric analysis in which Perdiz arrow points from the ancestral Caddo area were found to differ as a function of *time*, *raw material*, and *burial context* [@RN9551].

### 2022-11-27 supp/gmszclass.qmd
<edit struc>
<edit index>

--- 2022-11-22 supp/03-2dgm-sizeclass.Rmd
@@
-# GM - Size Class
+---
+title: "GM: Size Class"
+bibliography: [book.bib]
+---
 
@@
 
-This analysis follows recent analyses of Perdiz arrow points ([Selden and Dockall, in review](https://github.com/seldenlab/perdiz3/blob/main/ms/perdiz3/perdiz3.pdf)) and Gahagan biface morphology across the southern Caddo area and central Texas [@RN11783;@RN21001;@RN11264], where Perdiz arrow points and Gahagan bifaces have been found to differ in shape across the same geography as Caddo bottles [@RN11801;@RN11782;@RN11716;@RN8312].
+This analysis follows recent formal analyses of Perdiz arrow points [@RN9551;@RN11264] as well as another that is in press ([Selden and Dockall, in press](https://github.com/seldenlab/perdiz3/blob/main/ms/perdiz3.pdf)), and Gahagan biface morphology across the southern Caddo area and central Texas [@RN11783; @RN21001; @RN11264], where Perdiz arrow points and Gahagan bifaces have been found to differ in shape across the same geography as Caddo bottles [@RN11801; @RN11782; @RN11716; @RN8312].
 This analysis also follows another geometric morphometric analysis in which Perdiz arrow points from the ancestral Caddo area were found to differ as a function of _time_, _raw material_, and _burial context_ [@RN9551].
@@
 library(ggplot2)
+
+# read data and define number of sLMs ----
+shapes <- readShapes("shapes")
+shapesGM <- readland.shapes(shapes, nCurvePts = c(15, 20, 35))
 

### 2022-11-28 supp/index.qmd
<edit index>

@@
 
-# Preface {-}
+>This process of comparison, of recognising in one form a definite permutation or _deformation_ of another, apart altogether from a precise and adequate understanding of the original "type" or standard of comparison, lies within the immediate province of mathematics, and finds its solution in the elementary use of a certain method of the mathematician [@RN7522].
 
->This process of comparison, of recognising in one form a definite permutation or _deformation_ of another, apart altogether from a precise and adequate understanding of the original "type" or standard of comparison, lies within the immediate province of mathematics, and finds its solution in the elementary use of a certain method of the mathematician [@RN7522].
+# Basis of inquiry
 
@@
 
-American Southeast, Caddo, NAGPRA, computational archaeology, archaeoinformatics, museum studies, digital humanities, non-Western art history, geometric morphometrics, STEM, STEAM, ovoid biface, Jowell knife, Jowell knives, Jowell biface, lithics, oblique parallel flaking
+American Southeast, Caddo, NAGPRA, computational archaeology, archaeoinformatics, museum studies, digital humanities, non-Western art history, geometric morphometrics, STEM, STEAM, ovoid biface, Galt biface, Jowell knife, Jowell knives, Jowell biface, Nikolas knife, lithics, oblique parallel flaking
 
@@
 
-### Summary
+## Summary
 

### 2022-11-28 ms/ms.Rmd
<edit ms>

@@
 keywords: |
-  American Southeast; Caddo; NAGPRA; archaeology; ovoid biface; Jowell knife; Jowell knives; Jowell biface; lithics; parallel oblique flaking; archaeoinformatics; museum studies; digital humanities; non-Western art history; geometric morphometrics; STEM; STEAM
+  American Southeast; Caddo; NAGPRA; computational archaeology; Galt biface;  Jowell knife; Jowell knives; Jowell biface; Nikolas knife; Nikolas knives; ovoid biface; lithics; parallel oblique flaking; archaeoinformatics; museum studies; digital humanities; non-Western art history; geometric morphometrics; STEM; STEAM
 header-includes: |
@@
 \label{fig:map}
-\end{figure}
-
-## Parallel-oblique flaking
-
-When compared with other large finished bifaces in the region, the distinctive parallel oblique flake scar patterns on both faces of the elliptical bifaces for all, or virtually all, of their length is unique (Figure \ref{fig:illustration}).
-The absence of identified preforms for this artifact type makes technological and/or attribute-based discriminations related to manufacture a challenge; however, one potential preform was recovered from mortuary contexts at Belcher Mound (Figure \ref{fig:illustration}c).
-The projection of ritual and/or symbolic interpretation/s that articulate with this flake scar pattern is not possible at this time.
-Parallel oblique flake scar patterns are known to be characteristic of Guerrero arrow points found at missions in Texas \citep{RN9610}, several dart point types of the late Paleoindian period \citep{RN8584,RN7935,RN11182, RN9843,RN8143,RN10117,RN11181}, and this distinctive flaking pattern has also been inferred from ceramic motifs in Peru \citep{RN7938}.
-
-Parallel oblique flake scar patterns are not characteristic of any of the larger bifaces and bifacial knives in the American Southeast; thus, technological, cultural, or symbolic relationships between the elliptical bifaces and Guerrero arrow points or Angostura dart points are not suggested here.
-The technological choice to use parallel oblique flaking in finishing these elliptical bifaces most likely represents, at minimum, a stylistic choice by knappers limited to this specific implement; although, that may have been a functionally-relevant choice given their shape and size.
-In some respects, the decision to employ parallel oblique flaking is as distinctive as the presence---although not diagnostic \citep{RN6170}---of the flat and expanding flake scar pattern observed on Gahagan and Copena bifaces.
-
-\begin{figure}\centering
-\includegraphics[width=0.9\linewidth]{figs/elliptical.illustration.png}
-\caption{Illustrated elliptical bifaces from a, an unknown location; b, Jowell Farm; and c, Belcher Mound highlighting parallel oblique flaking used in production.
-The specimen from Belcher Mound is a potential preform exhibiting parallel oblique removals consistent with elliptical bifaces from other sites in the region.}
-\label{fig:illustration}
 \end{figure}
@@
 
+## Parallel-oblique flaking
+
+When compared with other large finished bifaces in the region, the distinctive parallel oblique flake scar patterns on both faces of the elliptical bifaces for all, or virtually all, of their length is unique (Figure \ref{fig:illustration}).
+The absence of identified preforms for this artifact type makes technological and/or attribute-based discriminations related to manufacture a challenge; however, one potential preform was recovered from mortuary contexts at Belcher Mound (Figure \ref{fig:illustration}c).
+The projection of ritual and/or symbolic interpretation/s that articulate with this flake scar pattern is not possible at this time.
+Parallel oblique flake scar patterns are known to be characteristic of Guerrero arrow points found at missions in Texas \citep{RN9610}, several dart point types of the late Paleoindian period \citep{RN8584,RN7935,RN11182, RN9843,RN8143,RN10117,RN11181}, and this distinctive flaking pattern has also been inferred from ceramic motifs in Peru \citep{RN7938}.
+
+Parallel oblique flake scar patterns are not characteristic of any of the larger bifaces and bifacial knives in the American Southeast; thus, technological, cultural, or symbolic relationships between the elliptical bifaces and Guerrero arrow points or Angostura dart points are not suggested here.
+The technological choice to use parallel oblique flaking in finishing these elliptical bifaces most likely represents, at minimum, a stylistic choice by knappers limited to this specific implement; although, that may have been a functionally-relevant choice given their shape and size.
+In some respects, the decision to employ parallel oblique flaking is as distinctive as the presence---although not diagnostic \citep{RN6170}---of the flat and expanding flake scar pattern observed on Gahagan and Copena bifaces.
+
+\begin{figure}\centering
+\includegraphics[width=0.9\linewidth]{figs/elliptical.illustration.png}
+\caption{Illustrated elliptical bifaces from a, an unknown location; b, Jowell Farm; and c, Belcher Mound highlighting parallel oblique flaking used in production.
+The specimen from Belcher Mound is a potential preform exhibiting parallel oblique removals consistent with elliptical bifaces from other sites in the region.}
+\label{fig:illustration}
+\end{figure}
+
 # Conclusion

### 2022-11-28 ms/ms.tex
<edit ms>

@@
 \begin{keywords}
-American Southeast; Caddo; NAGPRA; archaeology; ovoid biface; Jowell
-knife; Jowell knives; Jowell biface; lithics; parallel oblique flaking;
+American Southeast; Caddo; NAGPRA; computational archaeology; Galt
+biface; Jowell knife; Jowell knives; Jowell biface; Nikolas knife;
+Nikolas knives; ovoid biface; lithics; parallel oblique flaking;
 archaeoinformatics; museum studies; digital humanities; non-Western art
@@
 \label{fig:map}
-\end{figure}
-
-\hypertarget{parallel-oblique-flaking}{%
-\subsection{Parallel-oblique flaking}\label{parallel-oblique-flaking}}
-
-When compared with other large finished bifaces in the region, the
-distinctive parallel oblique flake scar patterns on both faces of the
-elliptical bifaces for all, or virtually all, of their length is unique
-(Figure \ref{fig:illustration}). The absence of identified preforms for
-this artifact type makes technological and/or attribute-based
-discriminations related to manufacture a challenge; however, one
-potential preform was recovered from mortuary contexts at Belcher Mound
-(Figure \ref{fig:illustration}c). The projection of ritual and/or
-symbolic interpretation/s that articulate with this flake scar pattern
-is not possible at this time. Parallel oblique flake scar patterns are
-known to be characteristic of Guerrero arrow points found at missions in
-Texas \citep{RN9610}, several dart point types of the late Paleoindian
-period \citep{RN8584,RN7935,RN11182, RN9843,RN8143,RN10117,RN11181}, and
-this distinctive flaking pattern has also been inferred from ceramic
-motifs in Peru \citep{RN7938}.
-
-Parallel oblique flake scar patterns are not characteristic of any of
-the larger bifaces and bifacial knives in the American Southeast; thus,
-technological, cultural, or symbolic relationships between the
-elliptical bifaces and Guerrero arrow points or Angostura dart points
-are not suggested here. The technological choice to use parallel oblique
-flaking in finishing these elliptical bifaces most likely represents, at
-minimum, a stylistic choice by knappers limited to this specific
-implement; although, that may have been a functionally-relevant choice
-given their shape and size. In some respects, the decision to employ
-parallel oblique flaking is as distinctive as the presence---although
-not diagnostic \citep{RN6170}---of the flat and expanding flake scar
-pattern observed on Gahagan and Copena bifaces.
-
-\begin{figure}\centering
-\includegraphics[width=0.9\linewidth]{figs/elliptical.illustration.png}
-\caption{Illustrated elliptical bifaces from a, an unknown location; b, Jowell Farm; and c, Belcher Mound highlighting parallel oblique flaking used in production.
-The specimen from Belcher Mound is a potential preform exhibiting parallel oblique removals consistent with elliptical bifaces from other sites in the region.}
-\label{fig:illustration}
 \end{figure}
@@
 
+\hypertarget{parallel-oblique-flaking}{%
+\subsection{Parallel-oblique flaking}\label{parallel-oblique-flaking}}
+
+When compared with other large finished bifaces in the region, the
+distinctive parallel oblique flake scar patterns on both faces of the
+elliptical bifaces for all, or virtually all, of their length is unique
+(Figure \ref{fig:illustration}). The absence of identified preforms for
+this artifact type makes technological and/or attribute-based
+discriminations related to manufacture a challenge; however, one
+potential preform was recovered from mortuary contexts at Belcher Mound
+(Figure \ref{fig:illustration}c). The projection of ritual and/or
+symbolic interpretation/s that articulate with this flake scar pattern
+is not possible at this time. Parallel oblique flake scar patterns are
+known to be characteristic of Guerrero arrow points found at missions in
+Texas \citep{RN9610}, several dart point types of the late Paleoindian
+period \citep{RN8584,RN7935,RN11182, RN9843,RN8143,RN10117,RN11181}, and
+this distinctive flaking pattern has also been inferred from ceramic
+motifs in Peru \citep{RN7938}.
+
+Parallel oblique flake scar patterns are not characteristic of any of
+the larger bifaces and bifacial knives in the American Southeast; thus,
+technological, cultural, or symbolic relationships between the
+elliptical bifaces and Guerrero arrow points or Angostura dart points
+are not suggested here. The technological choice to use parallel oblique
+flaking in finishing these elliptical bifaces most likely represents, at
+minimum, a stylistic choice by knappers limited to this specific
+implement; although, that may have been a functionally-relevant choice
+given their shape and size. In some respects, the decision to employ
+parallel oblique flaking is as distinctive as the presence---although
+not diagnostic \citep{RN6170}---of the flat and expanding flake scar
+pattern observed on Gahagan and Copena bifaces.
+
+\begin{figure}\centering
+\includegraphics[width=0.9\linewidth]{figs/elliptical.illustration.png}
+\caption{Illustrated elliptical bifaces from a, an unknown location; b, Jowell Farm; and c, Belcher Mound highlighting parallel oblique flaking used in production.
+The specimen from Belcher Mound is a potential preform exhibiting parallel oblique removals consistent with elliptical bifaces from other sites in the region.}
+\label{fig:illustration}
+\end{figure}
+
 \hypertarget{conclusion}{%

### 2022-11-30 supp/index.qmd
<edit index>

@@
 
-American Southeast, Caddo, NAGPRA, computational archaeology, archaeoinformatics, museum studies, digital humanities, non-Western art history, geometric morphometrics, STEM, STEAM, ovoid biface, Galt biface, Jowell knife, Jowell knives, Jowell biface, Nikolas knife, lithics, oblique parallel flaking
+American Southeast, Caddo, NAGPRA, computational archaeology, archaeoinformatics, museum studies, digital humanities, non-Western art history, geometric morphometrics, STEM, STEAM, ovoid biface, Galt biface, Jowell knife, Jowell knives, Jowell biface, Nikolas knife, lithics, parallel-oblique flaking
 
@@
 
+## Parallel-oblique flaking
+
+When compared with other large finished bifaces in the region, the distinctive parallel oblique flake scar patterns on both faces of the elliptical bifaces for all, or virtually all, of their length is unique.
+The absence of identified preforms for this artifact type makes technological and/or attribute-based discriminations related to manufacture a challenge; however, one potential preform was recovered from mortuary contexts at Belcher Mound.
+The projection of ritual and/or symbolic interpretation/s that articulate with this flake scar pattern is not possible at this time.
+Parallel oblique flake scar patterns are known to be characteristic of several dart point types of the late Paleoindian period [@RN8584;@RN7935;@RN11182;@RN9843;@RN8143;@RN10117;@RN11181], and has even been inferred from ceramic motifs in Peru [@RN7938].
+
+Parallel oblique flake scar patterns are not characteristic of larger bifaces and bifacial knives in the American Southeast; thus, technological, cultural, or symbolic relationships between the elliptical bifaces and Angostura/Paleoindian dart points are not suggested here.
+The technological choice to use parallel oblique flaking in finishing the elliptical bifaces most likely represents, at minimum, a stylistic choice by knappers limited to this specific implement; although, that may have been a functionally-relevant choice given their shape and size.
+In some respects, the decision to employ parallel oblique flaking is as distinctive as the presence---although not diagnostic [@RN6170]---of the flat and expanding flake scar pattern observed on Gahagan and Copena bifaces.
+
 ```{r figillust, fig.cap="Illustrated elliptical bifaces from a, an unknown location; b, Jowell Farm; and c, Belcher Mound highlighting oblique parallel flaking used in production.

### 2022-12-02 ms/ms.Rmd

@@
 The projection of ritual and/or symbolic interpretation/s that articulate with this flake scar pattern is not possible at this time.
-Parallel oblique flake scar patterns are known to be characteristic of Guerrero arrow points found at missions in Texas \citep{RN9610}, several dart point types of the late Paleoindian period \citep{RN8584,RN7935,RN11182, RN9843,RN8143,RN10117,RN11181}, and this distinctive flaking pattern has also been inferred from ceramic motifs in Peru \citep{RN7938}.
+Parallel oblique flake scar patterns are known to be characteristic of several dart point types of the late Paleoindian period \citep{RN8584,RN7935,RN11182, RN9843,RN8143,RN10117,RN11181}, and has even been inferred from ceramic motifs in Peru \citep{RN7938}.
 
-Parallel oblique flake scar patterns are not characteristic of any of the larger bifaces and bifacial knives in the American Southeast; thus, technological, cultural, or symbolic relationships between the elliptical bifaces and Guerrero arrow points or Angostura dart points are not suggested here.
-The technological choice to use parallel oblique flaking in finishing these elliptical bifaces most likely represents, at minimum, a stylistic choice by knappers limited to this specific implement; although, that may have been a functionally-relevant choice given their shape and size.
+Parallel oblique flake scar patterns are not characteristic of larger bifaces and bifacial knives in the American Southeast; thus, technological, cultural, or symbolic relationships between the elliptical bifaces and Angostura/Paleoindian dart points are not suggested here.
+The technological choice to use parallel oblique flaking in finishing the elliptical bifaces most likely represents, at minimum, a stylistic choice by knappers limited to this specific implement; although, that may have been a functionally-relevant choice given their shape and size.
 In some respects, the decision to employ parallel oblique flaking is as distinctive as the presence---although not diagnostic \citep{RN6170}---of the flat and expanding flake scar pattern observed on Gahagan and Copena bifaces.

### 2022-12-02 ms/ms.tex

@@
 is not possible at this time. Parallel oblique flake scar patterns are
-known to be characteristic of Guerrero arrow points found at missions in
-Texas \citep{RN9610}, several dart point types of the late Paleoindian
-period \citep{RN8584,RN7935,RN11182, RN9843,RN8143,RN10117,RN11181}, and
-this distinctive flaking pattern has also been inferred from ceramic
-motifs in Peru \citep{RN7938}.
+known to be characteristic of several dart point types of the late
+Paleoindian period
+\citep{RN8584,RN7935,RN11182, RN9843,RN8143,RN10117,RN11181}, and has
+even been inferred from ceramic motifs in Peru \citep{RN7938}.
 
-Parallel oblique flake scar patterns are not characteristic of any of
-the larger bifaces and bifacial knives in the American Southeast; thus,
+Parallel oblique flake scar patterns are not characteristic of larger
+bifaces and bifacial knives in the American Southeast; thus,
 technological, cultural, or symbolic relationships between the
-elliptical bifaces and Guerrero arrow points or Angostura dart points
-are not suggested here. The technological choice to use parallel oblique
-flaking in finishing these elliptical bifaces most likely represents, at
-minimum, a stylistic choice by knappers limited to this specific
-implement; although, that may have been a functionally-relevant choice
-given their shape and size. In some respects, the decision to employ
-parallel oblique flaking is as distinctive as the presence---although
-not diagnostic \citep{RN6170}---of the flat and expanding flake scar
-pattern observed on Gahagan and Copena bifaces.
+elliptical bifaces and Angostura/Paleoindian dart points are not
+suggested here. The technological choice to use parallel oblique flaking
+in finishing the elliptical bifaces most likely represents, at minimum,
+a stylistic choice by knappers limited to this specific implement;
+although, that may have been a functionally-relevant choice given their
+shape and size. In some respects, the decision to employ parallel
+oblique flaking is as distinctive as the presence---although not
+diagnostic \citep{RN6170}---of the flat and expanding flake scar pattern
+observed on Gahagan and Copena bifaces.
 

### 2022-12-08 supp/index.qmd
<edit supp>
<edit index>

@@
 The landmarking protocol uses three landmarks; two horizontal tangents (top/bottom), and the third placed at the furthest extent of the retouched edge bearing the heaviest amount of retouch.
-Landmarks and semilandmarks were applied in R 4.2.1 [@R] using the `StereoMorph` library v.
-1.6.7 [@RN8973].
+Landmarks and semilandmarks were applied in R [@R] using the `StereoMorph` library [@RN8973].
 Some bifaces include multiple areas of heavy retouch at the top and bottom of the same lateral edge, while others include one retouched edge at top right and another at bottom left, similar to alternately beveled bifaces.
@@
 
+```{r figelliptical, fig.cap="A. T. Jackson with the elliptical bifaces from Galt Farm in Franklin County during excavations by the University of Texas in 1931."}
+
+knitr::include_graphics('./images/41fk2-2.jpg')
+```
+
 Further, the comparison of morphological diversity by size class demonstrated that elliptical biface shapes in the small size class at Jowell Farm occupy a significantly greater range of morphospace than those from Millsap Cache.

### 2022-12-16 ms/ms.Rmd
<edit index>

@@
 \includegraphics[width=0.9\linewidth]{figs/map.png}
-\caption{Location of Millsap Cache, Jowell Farm, and other sites mentioned in the text. Elliptical bifaces have been found in every major drainage basin in the ancestral Caddo area (white).}
+\caption{Location of Millsap Cache, Jowell Farm, and other sites mentioned in the text (additional information included at https://seldenlab.github.io/elliptical.bifaces/).
+Elliptical bifaces have been found in all but one major drainage basin in the ancestral Caddo area (white).}
 \label{fig:map}

### 2022-12-16 supp/index.qmd
<edit index>

@@
 
+Elliptical bifaces have a deep history in eastern North America extending from the Middle Archaic to well past European contact.
+In many instances they have been deposited in caches, which may represent large, corporate bundles decommissioned amid polity demise or radical transformations in social practice.
+Stylistic changes are evident in both length and width for these bifaces, and result in great variability in shape and size.
+While some undoubtedly were knapped for every-day, mundane uses, others were crafted from exotic cherts and reflect non-utilitarian uses.
+The _oversize_ or _exaggerated_ forms, mimicking combat weapons, are often referred to as symbolic weaponry.
+Such objects are typically those “whose size, delicacy, material of manufacture, and craftsmanship are outside the bounds of any weapon that might have been actually used” [@RN11177].
+
+The earliest elliptical bifaces, associated with Middle Archaic period, recovered as Benton caches (4500—4000 BCE) in the Mid-South [@RN11179], are clearly non-utilitarian, as they “meet the expectations of sacred markers for secular exchange” [@RN11178].
+In the western Middle Tennessee Valley, the Pickwick Burial Complex [@RN11180], dating to the Late Archaic/Gulf Formational period (3000—300 BCE) exhibits a continuation of bipointed blades, some notched or truncated, found in ritual caches and ceremonially broken.
+Elliptical bifaces find their most dramatic expression as Caddoan and Mississippian exaggerated, symbolic weaponry.
+Mississippian Ramey knives, dating from c.
+CE 1000 to 1300 are found associated with Cahokia political dynamics.
+The slightly later Duck River sword-form bifaces (CE 1275-1400) continue the emphasis on high-quality knapping and long-distance circulation.
+Such lanceolate blades, although smaller, continue into the early seventeenth century when they serve as regalia among political and social elites in ritual performances.
+
+```{r figorget, fig.cap="Shell gorgets from sites in the American Southeast depicting elliptical bifaces."}
+
+knitr::include_graphics('./images/composite.figure2.jpg')
+```
+
+Duck River and other sword-form bifaces are also present at Spiro Mound [@RN2720], with the longest measuring 55.9cm.
+Smaller elliptical biface forms, roughly half the size of the larger form, have also been identified at Spiro, along with several wooden implements that Early [-@RN1052] interpreted as memetic of elliptical bifaces.
+The bifaces from this latter category are the focus of this study, which is aimed at assessing whether resharpening and/or retouch may have resulted in convergent morphology at two sites in the ancestral Caddo area.
+A measure of morphological disparity is subsequently to assess whether biface morphology is more diverse at the larger size, which would suggest greater diversity in biface shape or size related to production activities.
+
 ## Analysis overview
 
-Prior to landmarking, elliptical bifaces were oriented with the most heavily retouched edge at top right.
+Prior to landmarking, elliptical bifaces from Millsap Cache and Jowell Farm were oriented with the most heavily retouched edge at top right.
 The landmarking protocol uses three landmarks; two horizontal tangents (top/bottom), and the third placed at the furthest extent of the retouched edge bearing the heaviest amount of retouch.
-Landmarks and semilandmarks were applied in R [@R] using the `StereoMorph` library [@RN8973].
+Landmarks and semilandmarks were applied in R 4.2.1 [@R] using the `StereoMorph` library v.
+1.6.7 [@RN8973].
 Some bifaces include multiple areas of heavy retouch at the top and bottom of the same lateral edge, while others include one retouched edge at top right and another at bottom left, similar to alternately beveled bifaces.

### 2022-12-19 supp/index.qmd
<edit index>

@@
 
-Duck River and other sword-form bifaces are also present at Spiro Mound [@RN2720], with the longest measuring 55.9cm.
+Duck River and other sword-form bifaces are also present at Spiro Mound [@RN2720], with the longest measuring 55.9cm, or just over 22 inches.
 Smaller elliptical biface forms, roughly half the size of the larger form, have also been identified at Spiro, along with several wooden implements that Early [-@RN1052] interpreted as memetic of elliptical bifaces.
-The bifaces from this latter category are the focus of this study, which is aimed at assessing whether resharpening and/or retouch may have resulted in convergent morphology at two sites in the ancestral Caddo area.
-A measure of morphological disparity is subsequently to assess whether biface morphology is more diverse at the larger size, which would suggest greater diversity in biface shape or size related to production activities.
+However, there is a morphological difference in the sword-form and smaller elliptical biface forms at Spiro, where both ends are mirrored, and are either pointed _or_ rounded, and not pointed _and_ rounded.
+
+Among the shell cup illustrations assembled by Phillips and Brown [-@RN2447] for the materials from Spiro Mound, is a depiction of smaller elliptical bifaces, which are conveyed as elements of the serpent/snake dance rather than elements of the eagle/bird dance.
+It is entirely possible that biface **shape** may convey a relevant---and ceremonially important---depiction related to the difference in local viper species.
+Differences in biface **size** may articulate with the size of specific viper species, as some (i.e.
+_Sistrurus miliarius streckeri_ or western pygmy rattlesnake) are physically smaller than others.
+Alternatively, **size** differences might also depict allometric trends associated with the development and maturation processes for various vipers.
+
+```{r figspiro, fig.cap="Shell cup from Spiro Mound depicting the serpent dance, where the difference in elliptical biface morphology is made plain.
+This figure was redrafted from Figure 44 in the Phillips and Brown series [-@RN2447]."}
+
+knitr::include_graphics('./images/spiro.png')
+```
+
+Using the tools of landmark geometric morphometrics, this study assesses whether resharpening and/or retouch of these important ceremonial bifaces may have resulted in convergent morphology at two sites in the ancestral Caddo area.
+A measure of morphological disparity is then used to assess whether biface morphology may have been more diverse at the larger size, suggestive of greater diversity in biface shape or size related to production activities.
 

### 2022-12-19 supp/gmraw.qmd
<edit index>

--- 2022-11-27 supp/gmraw.qmd
@@
 
-Elliptical bifaces from the Millsap Cache and Jowell Farm were used to examine whether their morphology remains stable, or expresses morphological variability by site/raw material.
-Variability in elliptical biface morphology provides evidence associated with differences in morphology as a function of geography/raw material.
+Bifaces from Millsap Cache and Jowell Farm were used to assess morphological stability/variability.
+Variability in elliptical biface morphology would provide evidence of local differences in morphology as a function of geography and/or raw material.
+Further, the geographic range/s of local elapid and viper populations differ between sites, thus, if elliptical bifaces are considered morphologically representative, then a difference in morphology should be apparent between the sites.
 
-This analysis follows recent analyses of Perdiz arrow points ([Selden and Dockall, in review](https://github.com/seldenlab/perdiz3/blob/main/ms/perdiz3/perdiz3.pdf)) and Gahagan biface morphology across the southern Caddo area and central Texas [@RN11783; @RN21001; @RN11264], where Perdiz arrow points and Gahagan bifaces have been found to differ in shape across the same geography as Caddo bottles [@RN11801; @RN11782; @RN11716; @RN8312].
-This analysis also follows another geometric morphometric analysis in which Perdiz arrow points from the ancestral Caddo area were found to differ as a function of *time*, *raw material*, and *burial context* [@RN9551].
+This analysis follows two recent formal analyses of Perdiz arrow point morphology [@RN9551;@RN11264] and another that is in press ([Selden and Dockall, in press](https://github.com/seldenlab/perdiz3/blob/main/ms/perdiz3.pdf)), as well as Gahagan biface morphology across the southern Caddo area and central Texas [@RN11783; @RN21001; @RN11264], where Perdiz arrow points and Gahagan bifaces have been found to differ in shape across the same geography as Caddo bottles [@RN11801; @RN11782; @RN11716; @RN8312].
+Any morphological difference that occurs between Millsap Cache and Jowell Farm would also provide further evidence in support of the shape boundary hypothesis.
+The analysis also follows another geometric morphometric analysis in which Perdiz arrow points from the ancestral Caddo area were found to differ as a function of *time*, *raw material*, and *burial context* [@RN9551].
+While temporal and depositional contexts are unknown for both caches of bifaces used in this study, any finding/s of extant morphological differences would support the hypothesis of elliptical biface morphology as a function of geography and/or raw material.
 

### 2022-12-19 supp/gmszclass.qmd
<edit index>

@@
 Two `size classes` (large/small) were used to assess whether the morphology of elliptical bifaces from Millsap Cache and Jowell Farm remains morphologically stable/consistent through reduction.
-
-This analysis follows recent formal analyses of Perdiz arrow points [@RN9551;@RN11264] as well as another that is in press ([Selden and Dockall, in press](https://github.com/seldenlab/perdiz3/blob/main/ms/perdiz3.pdf)), and Gahagan biface morphology across the southern Caddo area and central Texas [@RN11783; @RN21001; @RN11264], where Perdiz arrow points and Gahagan bifaces have been found to differ in shape across the same geography as Caddo bottles [@RN11801; @RN11782; @RN11716; @RN8312].
-This analysis also follows another geometric morphometric analysis in which Perdiz arrow points from the ancestral Caddo area were found to differ as a function of _time_, _raw material_, and _burial context_ [@RN9551].
 

### 2022-12-19 supp/index.qmd
<edit index>

--- 2022-12-16 supp/index.qmd
@@
 ---
-title: "Size matters: Divergent morphology in elliptical bifaces from sites in the American Southeast articulates with distinct local reduction practices"
+title: "Divergent elliptical biface morphology: Distinct local retouch practice or serpent dance proxy for Elapids and Vipers?"
 author:
@@
 In the western Middle Tennessee Valley, the Pickwick Burial Complex [@RN11180], dating to the Late Archaic/Gulf Formational period (3000—300 BCE) exhibits a continuation of bipointed blades, some notched or truncated, found in ritual caches and ceremonially broken.
-Elliptical bifaces find their most dramatic expression as Caddoan and Mississippian exaggerated, symbolic weaponry.
+Elliptical bifaces find their most dramatic expression as Mississippian and Caddoan exaggerated, symbolic weaponry.
 Mississippian Ramey knives, dating from c.
@@
 
-```{r figorget, fig.cap="Shell gorgets from sites in the American Southeast depicting elliptical bifaces."}
+```{r figorget, fig.cap="Shell gorgets from sites in the American Southeast depicting elliptical bifaces in the eagle/bird dance."}
 
@@
 
-Duck River and other sword-form bifaces are also present at Spiro Mound [@RN2720], with the longest measuring 55.9cm.
+Duck River and other sword-form bifaces are also present at Spiro Mound [@RN2720], with the longest measuring 55.9cm, or just over 22 inches.
 Smaller elliptical biface forms, roughly half the size of the larger form, have also been identified at Spiro, along with several wooden implements that Early [-@RN1052] interpreted as memetic of elliptical bifaces.
-The bifaces from this latter category are the focus of this study, which is aimed at assessing whether resharpening and/or retouch may have resulted in convergent morphology at two sites in the ancestral Caddo area.
-A measure of morphological disparity is subsequently to assess whether biface morphology is more diverse at the larger size, which would suggest greater diversity in biface shape or size related to production activities.
+Elliptical biface forms at Spiro differ from those to the east, as both ends are mirrored, and are either pointed _or_ rounded, and not pointed _and_ rounded.
+
+Among the shell cup illustrations assembled by Phillips and Brown [-@RN2447] that characterize the material culture from Spiro Mound is a depiction of elliptical bifaces conveyed as elements of the serpent/snake dance rather than elements of the eagle/bird dance.
+It is plausible that in this case biface **shape** may convey a relevant---and ceremonially important---representation related to extant morphological differences expressed in local elapid and viper species.
+Differences in biface **size** may articulate with the size of a specific species, as some are smaller than others (i.e.
+_Sistrurus miliarius streckeri_ or western pygmy rattlesnake).
+Alternatively, **size** differences might also depict allometric trends associated with the development and maturation of distinct species, as well as sexual size dimorphism.
+
+```{r figspiro, fig.cap="Shell cup from Spiro Mound depicting the serpent dance, where the difference in elliptical biface morphology between snake dancers and eagle dancers is made plain.
+This figure was redrafted from Figure 44 in the Phillips and Brown series [-@RN2447]."}
+
+knitr::include_graphics('./images/spiro.png')
+```
+
+This study uses the tools of landmark geometric morphometrics to assess whether resharpening and/or retouch of these important ceremonial bifaces may have resulted in convergent morphology at two sites in the ancestral Caddo area.
+A measure of morphological disparity was subsequently used to assess whether biface morphology may have been more diverse at the larger size, suggestive of greater diversity in biface shape or size related to production activities.
+
+Results can also be used to evaluate the hypothesis that elliptical biface morphology is representative of local species of the Elapidae and Viperidae families, where greater diversity would be expected in the larger size class.
+If representative of local snake species, the analysis of morphological disparity would indicate a greater diversity of size in the larger size class.
+This is due to the fact that all species will traverse the smaller size class during maturation and development before reaching adulthood, where morphology is known to differ across species.
 
@@
 
-```{r figelliptical, fig.cap="A. T. Jackson with the elliptical bifaces from Galt Farm in Franklin County during excavations by the University of Texas in 1931."}
+```{r figelliptical, fig.cap="A. T. Jackson with the elliptical bifaces found at Galt Farm in Franklin County, Texas during excavations by the University of Texas in 1931."}
 
@@
 Our thanks to the Caddo Nation of Oklahoma, the Caddo Nation Tribal Council, Tribal Chairman, and Tribal Historic Preservation Office for permission and access to NAGPRA and previously repatriated collections.
-Our gratitude is also extended to Marybeth Tomka and Lauren Bussiere at the Texas Archeological Research Laboratory for their assistance with access to the bifaces and associated records, to Sergio Ayala and the Gault School of Archaeological Research for access to the UV light, to Scott Hammerstedt and Debra K.
+Our gratitude is also extended to Lauren Bussiere at the Texas Archeological Research Laboratory for her assistance with access to the bifaces and associated records, to Sergio Ayala and the Gault School of Archaeological Research for access to the UV light, to Scott Hammerstedt and Debra K.
 Green at the Oklahoma Archeological Survey for their assistance with records requests, and to Marc Lavine and Susie Fishman-Armstrong at the Sam Noble Oklahoma Museum of Natural History for providing digital images from their collections.

### 2022-12-20 supp/index.qmd
<edit index>

@@
 
-Elliptical bifaces are prevalent at Mississipian sites throughout the American Southeast.
-Those from Millsap Cache and Jowell Farm comprise two of the largest samples of this ill-understood stone tool from the ancestral Caddo area.
-Bifaces from Millsap Cache were reportedly produced using Kay County flint, while those from Jowell Farm were manufactured using Edwards chert, providing for an empirical test of morphological differences as a function of raw material.
-The sample of elliptical bifaces was divided into two size classes; one conceptually reflective of production (large), and the other with local reduction practices (small).
-Size classes were used to assess whether modifications by Caddo knappers may have yielded similar---convergent---biface shapes in the small class.
-Size classes were also used to test the hypothesis that greater morphological variation would be apparent in the small class due to idiosyncratic responses related to local retouch practices.
-Results demonstrate that elliptical biface shape does not differ by raw material, but size does, suggesting that a shared and morphologically-consistent mental template was maintained independent of biface size.
+## Background
+
+Elliptical bifaces are prevalent at Mississipian and Caddo sites throughout the American Southeast, and are depicted in motifs that adorn shell gorgets and bowls, representing two distinct dances.
+Morphological attributes of elliptical bifaces used in Mississipian and Caddo imagery differs in noticeable ways that are manifest in associated material culture elements, and bifaces from Caddo sites more readily articulate with the serpent dance, where the motif portrays the emergence of the Hero Twins from the Beneath World; each grasping an elliptical biface.
+
+## Hypotheses
+
+The elliptical bifaces used in this study represent two of the largest samples cached at Caddo sites, and each was manufactured with a distinct raw material, providing for an empirical test of morphological differences as a function of raw material.
+If the elliptical bifaces were intended to represent local elapid and viper populations in the serpent dance, then their morphology may differ due to regional differences in biogeography.
+The sample of elliptical bifaces was also divided into two size classes; one conceptually reflective of production (large) or adult elapids and vipers, and the other with local reduction practices (small) or juvenile elapids and vipers.
+
+Size classes were used to assess whether modifications by Caddo knappers may have yielded similar---convergent---biface morphologies in the small class suggestive of relative standardization due to resharpening and/or retouch in one instance, and morphological divergence associated with adult elapids and vipers in the other.
+Size classes were also used to test two hypotheses related to morphological diversity; one where greater morphological variation would be apparent in the small size class due to idiosyncratic responses related to local retouch practices, and a second where greater morphological variation would be apparent in the large size class due to differences in constituent elapid and viper communities depicted by the bifaces.
+
+## Results
+
+Results demonstrate that elliptical biface shape does not differ by raw material, but size does, suggesting that a shared---and morphologically-consistent---mental template was maintained independent of biface size and constituent community differences in elapids and vipers.
 The subsequent analysis of elliptical biface morphology by size class demonstrated that size does not differ by raw material in the large class, but in the small class, it does.
-This finding supports the argument that elliptical biface morphology diverges through local reduction practices.
-As expected, greater shape diversity occurs in the small class, where Jowell Farm bifaces were found to be more morphologically diverse than those from Millsap Cache.
-Distinct local reduction practices are advanced as the driver of extant morphological differences found in elliptical bifaces.
+These findings support the argument that elliptical biface morphology diverges through local reduction practices, while simultaneously nullifying the hypothesis that biface shape reflects the morphology of local elapid and viper communities.
+As expected, greater shape diversity occurs in the small size class, where the Jowell Farm bifaces were found to be more morphologically diverse than those from Millsap Cache.
+Thus, distinct local reduction practices are advanced as the driver of extant morphological differences expressed in elliptical bifaces.
 
@@
 ```
+
+# Background
 
@@
 
-Among the shell cup illustrations assembled by Phillips and Brown [-@RN2447] that characterize the material culture from Spiro Mound is a depiction of elliptical bifaces conveyed as elements of the serpent/snake dance rather than elements of the eagle/bird dance.
+Among the shell cup illustrations assembled by Phillips and Brown [-@RN2447] that characterize the material culture at Spiro Mound is a depiction of elliptical bifaces conveyed as elements of the serpent/snake dance rather than elements of the eagle/bird dance.
 It is plausible that in this case biface **shape** may convey a relevant---and ceremonially important---representation related to extant morphological differences expressed in local elapid and viper species.
@@
 
-```{r figspiro, fig.cap="Shell cup from Spiro Mound depicting the serpent dance, where the difference in elliptical biface morphology between snake dancers and eagle dancers is made plain.
+```{r figspiro, fig.cap="Shell cup from Spiro Mound depicting the serpent dance, where the difference in elliptical biface morphology between eagle dancers and snake dancers is made plain.
 This figure was redrafted from Figure 44 in the Phillips and Brown series [-@RN2447]."}
@@
 
-This study uses the tools of landmark geometric morphometrics to assess whether resharpening and/or retouch of these important ceremonial bifaces may have resulted in convergent morphology at two sites in the ancestral Caddo area.
-A measure of morphological disparity was subsequently used to assess whether biface morphology may have been more diverse at the larger size, suggestive of greater diversity in biface shape or size related to production activities.
+## Hypotheses
 
-Results can also be used to evaluate the hypothesis that elliptical biface morphology is representative of local species of the Elapidae and Viperidae families, where greater diversity would be expected in the larger size class.
-If representative of local snake species, the analysis of morphological disparity would indicate a greater diversity of size in the larger size class.
-This is due to the fact that all species will traverse the smaller size class during maturation and development before reaching adulthood, where morphology is known to differ across species.
+1.
+Elliptical bifaces found at Millsap Cache were manufactured using Kay County Chert, while those from Jowell Farm were manufactured using Edwards Chert.
+Differences in stone tool shape as a function of raw material have been reported for Perdiz arrow points in the region [@RN9364]; however, others have demonstrated that the attribution of morphological variability to raw material differences is an assumption that warrants regular testing, suggesting that primacy should not be assumed [@RN8442].
+Thus, a comparison of between-site morphology was included to assess elliptical biface morphology as a function of raw material.
+  a)    The biogeography of elapid and viper species differs within the ancestral Caddo area, where elapid populations occur south of the Red River, excluding them from the geography surrounding Millsap Cache.
+If elliptical biface morphology is intended to represent local elapid and viper populations, then elliptical bifaces from these sites will differ due to differences in the constituent communities of elapids and vipers.
 
-## Analysis overview
+2.
+The elliptical bifaces from Jowell Farm were discovered in an area of the ancestral Caddo region where others have assumed large bifaces to be products of external manufacture due to the absence of local production evidence and the availability of raw material/s of similar size and quality [@RN3682;@RN11783;@RN21001].
+Given the potential that elliptical bifaces were produced by knappers outside of the ancestral Caddo region for at least one of the caches (Jowell Farm), it is assumed that the smaller elliptical bifaces will more readily reflect morphology imposed by Caddo knappers through resharpening and/or retouch incurred through normal use [@RN11264].
+Thus, it is expected that elliptical bifaces will express convergent morphology, where morphological consistency is achieved through the idiosyncratic approaches to resharpening and/or retouch by Caddo knappers in the small size class.
+  a)   Elapid and viper species will traverse the smaller size class through normal developmental stages before reaching adulthood, where their morphology is known to differ within and among species.
+If elliptical biface morphology is thought to represent distinct elapid and viper populations, then the most pronounced morphological difference will occur in the large size class among the adults.
+Thus, if elliptical bifaces were intended to represent elapid and viper species, then the morphological pattern will be identical to H2.
+However, the analytical results would be read in reverse, where expected similarities occur at the smaller size class, before diverging in the large size class.
+
+3.
+A measure of morphological disparity was subsequently used to assess whether biface morphology may have been more morphologically diverse in the large size class, suggestive of greater diversity in biface shape or size related to production activities [@RN11264].
+An alternative hypothesis is that the idiosyncratic responses to resharpening and/or retouch by Caddo knappers required to address damage incurred through normal use generate greater morphological diversity in the small size class [@RN11264].
+  a)    Elapid and viper species express greater morphological diversity at their larger size.
+Thus, if elliptical bifaces were intended to represent elapid and viper species, and biological diversity differs between sites---elapid biogeography restricted to the area of Jowell Farm---then morphological disparity will occur between sites in the large size class.
+
+# Analysis overview
 
@@
 
-### Generalised Procrustes Analysis
+## Generalised Procrustes Analysis
 
@@
 
-### Principal Components Analysis
+## Principal Components Analysis
 
@@
 
-### Procrustes ANOVA
+## Procrustes ANOVA
 
@@
 
-## Primary findings
+# Primary findings
 
+  -   Allometry (shape change with size) was not significant.
   -   Bifaces _do not differ_ in shape, but **differ significantly** in size by raw material.
-  -   Allometry (shape change with size) was not significant.
-  -   Bifaces in the _large class_ recovered from Millsap Cache and Jowell Farm do not differ in shape or size, suggesting consistent morphology in bifaces from the _large class_, supporting arguments for production-based standardization.
+  -   Bifaces in the _large class_ recovered from Millsap Cache and Jowell Farm do not differ in shape or size, suggesting consistent morphology in bifaces from the _large class_, supporting arguments for relative production-based standardization.
   -   Bifaces in the _small class_ recovered from Millsap Cache and Jowell Farm **differ significantly** in size.
-Thus, while shape remains consistent, size differs, suggesting---at minimum---locally divergent approaches to retouch or retouch intensity.
-  -   Morphological disparity in the shape of the _small class_ is **significantly greater** at Jowell Farm, providing evidence for a more standardized approach to local retouch at Millsap Cache.
-This finding provides additional support for the hypothesis that greater shape diversity is associated with retouch efforts, rather than production.
+Thus, while shape remains consistent, size differs, suggesting---at minimum---locally divergent approaches to retouch or retouch _intensity_.
+  -   Morphological disparity in the **shape** of elliptical bifaces in the _small size class_ is **significantly greater** at Jowell Farm, providing evidence for a more standardized approach to local retouch at Millsap Cache.
+This finding also yields added support for the hypothesis that greater shape diversity is associated with local retouch efforts, rather than production.
 
-## Parallel-oblique flaking
+# Parallel-oblique flaking
 
@@
 
-## Summary
+In the case of ceremonial bifaces, the intentional use of parallel oblique flake scars may have been intended to convey motion, where flake scars on one face represent muscle constriction on one edge and muscle dilation on the other, with an inverted pattern on the opposite face.
+Thus, in intentionally flipping between the two faces---which might have been a component of the serpent/snake dance---dancers could convey motion with a static implement.
+Additionally, the parallel oblique flake scars closely mimic the parallel scales that occur beneath both elapids and vipers, which shift in similar ways to generate motion.
+
+# Summary
 
@@
 
-## Acknowledgments {-}
+# Acknowledgments {-}
 
@@
 
-## Funding {-}
+# Funding {-}
 
@@
 
-## Data management {-}
+# Data management {-}
 
@@
 
-## Colophon {-}
+# Colophon {-}
 

### 2022-12-20 supp/index.qmd
<edit index>

@@
 
-Elliptical bifaces are prevalent at Mississipian and Caddo sites throughout the American Southeast, and are depicted in motifs that adorn shell gorgets and bowls, representing two distinct dances.
-Morphological attributes of elliptical bifaces used in Mississipian and Caddo imagery differs in noticeable ways that are manifest in associated material culture elements, and bifaces from Caddo sites more readily articulate with the serpent dance, where the motif portrays the emergence of the Hero Twins from the Beneath World; each grasping an elliptical biface.
+Elliptical bifaces are prevalent at Mississippian and Caddo sites throughout the American Southeast, and are depicted in motifs that adorn shell gorgets and bowls, representing two distinct dances.
+Morphological attributes of elliptical bifaces used in Mississippian and Caddo imagery differs in noticeable ways that are manifest in associated material culture elements, and bifaces from Caddo sites more readily articulate with the serpent dance, where the motif portrays the emergence of the Hero Twins from the Beneath World; each grasping an elliptical biface.
 
@@
 
-The earliest elliptical bifaces, associated with Middle Archaic period, recovered as Benton caches (4500—4000 BCE) in the Mid-South [@RN11179], are clearly non-utilitarian, as they “meet the expectations of sacred markers for secular exchange” [@RN11178].
+The earliest elliptical bifaces, associated with Middle Archaic period, recovered as Benton caches (4500—4000 BCE) in the Mid-South [@RN11179], and are clearly non-utilitarian, as they “meet the expectations of sacred markers for secular exchange” [@RN11178].
 In the western Middle Tennessee Valley, the Pickwick Burial Complex [@RN11180], dating to the Late Archaic/Gulf Formational period (3000—300 BCE) exhibits a continuation of bipointed blades, some notched or truncated, found in ritual caches and ceremonially broken.
@@
   a)    The biogeography of elapid and viper species differs within the ancestral Caddo area, where elapid populations occur south of the Red River, excluding them from the geography surrounding Millsap Cache.
-If elliptical biface morphology is intended to represent local elapid and viper populations, then elliptical bifaces from these sites will differ due to differences in the constituent communities of elapids and vipers.
+If elliptical biface morphology was intended to represent local elapid and viper populations, then elliptical bifaces from these sites will differ due to differences in the constituent communities of elapids and vipers.
 
@@
 Thus, it is expected that elliptical bifaces will express convergent morphology, where morphological consistency is achieved through the idiosyncratic approaches to resharpening and/or retouch by Caddo knappers in the small size class.
-  a)   Elapid and viper species will traverse the smaller size class through normal developmental stages before reaching adulthood, where their morphology is known to differ within and among species.
-If elliptical biface morphology is thought to represent distinct elapid and viper populations, then the most pronounced morphological difference will occur in the large size class among the adults.
+  a)   Elapid and viper species will progress through the smaller size class during normal developmental stages before reaching adulthood, where their morphology differs within and among species.
+If elliptical biface morphology is thought to represent distinct elapid and viper populations, then the most pronounced morphological difference will occur in the large size class among adults.
 Thus, if elliptical bifaces were intended to represent elapid and viper species, then the morphological pattern will be identical to H2.
-However, the analytical results would be read in reverse, where expected similarities occur at the smaller size class, before diverging in the large size class.
+However, analytical results would be read in reverse, where expected similarities occur at the smaller size class, before diverging in the large size class.
 
@@
 
-In the case of ceremonial bifaces, the intentional use of parallel oblique flake scars may have been intended to convey motion, where flake scars on one face represent muscle constriction on one edge and muscle dilation on the other, with an inverted pattern on the opposite face.
-Thus, in intentionally flipping between the two faces---which might have been a component of the serpent/snake dance---dancers could convey motion with a static implement.
-Additionally, the parallel oblique flake scars closely mimic the parallel scales that occur beneath both elapids and vipers, which shift in similar ways to generate motion.
+In the case of ceremonial bifaces, the intentional use of parallel oblique flake scars may have been meant to convey motion, where flake scars on one face represent muscle constriction along one edge and muscle dilation on the other, where the pattern is inverted on the opposite face.
+Thus, in intentionally flipping between the two faces---a potential component of the serpent/snake dance---dancers could convey motion using a static implement.
+The parallel oblique flake scars closely mimic the parallel scales that occur beneath both elapids and vipers, which shift in similar ways to generate motion.
 

### 2022-12-20 supp/index.qmd
<edit index>

@@
 Elliptical bifaces are prevalent at Mississippian and Caddo sites throughout the American Southeast, and are depicted in motifs that adorn shell gorgets and bowls, representing two distinct dances.
-Morphological attributes of elliptical bifaces used in Mississippian and Caddo imagery differs in noticeable ways that are manifest in associated material culture elements, and bifaces from Caddo sites more readily articulate with the serpent dance, where the motif portrays the emergence of the Hero Twins from the Beneath World; each grasping an elliptical biface.
+Morphological attributes of elliptical bifaces used in Mississippian and Caddo imagery differ in noteworthy ways that are manifest in the associated material culture assemblages, and bifaces from Caddo sites more readily articulate with the serpent dance, where the motif portrays the emergence of the Hero Twins from the Beneath World; each grasping an elliptical biface.
 
@@
 Among the shell cup illustrations assembled by Phillips and Brown [-@RN2447] that characterize the material culture at Spiro Mound is a depiction of elliptical bifaces conveyed as elements of the serpent/snake dance rather than elements of the eagle/bird dance.
-It is plausible that in this case biface **shape** may convey a relevant---and ceremonially important---representation related to extant morphological differences expressed in local elapid and viper species.
-Differences in biface **size** may articulate with the size of a specific species, as some are smaller than others (i.e.
+Snakes occupy an important place in Caddo lore where the Snake-Woman was the bringer of seeds to the Caddo, and no one, especially children, was permitted to point toward or touch the plants as they grow, since they belong to her until they are ripe [@RN11189, 18].
+
+It is plausible that in the case of the serpent/snake dance, biface **shape** may convey a relevant---and ceremonially important---representation related to extant morphological differences expressed in local elapid and viper species.
+Differences in biface **size** may articulate with the size of a specific species, as some snakes are smaller than others (i.e.
 _Sistrurus miliarius streckeri_ or western pygmy rattlesnake).
@@
 1.
-Elliptical bifaces found at Millsap Cache were manufactured using Kay County Chert, while those from Jowell Farm were manufactured using Edwards Chert.
+Elliptical bifaces found at Millsap Cache were manufactured using Kay County Chert [@RN11461], while those from Jowell Farm were manufactured using Edwards Chert.
 Differences in stone tool shape as a function of raw material have been reported for Perdiz arrow points in the region [@RN9364]; however, others have demonstrated that the attribution of morphological variability to raw material differences is an assumption that warrants regular testing, suggesting that primacy should not be assumed [@RN8442].
@@
 
-In the case of ceremonial bifaces, the intentional use of parallel oblique flake scars may have been meant to convey motion, where flake scars on one face represent muscle constriction along one edge and muscle dilation on the other, where the pattern is inverted on the opposite face.
+In the case of ceremonial bifaces, the intentional use of parallel oblique flake scars may have been meant to convey motion, where flake scars on one face represent muscle constriction along one edge and muscle dilation on the other, and where that pattern is inverted on the opposite face.
 Thus, in intentionally flipping between the two faces---a potential component of the serpent/snake dance---dancers could convey motion using a static implement.

### 2022-12-21 supp/dance.qmd
<init lit>

---
title: "Literature: Dance"
bibliography: [book.bib]
---

The dataset used in this analysis was harvested from the [Web of Science](https://clarivate.com/webofsciencegroup/solutions/web-of-science/), includes all 611 publications returned for the query **_dance AND (archaeology OR archeology OR anthropology)_**, and uses the `bibliometrix` package for analysis.

```{r bibliometrix, echo=TRUE, warning=FALSE}
# install
#devtools::install_github("massimoaria/bibliometrix")

# load
library(here)
library(bibliometrix)
library(reshape2)
library(ggplot2)

# build data frame
df <- convert2df(file = "savedrecs.bib",
                 dbsource = "wos",
                 format = "bibtex")
```

## Descriptive analysis

```{r summary, out.width = "100%", dpi = 300, echo=TRUE, warning=FALSE}
results <- biblioAnalysis(df, 
                          sep = ";")
options(width = 100)
s <- summary(object = results,
             k = 10,
             pause = FALSE)

plot(x = results,
     k = 10,
     pause = FALSE)
```

### Attributes of the local network

```{r local.attr, out.width = "100%", dpi = 300, echo=TRUE, warning=FALSE}
# calculate citations in local network
CR <- localCitations(df, sep = ";")

# top 10 cited authors in local network
knitr::kable((CR$Authors[1:10,]), "simple")

# top 10 cited papers in local network
knitr::kable((CR$Papers[1:10,]), "simple")
```

## Most cited

### Most cited articles

```{r most.cited.refs, out.width = "100%", dpi = 300, echo=TRUE, warning=FALSE}
# most cited references in global network
mcr <- citations(df,
                 field = "article",
                 sep = ";")

cbind(mcr$Cited[1:10])
```

### Most cited authors

```{r most.cited.auths, out.width = "100%", dpi = 300, echo=TRUE, warning=FALSE}
# most cited authors in global network
mcr <- citations(df,
                 field = "author",
                 sep = ";")

cbind(mcr$Cited[1:10])
```

### Author dominance ranking

```{r auth.dom, out.width = "100%", dpi = 300, echo=TRUE, warning=FALSE}
dom <- biblioAnalysis(df)
dom.r <- dominance(dom)

knitr::kable(dom.r, "simple")
```

### Productivity over time

```{r prod, out.width = "100%", dpi = 300, echo=TRUE, warning=FALSE}
topAU <- authorProdOverTime(df, k = 10, graph = TRUE)
```

## Historiograph direct citation linkages

The historiographic map is a chronological network map of the most relevant direct citations resulting from this bibliographic collection.

```{r hdcn, out.width = "100%", dpi = 300, echo=TRUE, warning=FALSE}
# historical citation network
options(width = 100)
histResults <- histNetwork(df,
                           min.citations = 20,
                           sep = ";")

# plot historical co-citation network
hnet <- histPlot(histResults,
                 n = 35,
                 size = 4,
                 labelsize = 1)
```

### ID and DE keyword associations

```{r key.assocx, out.width = "100%", dpi = 300, echo=TRUE, warning=FALSE}
list <- keywordAssoc(df,
                     sep = ";",
                     n = 10)

list[[1]][1:10]
```

### Yearly occurrences of top keywords/terms

#### Authors' keywords

```{r key.growth, out.width = "100%", dpi = 300, echo=TRUE, warning=FALSE}
topKW = KeywordGrowth(df,
                      Tag = "DE",
                      sep = ";",
                      top = 10,
                      cdf = TRUE)

topKW

# plot results
key.plot = melt(topKW,
                id ='Year')

ggplot(key.plot, aes(Year,
                     value,
                     group = variable,
                     color = variable)) + 
  geom_line()
```

#### Publisher's keywords

```{r key.growth2x, out.width = "100%", dpi = 300, echo=TRUE, warning=FALSE}
topKW = KeywordGrowth(df,
                      Tag = "ID",
                      sep = ";",
                      top = 10,
                      cdf = TRUE)

topKW

# plot results
key.plot = melt(topKW,
                id ='Year')

ggplot(key.plot, aes(Year,
                     value,
                     group = variable,
                     color = variable)) + 
  geom_line()
```

## Conceptual structure

### Co-word analysis

The co-word analysis maps the conceptual structure of a research domain using the co-occurrence of author keywords in the bibliographic collection.

#### Author's keywords

```{r co.word, out.width = "100%", dpi = 300, echo=TRUE, warning=FALSE}
# using authors keywords
cw <- conceptualStructure(df,
                          field = "DE",
                          method = "MCA",
                          minDegree = 5,
                          clust = "auto",
                          stemming = FALSE,
                          labelsize = 8)
```

#### Publisher's keywords

```{r co.word2i, out.width = "100%", dpi = 300, echo=TRUE, warning=FALSE}
# using publishers keywords
cw <- conceptualStructure(df,
                          field = "ID",
                          method = "MCA",
                          minDegree = 5,
                          clust = "auto",
                          stemming = FALSE,
                          labelsize = 8)
```

## Thematic mapping

From [@RN20996, 150-151]:

* Themes in the upper-right quadrant are both well developed and important for the structuring ofa research field. They are known as the motor-themes of the specialty, given that they present strong centrality and high density. The placement of themes in this quadrant implies that they are related externally to concepts applicable to other themes that are conceptually closely related.
* Themes in the upper-left quadrant have well developed internal ties but unimportant external ties and so are of only marginal importance for the field. These themes are very specialized and peripheral in character.
* Themes in the lower-left quadrant are both weakly developed and marginal. The themes of this quadrant have low density and low centrality, mainly representing either emerging or disappearing themes.
* Themes in the lower-right quadrant are important for a research field but are not developed. So, this quadrant groups transversal and general, basic themes.

### Authors' keywords

```{r thematic.map, out.width = "100%", dpi = 300, echo=TRUE, warning=FALSE}
# keyword map
map1 = thematicMap(df,
                   field = "DE",
                   n = 1000,
                   minfreq = 4,
                   ngrams = 1,
                   stemming = FALSE,
                   size = 0.3,
                   n.labels = 1,
                   community.repulsion = 0.1,
                   cluster = "walktrap",
                   repel = TRUE)
# plot map
plot(map1$map)
```

### Publisher's keywords

```{r thematic.map2b, out.width = "100%", dpi = 300, echo=TRUE, warning=FALSE}
# keyword map
map2 = thematicMap(df,
                   field = "ID",
                   n = 1000,
                   minfreq = 4,
                   ngrams = 1,
                   stemming = FALSE,
                   size = 0.6,
                   n.labels = 1,
                   community.repulsion = 0.1,
                   cluster = "walktrap",
                   repel = TRUE)
# plot map
plot(map2$map)
```

## Social structure

### Author collaboration

Scientific collaborations are plotted where nodes are authors and links are co-authorships, illustrating collaborations between authors.

```{r auth.collab, out.width = "100%", dpi = 300, echo=TRUE, warning=FALSE}
# author collaboration network
auth.collab <- biblioNetwork(df,
                             analysis = "collaboration",
                             network = "authors",
                             sep = ";")

# network plot
auth.collabnet = networkPlot(auth.collab,
                             n = 75,
                             Title = "Author Collaboration",
                             type = "mds",
                             halo = TRUE,
                             weighted = TRUE,
                             size = 5,
                             edgesize = 1,
                             labelsize = 0.5,
                             remove.isolates = TRUE)
```

### Edu collaboration

Scientific collaborations are plotted where nodes are institutions and links are co-authorships, illustrating collaborations between institutions.

```{r edu.network, out.width = "100%", dpi = 300, echo=TRUE, warning=FALSE}
# author collaboration network
edu.collab <- biblioNetwork(df,
                            analysis = "collaboration",
                            network = "universities",
                            sep = ";")

# network plot
edu.collabnet = networkPlot(edu.collab,
                            n = 75,
                            Title = "Edu Collaboration", 
                            type = "mds",
                            halo = TRUE,
                            size = 15, 
                            size.cex = T, 
                            edgesize = 1, 
                            labelsize = 0.5, 
                            remove.isolates = TRUE)

# end of code
```

### 2022-12-21 supp/index.qmd
<init lit>

@@
 Elliptical bifaces are prevalent at Mississippian and Caddo sites throughout the American Southeast, and are depicted in motifs that adorn shell gorgets and bowls, representing two distinct dances.
-Morphological attributes of elliptical bifaces used in Mississippian and Caddo imagery differ in noteworthy ways that are manifest in the associated material culture assemblages, and bifaces from Caddo sites more readily articulate with the serpent dance, where the motif portrays the emergence of the Hero Twins from the Beneath World; each grasping an elliptical biface.
+Morphological attributes of elliptical bifaces used in Mississippian and Caddo imagery differ in noteworthy ways that are manifest in the associated material culture assemblages.
+Those from Caddo sites more readily articulate with the serpent dance, where the decorative motif portrays the emergence of the Hero Twins from the Beneath World; each grasping an elliptical biface.
 
@@
 
-The elliptical bifaces used in this study represent two of the largest samples cached at Caddo sites, and each was manufactured with a distinct raw material, providing for an empirical test of morphological differences as a function of raw material.
+The elliptical bifaces used in this study come from two of the largest caches found at Caddo sites, and each was manufactured with a distinct raw material, providing for an empirical test of morphological differences as a function of raw material.
 If the elliptical bifaces were intended to represent local elapid and viper populations in the serpent dance, then their morphology may differ due to regional differences in biogeography.
@@
 
-Size classes were used to assess whether modifications by Caddo knappers may have yielded similar---convergent---biface morphologies in the small class suggestive of relative standardization due to resharpening and/or retouch in one instance, and morphological divergence associated with adult elapids and vipers in the other.
+Size classes were used to assess whether modifications by Caddo knappers may have yielded similar---convergent---biface morphologies in the small size class suggestive of relative standardization due to resharpening and/or retouch in one instance, and morphological divergence associated with adult elapids and vipers in the other.
 Size classes were also used to test two hypotheses related to morphological diversity; one where greater morphological variation would be apparent in the small size class due to idiosyncratic responses related to local retouch practices, and a second where greater morphological variation would be apparent in the large size class due to differences in constituent elapid and viper communities depicted by the bifaces.
@@
 Thus, in intentionally flipping between the two faces---a potential component of the serpent/snake dance---dancers could convey motion using a static implement.
-The parallel oblique flake scars closely mimic the parallel scales that occur beneath both elapids and vipers, which shift in similar ways to generate motion.
+The parallel oblique flake scars closely mimic the parallel scales that occur beneath both elapids and vipers, which can shift in similar ways to generate motion.
 

### 2022-12-24 supp/index.qmd
<edit index>

@@
 ---
-title: "Divergent elliptical biface morphology: Distinct local retouch practice or serpent dance proxy for Elapids and Vipers?"
+title: "Divergent elliptical biface morphology: Distinct local retouch practice or Serpent Dance proxy for elapids and vipers?"
 author:
@@
 github-repo: seldenlab/elliptical.bifaces
+description: Supplementary Materials
 ---
@@
 
-### Keywords {-}
+### Keywords
 
-American Southeast, Caddo, NAGPRA, computational archaeology, archaeoinformatics, museum studies, digital humanities, non-Western art history, geometric morphometrics, STEM, STEAM, ovoid biface, Galt biface, Jowell knife, Jowell knives, Jowell biface, Nikolas knife, lithics, parallel-oblique flaking
+_American Southeast, Caddo, NAGPRA, computational archaeology, archaeoinformatics, museum studies, digital humanities, non-Western art history, geometric morphometrics, STEM, STEAM, ovoid biface, Galt biface, Jowell knife, Jowell knives, Jowell biface, Nikolas knife, lithics, parallel-oblique flaking_
 
@@
 Thus, a comparison of between-site morphology was included to assess elliptical biface morphology as a function of raw material.
-  a)    The biogeography of elapid and viper species differs within the ancestral Caddo area, where elapid populations occur south of the Red River, excluding them from the geography surrounding Millsap Cache.
+  a)    The biogeography of elapid and viper species differs within the ancestral Caddo area, where elapids are absent from the geography surrounding Millsap Cache, and one viper species (_Crotalus atrox_, or western/Texas diamondback rattlesnake) is absent from the geography surrounding Jowell Farm.
 If elliptical biface morphology was intended to represent local elapid and viper populations, then elliptical bifaces from these sites will differ due to differences in the constituent communities of elapids and vipers.
@@
 Given the potential that elliptical bifaces were produced by knappers outside of the ancestral Caddo region for at least one of the caches (Jowell Farm), it is assumed that the smaller elliptical bifaces will more readily reflect morphology imposed by Caddo knappers through resharpening and/or retouch incurred through normal use [@RN11264].
-Thus, it is expected that elliptical bifaces will express convergent morphology, where morphological consistency is achieved through the idiosyncratic approaches to resharpening and/or retouch by Caddo knappers in the small size class.
-  a)   Elapid and viper species will progress through the smaller size class during normal developmental stages before reaching adulthood, where their morphology differs within and among species.
-If elliptical biface morphology is thought to represent distinct elapid and viper populations, then the most pronounced morphological difference will occur in the large size class among adults.
+Thus, it is expected that elliptical bifaces will express convergent morphology, where morphological consistency is achieved through the idiosyncratic approaches to resharpening and/or retouch by Caddo knappers in the _`small size class`_.
+  a)   Elapid and viper species progress through the smaller size class during normal developmental stages before reaching adulthood, where their morphology differs within and among species.
+If elliptical biface morphology is thought to represent distinct elapid and viper populations, then the most pronounced morphological difference will occur in the _`large size class`_, associated with adult snakes.
 Thus, if elliptical bifaces were intended to represent elapid and viper species, then the morphological pattern will be identical to H2.
-However, analytical results would be read in reverse, where expected similarities occur at the smaller size class, before diverging in the large size class.
+However, analytical results would be read in reverse, where expected similarities occur at the smaller size class, before diverging in the _`large size class`_.
 
 3.
-A measure of morphological disparity was subsequently used to assess whether biface morphology may have been more morphologically diverse in the large size class, suggestive of greater diversity in biface shape or size related to production activities [@RN11264].
-An alternative hypothesis is that the idiosyncratic responses to resharpening and/or retouch by Caddo knappers required to address damage incurred through normal use generate greater morphological diversity in the small size class [@RN11264].
+A measure of morphological disparity was subsequently used to assess whether biface morphology may have been more morphologically diverse in the _`large size class`_, suggestive of greater diversity in biface shape or size related to production activities [@RN11264].
+An alternative hypothesis is that the idiosyncratic responses to resharpening and/or retouch by Caddo knappers required to address damage incurred through normal use generate greater morphological diversity in the _`small size class`_ [@RN11264].
   a)    Elapid and viper species express greater morphological diversity at their larger size.
-Thus, if elliptical bifaces were intended to represent elapid and viper species, and biological diversity differs between sites---elapid biogeography restricted to the area of Jowell Farm---then morphological disparity will occur between sites in the large size class.
+Thus, if elliptical bifaces were intended to represent elapid and viper species, and biological diversity differs between sites---elapid biogeography restricted to the area of Jowell Farm---then morphological disparity will occur between sites in the _`large size class`_.
 
